---
title: "Get ADHD medications from OpenPrescribing"
format: html
---

```{r}
#| label: setup
#| echo: false
library(DBI)
library(bigrquery)
library(dplyr)
library(dbplyr)
library(here)
library(tidyverse)
library(arrow)

source(here("functions", "opr.R"))
```


```{r}
#| label: connect-bq
con <- connect_openprescribing()
```

# GP prescribing aggregated by chemical

Each row represents the total items prescribed for one chemical by one practice in one month.

```{r}
#| label: get-gp-prescribing

adhd_med_bnf_codes <- c(
  # Dexamfetamine
  # pseudoBNF code: Dexamfetamine sulfate (0404000L0)
  "0404000L0%",
  # Methylphenidate
  # pseudo BNF code: Methylphenidate hydrochloride (0404000M0)
  "0404000M0%",
  # Atomoxetine
  # pseudoBNF code: Atomoxetine hydrochloride (0404000S0)
  "0404000S0%",
  # Lisdexamfetamine
  # pseudoBNF code: Lisdexamfetamine dimesylate (0404000U0)
  "0404000U0%",
  # Guanfacine
  # pseudoBNF code: Guanfacine (0404000V0)
  "0404000V0%"
)

query_gp_prescribing <- get_gp_prescribing(
  con,
  bnf_codes = adhd_med_bnf_codes,
  start_date = "2015-08-01",
  end_date = "2025-07-01"
)

bnf_chemical_names <- c(
  "0404000L0" = "Dexamfetamine",
  "0404000M0" = "Methylphenidate",
  "0404000S0" = "Atomoxetine",
  "0404000U0" = "Lisdexamfetamine",
  "0404000V0" = "Guanfacine"
)

df_gp_prescribing <- query_gp_prescribing |>
  select(
    sha,
    regional_team,
    stp,
    pct,
    practice,
    bnf_code,
    bnf_name,
    items,
    quantity,
    month,
    name,
    postcode,
    ccg_id,
    pcn_id,
    setting,
    status_code
  ) |>
  collect() |>
  mutate(
    bnf_chemical = substr(bnf_code, 1, 9),
    bnf_chemical_name = unname(bnf_chemical_names[bnf_chemical]),
    month = as.Date(month)
  )

df_gp_prescribing_chemical <- df_gp_prescribing |>
  group_by(practice, month, bnf_chemical) |>
  mutate(items_by_chemical = sum(items)) |>
  ungroup() |>
  arrange(month, bnf_chemical, practice) |>
  select(-c(bnf_code, bnf_name, items, quantity)) |>
  distinct() |>
  relocate(items_by_chemical, .after = bnf_chemical_name)

write_parquet(
  df_gp_prescribing_chemical,
  here("output", "openprescribing", "adhd_gp_prescribing_chemical.parquet")
)
```

# GP prescribing by full BNF code

Unlike the chemical-level dataset above, each row here retains the full BNF code (individual preparation), and includes an `items_x_quantity` column (items × quantity) needed for defined daily dose (DDD) calculations.
This will be important for comparison with OpenPrescribing Hospitals data.

```{r}
#| label: get-gp-prescribing-full-bnf

df_gp_prescribing_bnf <- df_gp_prescribing |>
  select(-bnf_chemical) |>
  # Items*quantity gives the total units of each preparation prescribed by a practice for DDD calculations
  mutate(items_x_quantity = items * quantity) |>
  relocate(items_x_quantity, .after = quantity) |>
  relocate(bnf_chemical_name, .before = bnf_code) |>
  arrange(month, bnf_code, practice) |>
  distinct()

write_parquet(
  df_gp_prescribing_bnf,
  here("output", "openprescribing", "adhd_gp_prescribing_bnf.parquet")
)
```

# Monthly GP prescribing for all chemicals and by chemical 

```{r}
# df_gp_prescribing_chemical <- read_parquet(
#   here(
#     "output",
#     "openprescribing",
#     "adhd_gp_prescribing_chemical.parquet"
#   )
# )

# Monthly issuing by chemical
df_gp_by_chemical <- df_gp_prescribing_chemical |> 
  group_by(month, bnf_chemical_name) |> 
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop")

write_csv(
  df_gp_by_chemical,
  here("output", "monthly_op_gp", "adhd_gp_by_chemical.csv")
)

# Monthly issuing for all chemicals combined
df_gp_all_chemicals <- df_gp_prescribing_chemical |>
  group_by(month) |>
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop") |>
  mutate(bnf_chemical_name = "All chemicals")

write_csv(
  df_gp_all_chemicals,
  here("output", "monthly_op_gp", "adhd_gp_all_chemicals.csv")
)

# Monthly issuing by chemical and for all chemicals combined
df_total_gp_over_time <- bind_rows(df_gp_by_chemical, df_gp_all_chemicals) |> 
  arrange(month, bnf_chemical_name) 

write_csv(
  df_total_gp_over_time,
  here("output", "monthly_op_gp", "adhd_gp_total.csv")
)
```