---
title: "Get ICB mappings from ODS website and OpenPrescribing BQ data"
format: html
---
# Overview

Although OpenPrescribing GP prescribing data and SCMD hospital issuing data are already mapped to the most recent ICB (as of 2026-03-31), I still need to link ICB codes to ICB names for the thesis using ODS data.

Also, part of the FOI data supplied by the NHSBSA has supplied Area Team codes instead of ICB codes, which I need to map to ICB. Thankfully this work has already been done in the OpenPrescribing BigQuery data so I can simply extract the area team (`sha`) to ICB code (`stp`) mapping fron OpenPrescribing GP data. 

# Methods

```{r}
#| label: set-up
#| output: false
library(tidyverse)
library(here)
```

## Mapping ICB codes to ICB names

The mapping can be obtained from [this link](https://www.odsdatasearchandexport.nhs.uk/?search=generalorg&query=INTEGRATED%20CARE%20BOARD&columns=111010100001111100010011011000000000000000000000000000000000000000000000000). The above link is generated by going to [ODS data search and export](https://www.odsdatasearchandexport.nhs.uk/) then searching for "Integrated Care Board". Click export on the right hand side of the blue banner then chose the "download as csv (include headers)" option. The generated file is named "ODS General Search (Include headers).csv". I have placed it in output/icb_mapping.

```{r}
# Rename the csv file
file.rename(
  from = here("openprescribing", "mapping_data", "ODS General Search (Include headers).csv"),
  to = here("openprescribing", "mapping_data", "ods_icb_search.csv")
)

# Read in csv file
icb_code_name_map <- read_csv(
  here("openprescribing", "mapping_data", "ods_icb_search.csv"),
) |> 
  janitor::clean_names() |> 
  select(code, name, legal_start_date) |> 
  # Filter out new ICBs created after the last day of data we analyse
  mutate(legal_start_date = as.Date(
    legal_start_date, format = "%d-%m-%Y"
    )
  ) |> 
  filter_out(legal_start_date > as.Date("2025-12-31")) |> 
  rename(icb_code = code, icb_name = name) |> 
  select(-legal_start_date)
  
# Write mapping file for ICB codes to ICB names to csv
write_csv(
    icb_code_name_map,
    here("output", "icb_mapping", "icb_code_name_map.csv")
)
```

