---
title: "ADHD Medications Descriptive Analysis and Results"
author:
  - Kaylee Poon
format: html
---
# Overview

This analysis explores prescribing/issuing data for ADHD medications using the following data sources:

1. GP prescribing data (OpenPrescribing) from 2015-08-01 to 2025-07-01 (yearly items). 
2. FOI prescribing data from 2015-08-01 to 2025-07-01 (yearly items).
3. SCMD issuing data (OpenPrescribing-Hospitals website) from 2019-04-01 to 2025-03-01 (yearly DDD quantity).

In reality it's 2015-08-01 to 2025-07-31 and so on, but each month's data is recorded using the first day of each month. 

I will produce the following figures for the thesis: 

- Annual total ADHD medication prescriptions from each data source

- Regional trends and variations in ADHD medications in primary care (OpenPrescribing), secondary care (OpenPrescribing-Hospitals), and private care (NHS BSA FOI)

# Methods
```{r}
library(tidyverse)
library(here)
library(arrow)
library(patchwork)
library(ggtext)
library(ggnewscale)
library(stringr)
library(flextable)
library(officer)
library(readxl)
library(lubridate)

options(scipen = 999) # Turn off scientific notation

source(here("functions", "medication_plots.R"))

# Read in data files
gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
    )
)

total_gp_over_time <- read_csv(here("output", "monthly_op_gp", "adhd_gp_total.csv"))

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)

total_foi_over_time <- read_csv(here("output", "foi_prescribing", "adhd_foi_total.csv"))

scmd_adhd_issuing <- read_csv(
  here(
    "output",
    "web-openprescribing-hospitals",
    "adhd_oph_web_ddd_quantity.csv"
    )
) |> 
  janitor::clean_names() |> 
  rename(
    month = date,
    bnf_chemical_name = vtm_name,
    icb_name = icb
  ) |> 
  # Filter for only months in our analysis period
  filter(between(month, as.Date("2019-04-01"), as.Date("2025-03-01")))

total_scmd_over_time <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_total.csv")) |> 
  # Filter for only months in our analysis period
  filter(between(month, as.Date("2019-04-01"), as.Date("2025-03-01")))

# Read in ICB code -> ICB name mapping
icb_code_name_map <- read_csv(here("output", "icb_mapping", "icb_code_name_map.csv"))

# Read in yearly ICB populations
population_path <- here("output", "icb_mapping", "ons_icb_population_estimates.csv")

df_population_annual <- read_csv(population_path, show_col_types = FALSE) |> 
  transmute(
    population_year = as.integer(estimate_year),
    icb_name = icb_name,
    population = as.numeric(population)
  ) |> 
  distinct(population_year, icb_name, .keep_all = TRUE) |> 
  mutate(icb_name = str_to_upper(icb_name)) |> 
  filter(!(population_year %in% 2011:2015))
```

# Results: Yearly medication coding 

## Prescribing/issuing by chemical (yearly)

Will need to interpret the first year or so carefully because of missing data.

```{r}
# Helper function for calculating yearly totals from Aug 1st to Jul 31st
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
calc_yearly_total <- function(data, yearly_measure, monthly_measure){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jul 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by(bnf_chemical_name, end_date) |> 
    summarise(
      !!yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

# GP prescribing
df_yearly_gp_by_chemical <- calc_yearly_total(
  total_gp_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_gp_by_chemical <- plot_med_usage_shapes(
  data = df_yearly_gp_by_chemical, 
  usage_measure = yearly_items, 
  title_label = "Yearly GP prescribing (items)",
  n_breaks = 10
)

fig_yearly_gp_by_chemical

# FOI prescribing
df_yearly_foi_by_chemical <- calc_yearly_total(
  total_foi_over_time,
  "yearly_items",
  monthly_items
) 

fig_yearly_foi_by_chemical <- plot_med_usage_shapes(
  data = df_yearly_foi_by_chemical,  
  usage_measure = yearly_items, 
  title_label = "Yearly private controlled drug prescribing (items)",
  n_breaks = 10
)

fig_yearly_foi_by_chemical

# SCMD issuing - we want to aggregate from Apr 1st to Mar 31st to enable comparisons with ICD-10 code usage
df_yearly_scmd_by_chemical <- total_scmd_over_time |> 
  mutate(
    # Calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
    # Add end date of reporting year
    end_date = as.Date(paste0(reporting_year, "-03-31"))
  ) |> 
  group_by(bnf_chemical_name, end_date) |> 
  summarise(
    yearly_ddd_quantity = sum(monthly_ddd_quantity, na.rm = TRUE), .groups = "drop"
  )

fig_yearly_scmd_by_chemical <- plot_med_usage_shapes(
  data = df_yearly_scmd_by_chemical, 
  usage_measure = yearly_ddd_quantity, 
  title_label = "Yearly hospital issuing (DDD quantity)",
  n_breaks = 6
)

fig_yearly_scmd_by_chemical
```

Not doing a shared x-axis as the yearly aggregation periods are different. 

### Combined figures

```{r}
combined_yearly_plot <- (fig_yearly_gp_by_chemical / fig_yearly_foi_by_chemical / fig_yearly_scmd_by_chemical) + 
  plot_annotation(tag_levels = "A") +
  # Shared legend
  plot_layout(guides = "collect") &
	theme(
		plot.tag = element_text(size = 20),
    legend.position = "bottom",
		legend.title = element_blank(),
		legend.box = "horizontal"
	)

ggsave(here("output", "figures", "fig_yearly_med_coding.png"), combined_yearly_plot, width = 10, height = 12)
```

## Summary statistics

### Yearly medication coding summary

Includes total prescribing/issuing across all years, actual prescribing/issuing numbers for the first and last years for each dataset, and percentage change in prescribing/issuing combined for all chemicals between the first and last years. Will be reported as text in the thesis.

```{r}
# Prepare GP, FOI, and SCMD dataframes to be combined
gp_summary <- df_yearly_gp_by_chemical |> 
  mutate(dataset = "GP", unit = "items") |> 
  rename(yearly_value = yearly_items)

foi_summary <- df_yearly_foi_by_chemical |> 
  mutate(dataset = "FOI", unit = "items") |> 
  rename(yearly_value = yearly_items)

scmd_summary <-  df_yearly_scmd_by_chemical |> 
  mutate(dataset = "SCMD", unit = "ddd_quantity") |> 
  rename(yearly_value = yearly_ddd_quantity)

# Create summary table for total usage
df_yearly_summary <- bind_rows(gp_summary, foi_summary, scmd_summary) |> 
  filter(bnf_chemical_name == "All chemicals") |> 
  group_by(dataset) |> 
  mutate(
    # Start and end dates for each dataset - can change if appropriate
    start_year = min(end_date),
    start_value = yearly_value[which.min(end_date)],
    end_year = max(end_date),
    end_value = yearly_value[which.max(end_date)],
    # Total prescribing/issuing across time period 
    total = sum(yearly_value)
  ) |> 
  select(-c(end_date, yearly_value, bnf_chemical_name)) |> 
  distinct() |> 
  # Percentage change of last aggregation year vs first aggregation year
  mutate(pct_change = round((end_value - start_value) / start_value * 100, digits = 2))

print(df_yearly_summary$total)
```

### Percentage of FOI/GP prescribing for each chemical

Need to think about how I do this, as FOI data for atomoxetine and guanfacine is missing for the first few months. Currently doing first year vs last year comparison (but I might want to do second year vs last year, given that the first year won't be complete?). Also reported as text in thesis.

```{r}
# Same analysis time period: 2015-08-01 to 2025-07-01
df_foi_gp_pct <- bind_rows(df_yearly_gp_by_chemical, df_yearly_foi_by_chemical, .id = "dataset") |> 
  mutate(
    dataset = recode(dataset, "1" = "GP", "2" = "FOI")
  ) |> 
  group_by(dataset, bnf_chemical_name) |> 
  mutate(
    start_value = yearly_items[which.min(end_date)],
    end_value = yearly_items[which.max(end_date)],
    period_dataset_total = sum(yearly_items)
  ) |> 
  ungroup() |>
  select(-c(end_date, yearly_items)) |> 
  distinct() |> 
  group_by(bnf_chemical_name) |> 
  mutate(
    start_total = sum(start_value),
    start_pct = round(start_value / start_total * 100, digits = 2),
    end_total = sum(end_value),
    end_pct = round(end_value / end_total * 100, digits = 2),
    period_total = sum(period_dataset_total),
    period_pct = round(period_dataset_total / period_total * 100, digits = 2)
  ) |>
  select(-c(start_value, end_value, start_total, end_total, period_total, period_dataset_total)) |> 
  arrange(bnf_chemical_name)
```

# Results: ICB-level analysis

Helper functions for calculating ICB yearly totals from 1st Aug to 31st Jul and adding ICB population by year (need to use 2024 mid-year population for Aug 2024 to Jul 2025 year since 2025 mid-year population hasn't been published yet):

```{r}
# Helper function for calculating ICB yearly totals for GP and FOI data ONLY
icb_yearly_total <- function(data, icb_col, monthly_col, dataset){
  data |> 
    mutate(
      # Calculate reporting year (1st Aug - 31st Jul) from `month`
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by({{icb_col}}, end_date) |> 
    summarise(
      yearly_items = sum({{monthly_col}}, na.rm = TRUE), .groups = "drop"
    ) |> 
    mutate(dataset = dataset) # This is useful for summary stats later
}

add_population_by_year <- function(data) {
  data |> 
    # Use previous year's mid-year population if aggregation period ends before 30 June
    mutate(
      population_year = if_else(month(end_date) < 7, as.integer(year(end_date) - 1), as.integer(year(end_date)))
    ) |> 
    # Use 2024 mid-year population if `end_date` is in 2025
    mutate(
      population_year = if_else(year(end_date) == as.integer(2025), as.integer(2024), population_year)
    ) |> 
    left_join(
      df_population_annual,
      by = c("population_year", "icb_name")
    ) |> 
    select(-population_year)
}

# Check that ICB names in the population estimate and name/code mapping files are identical
setequal(na.omit(df_population_annual$icb_name), icb_code_name_map$icb_name)
# [1] TRUE
```

## GP ICB-level analysis

We obtained GP data from 2015-08-01 to 2025-07-31 from BigQuery to match SNOMED CT code usage reporting years. Therefore, we will aggregate from Aug 1st to Jul 31st the next year. Also need to map ICB codes to ICB names for this one.

```{r}
# The `stp` column contains the correct ICB codes as of 2026-03-01
unique(gp_adhd_prescribing$stp)
# 43 unique values including "NA" (unidentified doctors)

# Map ICB codes to ICB names and calculate items / 1000 population
gp_adhd_icb_yearly <- icb_yearly_total(gp_adhd_prescribing, stp, items_by_chemical, "GP") |>  
  rename(icb_code = stp) |>
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_code) |> 
  # Add ICB population by year
  add_population_by_year() |>
  # Calculate prescribed items / 1000 population 
  mutate(items_per_1000_pop = round(yearly_items / population * 1000, 2))

# Create violin plots
# `plot_yearly_violins` is in `medication_plots.R`
fig_gp_icb_violins <- plot_yearly_violins(
  data = gp_adhd_icb_yearly,
  usage_measure = items_per_1000_pop,
  title_label = "Yearly GP prescribing by ICB (items per 1,000 population)",
  n_breaks = 10
)

fig_gp_icb_violins
```

It's safe to ignore warnings since they are caused by unidentified doctors. 

## FOI ICB-level analysis

FOI data prior to 2020-04-01 is given with Area Team codes instead of ICB codes. Without practice-level information, it's impossible to map Area Team onto ICB accurately. Therefore, I will only analyse FOI data from 2020-08-01 to 2025-07-31 (to match with SNOMED CT code usage reporting years).

```{r}
foi_adhd_20to25 <- foi_adhd_prescribing |> 
  filter(between(month, as.Date("2020-08-01"), as.Date("2025-07-01"))) |> 
  # Using the mapping file to ensure consistent formatting of ICB names
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_name.x) |> 
  rename(icb_name = icb_name.y)

# Checking if ICB names look correct
setequal(na.omit(foi_adhd_20to25$icb_name), icb_code_name_map$icb_name)
# ICB names match the mapping file but FOI data includes "Unidentified doctors")

# Calculate yearly prescribing for each ICB per 1000 population
foi_adhd_icb_yearly <- icb_yearly_total(foi_adhd_20to25, icb_name, items, "FOI") |> 
  # Add ICB population by year
  add_population_by_year() |>
  # Calculate prescribed items / 1000 population 
  mutate(items_per_1000_pop = round(yearly_items / population * 1000, 2))

fig_foi_icb_violins <- plot_yearly_violins(
  data = foi_adhd_icb_yearly,
  usage_measure = items_per_1000_pop,
  title_label = "Yearly private controlled drug prescribing by ICB (items per 1,000 population)",
  n_breaks = 5
)

fig_foi_icb_violins
```

## SCMD ICB-level analysis

SCMD data is only available from 2019-01-01 to 2025-12-31. To match analysis time periods with ICD-10 code usage reporting years, I will only analyse SCMD hospital issuing from 2019-04-01 to 2025-03-31. 

```{r}
# Reformatting ICB names in SCMD to match the ICB population estimate file
scmd_adhd_issuing<- scmd_adhd_issuing |> 
  mutate(
    icb_name = paste0(icb_name, " Integrated Care Board") |> 
    str_to_upper() 
  )
setequal(scmd_adhd_issuing$icb_name, icb_code_name_map$icb_name)
# [1] TRUE

# Calculate yearly DDD quantity issued for each ICB
# Can't use `icb_yearly_total` as I am aggregating from April 1st to March 31st but the code follows the same logic
scmd_adhd_icb_yearly <- scmd_adhd_issuing |>  
  mutate(
    # Calculate reporting year (Apr 1st - Mar 31st) from `month` 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
    # Add end date of reporting year
    end_date = as.Date(paste0(reporting_year, "-03-31"))
    ) |> 
  group_by(icb_name, end_date) |> 
  summarise(yearly_ddd_quantity = sum(value, na.rm = TRUE), .groups = "drop") |> 
  # This is useful for summary stats later
  mutate(dataset = "SCMD") |> 
  # Add ICB population by year
  add_population_by_year() |>
  # Calculate issued DDD quantity / 1000 population 
  mutate(ddd_quantity_per_1000_pop = round(yearly_ddd_quantity / population * 1000, 2))

fig_scmd_icb_violins <- plot_yearly_violins(
  data = scmd_adhd_icb_yearly,
  usage_measure = ddd_quantity_per_1000_pop,
  title_label = "Yearly hospital issuing by ICB (DDD quantity per 1,000 population)",
  n_breaks = 6
)

fig_scmd_icb_violins
```

Combined figure

```{r}
combined_yearly_violins <- (fig_gp_icb_violins / fig_foi_icb_violins / fig_scmd_icb_violins) + 
  plot_annotation(tag_levels = "A") &
	theme(
		plot.tag = element_text(size = 20)
	)

ggsave(here("output", "figures", "fig_med_icb_violins.png"), combined_yearly_violins, width = 10, height = 12)
```

### Top total prescribing/issuing ICBs per 1000 population across all years

```{r}
# Prepare dfs for combination
gp_icb_yearly_summary <- gp_adhd_icb_yearly |> 
  mutate(unit = "items") |> 
  rename(yearly_per_1000_pop = items_per_1000_pop) 

foi_icb_yearly_summary <- foi_adhd_icb_yearly |> 
  mutate(unit = "items") |> 
  rename(yearly_per_1000_pop = items_per_1000_pop)

scmd_icb_yearly_summary <- scmd_adhd_icb_yearly |> 
  mutate(unit = "ddd_quantity") |> 
  rename(yearly_per_1000_pop = ddd_quantity_per_1000_pop)

# Summary df with all datasets, also useful later
icb_yearly_summary <- bind_rows(gp_icb_yearly_summary, foi_icb_yearly_summary, scmd_icb_yearly_summary) |> 
  select(-c(yearly_items, yearly_ddd_quantity, population)) |> 
  # Remove "Unidentified doctors"
  filter_out(is.na(icb_name))

df_icbs_ranked <- icb_yearly_summary |> 
  group_by(dataset, icb_name) |> 
  # Total prescribing/issuing across entire analysis period for each dataset
  mutate(total_per_1000_pop = sum(yearly_per_1000_pop)) |> 
  ungroup() |> 
  select(-c(end_date, yearly_per_1000_pop)) |> 
  distinct() |> 
  group_by(dataset) |> 
  mutate(
    icb_name = coalesce(icb_name, "Unidentified doctors"),
    # Ranking ICBs by total prescribing/issuing
    rank = min_rank(desc(total_per_1000_pop))
  ) |> 
  ungroup()

df_icbs_summary <- df_icbs_ranked |> 
  group_by(dataset) |> 
  mutate(
    # Get top and bottom ICBs
    top_icb = icb_name[which.min(rank)], bottom_icb = icb_name[which.max(rank)],
    # Get top and bottom 5 ICBs
    top_5 = paste(icb_name[order(rank)][1:5], collapse = ", "),
    bottom_5 = paste(icb_name[order(rank, decreasing = T)][1:5], collapse = ", "),
    # Calculate quartiles and IQR
    q1 = round(quantile(total_per_1000_pop, probs = 0.25, na.rm = TRUE),2),
    median = round(quantile(total_per_1000_pop, probs = 0.5, na.rm = TRUE), 2),
    q3 = round(quantile(total_per_1000_pop, probs = 0.75, na.rm = TRUE), 2),
    iqr = q3 - q1
  ) |> 
  select(-c(icb_name, total_per_1000_pop, rank)) |> 
  distinct()
```

### Trying to rank ICBs each year and get the top/bottom 5 for each year

Not very concise but works for now. Need to update after using rolling yearly population estimates. 

```{r}
df_icb_ranked_yearly <- icb_yearly_summary |> 
  group_by(end_date, dataset) |>
  mutate(
    # Ranking ICBs by yearly prescribing/issuing each year
    year_rank = min_rank(desc(yearly_per_1000_pop)),
    max_rank = max(year_rank, na.rm = T),
    # Create columns to highlight top/bottom 5 ICBs in the fig
    year_top5 = year_rank <= 5,
    year_bottom5 = year_rank > max_rank - 5
  ) |> 
  ungroup() |> 
  # Turn ICB names into title case and remove "NHS" prefix
  mutate(icb_name = icb_name |> 
    str_remove("^NHS\\s*") |>
    str_remove("\\s*INTEGRATED CARE BOARD$") |> 
    str_to_title()
  ) 

collapse_years <- function(years){
  years <- sort(unique(as.integer(years)))

  # Identify breaks between non-consecutive years
  groups <- cumsum(c(1, diff(years) != 1))
  ranges <- split(years, groups)

  sapply(ranges, function(x){
    if (length(x) == 1) {
      as.character(x)
    } else {
      paste0(min(x), "-", max(x))
    }
  }) |> paste(collapse = ", ")
}

top_5 <- df_icb_ranked_yearly |> 
  filter(year_top5 == T) |> 
  mutate(year = format(as.Date(end_date), "%Y")) |> 
  distinct(icb_name, year, dataset) |> 
  group_by(icb_name, dataset) |> 
  summarise(
    years = collapse_years(year),
    .groups = "drop"
  ) |> 
  mutate(icb_name = paste0(icb_name, "(", years, ")")) |> 
  select(-years) |> 
  arrange(dataset)

bottom_5 <- df_icb_ranked_yearly |> 
  filter(year_bottom5 == T) |> 
  mutate(year = format(as.Date(end_date), "%Y")) |> 
  distinct(icb_name, year, dataset) |> 
  group_by(icb_name, dataset) |> 
  summarise(
    years = collapse_years(year),
    .groups = "drop"
  ) |> 
  mutate(icb_name = paste0(icb_name, "(", years, ")")) |> 
  select(-years) |> 
  arrange(dataset)

top_bottom_5 <- bind_rows(top_5, bottom_5)
```

Quick and dirty export into Word for easier viewing - unlikely to be a table in the thesis but useful for writing up.

```{r}
icb_summary_table <- flextable(df_icbs_summary) |> 
  theme_booktabs() |> 
  set_caption("Total prescribing/issuing summary for ICBs") 

icb_summary_table

# Save as Word document
save_as_docx(
  "ICB Medication Summary Table" = icb_summary_table,
  path = "output/tables/icb_meds_summary_table.docx"
)
```

### Percentage of GP and FOI prescribing by unidentified doctors (N/A for SCMD)

```{r}
df_unidentified <- bind_rows(gp_icb_yearly_summary, foi_icb_yearly_summary) |> 
  mutate(icb_name = if_else(is.na(icb_name), "unidentified", "identified")) |> 
  group_by(dataset, icb_name) |> 
  summarise(total_items = sum(yearly_items), .groups = "drop") |> 
  group_by(dataset) |> 
  summarise(pct_unidentified = (total_items[icb_name == "unidentified"] / sum(total_items)) * 100, .groups = "drop")
```

Unidentified doctors contributed to 0.468% and 0.972% of total GP and FOI prescribing respectively. 
Note: no GP prescribing from unidentified doctors after 2020-07-31.

### ICB quartiles for the first and last year (GP, FOI, SCMD)

```{r}
df_first_last_icb_quartiles <- icb_yearly_summary |>  
  group_by(dataset) |> 
  # Only keep rows for the first and last year of each dataset
  filter(end_date == min(end_date) | end_date == max(end_date)) |> 
  group_by(dataset, end_date) |> 
  mutate(
    # Calculate total yearly value per 1000 population
    total_yearly = sum(yearly_per_1000_pop),
    # Calculate quartiles and IQR
    q1 = round(quantile(yearly_per_1000_pop, probs = 0.25, na.rm = TRUE), 2),
    median = round(quantile(yearly_per_1000_pop, probs = 0.5, na.rm = TRUE), 2),
    q3 = round(quantile(yearly_per_1000_pop, probs = 0.75, na.rm = TRUE), 2),
    iqr = q3 - q1
  ) |> 
  ungroup() |> 
  select(-c(yearly_per_1000_pop, icb_name)) |> 
  distinct()
```

These quartiles exclude unidentified doctors for GP and FOI prescribing. 