---
title: "Exploratory analysis (OpenCodeCounts) for codelist development"
format:
    html:
        toc: true
        toc-location: body
execute:
  message: false
---

# Setup
```{r}
#| output: false
library(opencodecounts)
library(here)
library(tibble)
library(dplyr)
library(ggplot2)
library(gt)
```

# GP records (SNOMED CT)
## ADHD diagnosis code usage
### Get usage data for each code in codelist
```{r}
# Import codelist from OpenCodelists.org
snomed_diag_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/kayleepn/adhd-diagnosis-snomed-ct-clone1/62ec200e/")
as_tibble(snomed_diag_codelist)

# Filter code usage data to include codes from SNOMED-CT diagnosis codelist
df_snomed_diag_usage <- snomed_usage |>
    filter(snomed_code %in% snomed_diag_codelist$code)
```

### Show code usage table for all codes with usage data
```{r}
#| message: false

# Summarise total code usage for each code
all_snomed_diag_codes <- df_snomed_diag_usage |>
    group_by(snomed_code, description) |>
    summarise(total_usage = sum(usage)) |>
    ungroup() |> 
    arrange(desc(total_usage))

# Create list of borderline codes
borderline_snomed_diag_codes <- c(
    192132008,
    229715008,
    192131001,
    28271000087107,
    46745001
)

# Create code usage table with borderline codes highlighted. 
all_snomed_diag_usage <- all_snomed_diag_codes |> 
    gt() |> 
    tab_header(
        title = "SNOMED ADHD diagnosis code usage") |> 
    tab_style(
        cell_fill(color = "lightblue"),
        locations = cells_body(
            rows = snomed_code %in% borderline_snomed_diag_codes
        )
    )

all_snomed_diag_usage
```

Borderline codes that may not be relevant for the thesis are highlighted in light blue. We may also want to consider if we are including codes related to drug treatment of ADHD. 

### Select top 5 codes with most usage

```{r}
top5_snomed_diag_codes <- all_snomed_diag_codes |> 
    slice_max(total_usage, n = 5) |> 
    pull(snomed_code)
```

### Visualise trends over time (top 5 codes)
```{r}
#| fig-width: 10
#| fig-height: 6
plot_snomed_diag_usage <- df_snomed_diag_usage |> 
    filter(snomed_code %in% top5_snomed_diag_codes) |> 
    ggplot(aes(
        x = end_date,
        y = usage,
        colour = paste0(description, "(", snomed_code, ")"),
        shape = paste0(description, "(", snomed_code, ")" )
    )) +
    geom_line(alpha = .3) +
    geom_point() +
    scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
    scale_x_date(date_breaks = "2 years",
    date_labels = "%Y") +
    scale_colour_viridis_d(end = .7) +
    labs(
        x = NULL, y = "Usage count", 
        colour = NULL, shape = NULL
    ) +
    theme(legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "vertical")
plot_snomed_diag_usage    
```

## ADHD admin-related code usage

### Get usage data for each code in codelist
This codelist contains codes related to referral, assessment, and administrative activity for ADHD, but not diagnosis or treatment.
```{r}
# Import codelist from OpenCodelists.org
snomed_admin_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/kayleepn/adhd-referral-assessment-and-admin-related-codes/38f84b80/")
as_tibble(snomed_admin_codelist)

# Filter code usage data to include codes from SNOMED-CT admin codelist
df_snomed_admin_usage <- snomed_usage |>
    filter(snomed_code %in% snomed_admin_codelist$code)
```

### Show code usage table for all codes with usage data
```{r}
#| message: false
all_snomed_admin_codes <- df_snomed_admin_usage |>
    group_by(snomed_code, description) |>
    summarise(total_usage = sum(usage)) |>
    ungroup() |> 
    arrange(desc(total_usage))
print(all_snomed_admin_codes)
```

### Select top 5 codes with most usage

```{r}
top5_snomed_admin_codes <- all_snomed_admin_codes |> 
    slice_max(total_usage, n = 5) |> 
    pull(snomed_code)
```

### Visualise trends over time (top 5 codes)
```{r}
#| fig-width: 10
#| fig-height: 6
plot_snomed_admin_usage <- df_snomed_admin_usage |> 
    filter(snomed_code %in% top5_snomed_admin_codes) |> 
    ggplot(aes(
        x = end_date,
        y = usage,
        colour = paste0(description, "(", snomed_code, ")"),
        shape = paste0(description, "(", snomed_code, ")" )
    )) +
    geom_line(alpha = .3) +
    geom_point() +
    scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
    scale_x_date(date_breaks = "2 years",
    date_labels = "%Y") +
    scale_colour_viridis_d(end = .7) +
    labs(
        x = NULL, y = "Usage count", 
        colour = NULL, shape = NULL
    ) +
    theme(legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "vertical")
plot_snomed_admin_usage    
```

# Hospital admitted patient care (ICD-10) ADHD diagnosis code usage

## Get usage data for each code in codelist
```{r}
# Import codelist from OpenCodelists.org
icd10_diag_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/kayleepn/adhd-icd-10-test/27569f95/")
as_tibble(icd10_diag_codelist)

# Filter code usage data to include codes from SNOMED-CT codelist
df_icd10_diag_usage <- icd10_usage |>
    filter(icd10_code %in% icd10_diag_codelist$code)
```

## Show code usage table for all codes with usage data
```{r}
#| message: false
all_icd10_diag_codes <- df_icd10_diag_usage |>
    group_by(icd10_code, description) |>
    summarise(total_usage = sum(usage)) |>
    ungroup() |> 
    arrange(desc(total_usage))

# Create list of borderline codes
borderline_icd10_diag_codes <- c(
    "F90",
    "F901",
    "F908",
    "F909"
)

# Create code usage table with borderline codes highlighted. 
all_icd10_diag_usage <- all_icd10_diag_codes |> 
    gt() |> 
    tab_header(
        title = "ICD-10 ADHD diagnosis code usage") |> 
    tab_style(
        cell_fill(color = "lightblue"),
        locations = cells_body(
            rows = icd10_code %in% borderline_icd10_diag_codes
        )
    )

all_icd10_diag_usage
```

Borderline codes that may not be relevant for the thesis are highlighted in light blue.

## Visualise trends over time (all codes)
```{r}
#| fig-width: 10
#| fig-height: 6
plot_icd10_diag_usage <- df_icd10_diag_usage |> 
    ggplot(aes(
        x = end_date,
        y = usage,
        colour = paste0(description, "(", icd10_code, ")"),
        shape = paste0(description, "(", icd10_code, ")" )
    )) +
    geom_line(alpha = .3) +
    geom_point() +
    scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
    scale_x_date(date_breaks = "2 years",
    date_labels = "%Y") +
    scale_colour_viridis_d(end = .7) +
    labs(
        x = NULL, y = "Usage count", 
        colour = NULL, shape = NULL
    ) +
    theme(legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "vertical")
plot_icd10_diag_usage    
```