---
title: "ADHD Medications Descriptive Analysis and Results"
author:
  - Kaylee Poon
format: html
---
# Overview

This analysis explores prescribing/issuing data for ADHD medications using the following data sources:

1. GP prescribing data (OpenPrescribing) from 2015-08-01 to 2025-07-01 (monthly items). 
2. FOI prescribing data from 2015-08-01 to 2025-07-01 (monthly items).
3. SCMD issuing data (OpenPrescribing-Hospitals website) from 2019-01-01 to 2025-12-01 (monthly DDD quantity).

I will produce the following figures for the thesis: 

- Annual total ADHD medication prescriptions from each data source

- Regional trends and variations in ADHD medications in primary care (OpenPrescribing), secondary care (OpenPrescribing-Hospitals), and private care (NHS BSA FOI)

# Methods
```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(nhstt)
library(arrow)
library(patchwork)
library(ggtext)
library(ggnewscale)
library(stringr)

source(here("functions", "medication_plots.R"))

# Read in data files
gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
    )
)

total_gp_over_time <- read_csv(here("output", "monthly_op_gp", "adhd_gp_total.csv"))

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)

total_foi_over_time <- read_csv(here("output", "foi_prescribing", "adhd_foi_total.csv"))

scmd_adhd_issuing <- read_csv(
  here(
    "output",
    "web-openprescribing-hospitals",
    "adhd_oph_web_ddd_quantity.csv"
    )
) |> 
  janitor::clean_names() |> 
  rename(
    month = date,
    bnf_chemical_name = vtm_name,
    icb_name = icb
  ) 

total_scmd_over_time <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_total.csv"))

# Read in ICB code -> ICB name mapping
icb_code_name_map <- read_csv(here("output", "icb_mapping", "icb_code_name_map.csv"))
```

# Results: medication usage


## Prescribing/issuing by chemical (monthly)
```{r}
fig_monthly_gp_by_chemical <- plot_med_usage(total_gp_over_time, monthly_items, "Monthly items", "GP Prescribing")

fig_monthly_gp_by_chemical

fig_monthly_foi_by_chemical <- plot_med_usage(total_foi_over_time, monthly_items, "Monthly items", "FOI Prescribing")

fig_monthly_foi_by_chemical

fig_monthly_scmd_by_chemical <- plot_med_usage(total_scmd_over_time, monthly_ddd_quantity, "Monthly DDD quantity", "SCMD issuing")

fig_monthly_scmd_by_chemical

combined_monthly_fig_1col <- fig_monthly_gp_by_chemical / fig_monthly_foi_by_chemical / fig_monthly_scmd_by_chemical + 
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect")

combined_monthly_fig_1col

ggsave(
  here("output", "figures", "fig_1col_monthly_med_usage.png"),
  combined_monthly_fig_1col,
  width = 15,
  height = 15
)
```

I find that a 1*3 layout works better. If this ends up in the thesis, it would be nice to use a common x-axis. 

Might be good to do some % calculations for each chemical? Also think about if I should aggregate this into yearly data instead. 

## Prescribing/issuing by chemical (yearly)

Will need to interpret the first year or so carefully because of missing data.

```{r}
# Helper function for calculating yearly totals from Aug 1st to Jul 31st
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
calc_yearly_total <- function(data, yearly_measure, monthly_measure){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jul 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by(bnf_chemical_name, end_date) |> 
    summarise(
      !!yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

# GP prescribing
df_yearly_gp_by_chemical <- calc_yearly_total(
  total_gp_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_gp_by_chemical <- plot_med_usage_shapes(df_yearly_gp_by_chemical, end_date, yearly_items, "GP Prescribing")

fig_yearly_gp_by_chemical

# FOI prescribing
df_yearly_foi_by_chemical <- calc_yearly_total(
  total_foi_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_foi_by_chemical <- plot_med_usage_shapes(df_yearly_foi_by_chemical, end_date, yearly_items, "FOI Prescribing")

fig_yearly_foi_by_chemical

# SCMD issuing - we want to aggregate from Apr 1st to Mar 31st to enable comparisons with ICD-10 code usage
df_yearly_scmd_by_chemical <- total_scmd_over_time |> 
  mutate(
    # calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
    # Add end date of reporting year
    end_date = as.Date(paste0(reporting_year, "-03-31"))
  ) |> 
  group_by(bnf_chemical_name, end_date) |> 
  summarise(
    yearly_ddd_quantity = sum(monthly_ddd_quantity, na.rm = TRUE), .groups = "drop"
  )

fig_yearly_scmd_by_chemical <- plot_med_usage_shapes(df_yearly_scmd_by_chemical, end_date, yearly_ddd_quantity, "SCMD Prescribing")

fig_yearly_scmd_by_chemical
```

It's now easier to distinguish between different lines. Will need to fix x-axis tick labelling if we end up reporting yearly data at this point. 

### Percentage stacked bar chart for GP and FOI prescribing (total prescribing by chemical)

We will report the percentages shown in this chart as text instead - need to calculate!

```{r}
# df_total_gp_foi <- bind_rows(df_yearly_gp_by_chemical, df_yearly_foi_by_chemical, .id = "source") |> 
#   mutate(
#     source = recode(source, "1" = "GP", "2" = "FOI")
#   ) |> 
#   group_by(bnf_chemical_name, source) |> 
#   summarise(total_items = sum(yearly_items), na.rm = T, .groups = "drop")

# fig_total_gp_foi <- ggplot(
#   df_total_gp_foi,
#   aes(fill = source, y = total_items, x = bnf_chemical_name)
# ) +
#   # Stacked percentage bar chart
#   geom_bar(position = "fill", stat = "identity") +
#   scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
#   scale_fill_viridis_d(begin = 0.25, end = 0.5) +
#   labs(title = "Percentage of GP/FOI Prescribing by Chemical") +
#   theme_bw(base_size = 12) +
#   theme(text = element_text(family = "Times New Roman"))

# fig_total_gp_foi
```

Kind of ugly but does the job. Can add percentage labels. 

### Combined figures

```{r}
combined_monthly_plot <- ((fig_monthly_gp_by_chemical / fig_monthly_foi_by_chemical) | plot_spacer() | fig_monthly_scmd_by_chemical / fig_total_gp_foi) + 
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_monthly_plot

ggsave(here("output", "figures", "fig_monthly_med_usage.png"), combined_monthly_plot, width = 15, height = 15)

combined_yearly_plot <- ((fig_yearly_gp_by_chemical / fig_yearly_foi_by_chemical) | plot_spacer() | (fig_yearly_scmd_by_chemical / fig_total_gp_foi)) + 
  plot_annotation(title = "Version with yearly data", tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_yearly_plot

ggsave(here("output", "figures", "fig_yearly_med_usage.png"), combined_yearly_plot, width = 15, height = 15)
```


# Results: ICB-level analysis

## GP ICB-level analysis

We obtained GP data from 2015-08-01 to 2025-07-31 from BigQuery to match SNOMED CT code usage reporting years. Therefore, we will aggregate from Aug 1st to Jul 31st the next year. Also need to map ICB codes to ICB names for this one.

```{r}
# The `stp` column contains the correct ICB codes as of 2026-03-01
unique(gp_adhd_prescribing$stp)
# 43 unique values including "NA" (unidentified doctors)

# Helper function for calculating ICB yearly totals for GP and FOI data ONLY
icb_yearly_total <- function(data, icb_col, monthly_col){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jul 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by({{icb_col}}, end_date) |> 
    summarise(
      yearly_items = sum({{monthly_col}}, na.rm = TRUE), .groups = "drop"
    ) |> 
    mutate(end_date = as.factor(end_date))
}

# Map ICB codes to ICB names
gp_adhd_icb_yearly <- icb_yearly_total(gp_adhd_prescribing, stp, items_by_chemical) |>  
  rename(icb_code = stp) |>
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_code)

# Helper function for ranking ICBs according to yearly medication usage
# `yearly_col` is the column name of yearly quantity measures (items or ddd quantity)

rank_icb_yearly <- function(data, yearly_col){
  data |> 
    group_by(end_date) |>
    mutate(
      icb_name = coalesce(icb_name, "Unidentified doctors"),
      # Ranking ICBs by yearly items each year
      year_rank = min_rank(desc({{yearly_col}})),
      max_rank = max(year_rank, na.rm = T),
      # Create columns to highlight top/bottom 5 ICBs in the fig
      year_top5 = year_rank <= 5,
      year_bottom5 = year_rank > max_rank - 5
    ) |> 
    # Turn ICB names into title case and remove "NHS" prefix
    mutate(icb_name = icb_name |> 
        str_remove("^NHS\\s*") |>
        str_remove("\\s*Integrated Care Board$") |> 
        str_to_title()
    ) |> 
    ungroup()
}

gp_adhd_icb_ranked <- rank_icb_yearly(gp_adhd_icb_yearly, yearly_items)

# Helper function for creating a legend for top 5 ICBs
# Need to be very specific here because I want the legend as
# "ICB name, (years the ICB was in the top 5) 
create_top_legend <- function(data){
  data |> 
    filter(year_top5 == T) |> 
    mutate(year = format(as.Date(end_date), "%Y")) |> 
    distinct(icb_name, year) |> 
    group_by(icb_name) |> 
    summarise(
      years = paste(sort(unique(year)), collapse = ", "),
      .groups = "drop"
    ) |> 
    mutate(label = paste0(icb_name, "<br>(", years, ")"))
}

# Helper function for creating a legend for bottom 5 ICBs - same logic as above

create_bottom_legend <- function(data){
  data |> 
    filter(year_bottom5 == T) |> 
    mutate(year = format(as.Date(end_date), "%Y")) |> 
    distinct(icb_name, year) |> 
    group_by(icb_name) |> 
    summarise(
      years = paste(sort(unique(year)), collapse = ", "),
      .groups = "drop"
    ) |> 
    mutate(label = paste0(icb_name, "<br>(", years, ")"))
}

top_gp_legend <- create_top_legend(gp_adhd_icb_ranked)
bottom_gp_legend <- create_bottom_legend(gp_adhd_icb_ranked)

# Create violin plots
# `plot_yearly_violins` is in `medication_plots.R`
gp_shape_values <- c(21, 22, 23, 24, 25, 8)
fig_gp_icb_violins <- plot_yearly_violins(
  data = gp_adhd_icb_ranked,
  usage_measure = yearly_items,
  top_legend = top_gp_legend,
  bottom_legend = bottom_gp_legend,
  shape_values = gp_shape_values,
  title_label = "Yearly GP Prescribing by ICB"
)

fig_gp_icb_violins

ggsave(here("output", "figures", "fig_gp_icb_violins.png"), fig_gp_icb_violins, width = 15, height = 10)
```

## FOI ICB-level analysis

FOI data prior to 2020-04-01 is given with Area Team codes instead of ICB codes. Without practice-level information, it's impossible to map Area Team onto ICB accurately. Therefore, I will only analyse FOI data from 2020-08-01 to 2025-07-31 (to match with SNOMED CT code usage reporting years).

```{r}
foi_adhd_20to25 <- foi_adhd_prescribing |> 
  filter(month >= "2020-08-01" & month < "2025-08-01") |> 
  # Using the mapping file to ensure consistent formatting of ICB names
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_name.x) |> 
  rename(icb_name = icb_name.y)

setequal(na.omit(foi_adhd_20to25$icb_name), icb_code_name_map$icb_name)
# ICB names match the mapping file but FOI data includes "Unidentified doctors")

# Calculate yearly prescribing for each ICB
foi_adhd_icb <- icb_yearly_total(foi_adhd_20to25, icb_name, items)
# Rank ICBs according to yearly prescribing
foi_adhd_icb_ranked <- rank_icb_yearly(foi_adhd_icb, yearly_items)

# Create top and bottom legends
top_foi_legend <- create_top_legend(foi_adhd_icb_ranked)
bottom_foi_legend <- create_bottom_legend(foi_adhd_icb_ranked)

# Create violin plot
foi_shape_values <- c(21, 22, 23, 24, 25, 8, 11)

fig_foi_icb_violins <- plot_yearly_violins(
  data = foi_adhd_icb_ranked,
  usage_measure = yearly_items,
  top_legend = top_foi_legend,
  bottom_legend = bottom_foi_legend,
  shape_values = foi_shape_values,
  title_label = "Yearly FOI Prescribing by ICB"
)

fig_foi_icb_violins

ggsave(here("output", "figures", "fig_foi_icb_violins.png"), fig_foi_icb_violins, width = 15, height = 10)
```

## SCMD ICB-level analysis

SCMD data is only available from 2019-01-01 to 2025-12-31. To match analysis time periods with ICD-10 code usage reporting years, I will only analyse SCMD hospital issuing from 2019-04-01 to 2025-03-31. 

```{r}
# ICB names in SCMD essentially match the mapping file but are formatted differently
scmd_adhd_issuing_caps <- scmd_adhd_issuing |> 
  mutate(
    icb_name = paste0(icb_name, " Integrated Care Board") |> 
    str_to_upper() 
  )
setequal(scmd_adhd_issuing_caps$icb_name, icb_code_name_map$icb_name)
# [1] TRUE

# Calculate yearly DDD quantity for each ICB
# Can't use `icb_yearly_total` as I am aggregating from April 1st to March 31st but the code follows the same logic
scmd_adhd_icb_yearly <- scmd_adhd_issuing |> 
  filter(month >= "2019-04-01" & month < "2025-04-01") |> 
  mutate(
    # calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-03-31"))
    ) |> 
  group_by(icb_name, end_date) |> 
  summarise(yearly_ddd_quantity = sum(value, na.rm = TRUE), .groups = "drop") |> 
  # `end_date` needs to be a factor for the violin plot
  mutate(end_date = as.factor(end_date))

# Rank ICBs by yearly DDD quantity
scmd_adhd_icb_ranked <- rank_icb_yearly(scmd_adhd_icb_yearly, yearly_ddd_quantity)

# Create top and bottom legends
top_scmd_legend <- create_top_legend(scmd_adhd_icb_ranked)
bottom_scmd_legend <- create_bottom_legend(scmd_adhd_icb_ranked)

# Create violin plot
scmd_shape_values <- c(21, 22, 23, 24, 25, 8, 11)

fig_scmd_icb_violins <- plot_yearly_violins(
  data = scmd_adhd_icb_ranked,
  usage_measure = yearly_ddd_quantity,
  top_legend = top_scmd_legend,
  bottom_legend = bottom_scmd_legend,
  shape_values = scmd_shape_values,
  title_label = "Yearly SCMD Issuing by ICB"
)

fig_scmd_icb_violins

ggsave(here("output", "figures", "fig_scmd_icb_violins.png"), fig_scmd_icb_yearly, width = 15, height = 10)
```

## Summary statistics

Can have the following:

- Total prescribing/issuing quantity for each data source for the entire analysis period (for every single month or to match yearly totals?) and possibly percentage change
- Percentage of FOI/GP prescribing for each chemical
- Possibly the top total prescribing/issuing ICBs?
- Percentage of FOI and GP prescribing by unidentified doctors (N/A for SCMD)
- ICB quartiles for the first and last year (GP, FOI, SCMD)