---
title: "ADHD Medications Descriptive Analysis and Results"
author:
  - Kaylee Poon
format: html
---
# Overview

This analysis explores prescribing/issuing data for ADHD medications using the following data sources:

1. GP prescribing data (OpenPrescribing) from 2015-08-01 to 2025-07-01 (yearly items). 
2. FOI prescribing data from 2015-08-01 to 2025-07-01 (yearly items).
3. SCMD issuing data (OpenPrescribing-Hospitals website) from 2019-04-01 to 2025-03-01 (yearly DDD quantity).

In reality it's 2015-08-01 to 2025-07-31 and so on, but each month's data is recorded using the first day of each month. 

I will produce the following figures for the thesis: 

- Annual total ADHD medication prescriptions from each data source

- Regional trends and variations in ADHD medications in primary care (OpenPrescribing), secondary care (OpenPrescribing-Hospitals), and private care (NHS BSA FOI)

# Methods
```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(nhstt)
library(arrow)
library(patchwork)
library(ggtext)
library(ggnewscale)
library(stringr)

options(scipen = 999) # Turn off scientific notation

source(here("functions", "medication_plots.R"))

# Read in data files
gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
    )
)

total_gp_over_time <- read_csv(here("output", "monthly_op_gp", "adhd_gp_total.csv"))

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)

total_foi_over_time <- read_csv(here("output", "foi_prescribing", "adhd_foi_total.csv"))

scmd_adhd_issuing <- read_csv(
  here(
    "output",
    "web-openprescribing-hospitals",
    "adhd_oph_web_ddd_quantity.csv"
    )
) |> 
  janitor::clean_names() |> 
  rename(
    month = date,
    bnf_chemical_name = vtm_name,
    icb_name = icb
  ) |> 
  # Filter for only months in our analysis period
  filter(between(month, as.Date("2019-04-01"), as.Date("2025-03-01")))

total_scmd_over_time <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_total.csv")) |> 
  # Filter for only months in our analysis period
  filter(between(month, as.Date("2019-04-01"), as.Date("2025-03-01")))

# Read in ICB code -> ICB name mapping
icb_code_name_map <- read_csv(here("output", "icb_mapping", "icb_code_name_map.csv"))
```

# Results: Yearly medication coding 

## Prescribing/issuing by chemical (yearly)

Will need to interpret the first year or so carefully because of missing data.

```{r}
# Helper function for calculating yearly totals from Aug 1st to Jul 31st
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
calc_yearly_total <- function(data, yearly_measure, monthly_measure){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jul 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by(bnf_chemical_name, end_date) |> 
    summarise(
      !!yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

# GP prescribing
df_yearly_gp_by_chemical <- calc_yearly_total(
  total_gp_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_gp_by_chemical <- plot_med_usage_shapes(
  data = df_yearly_gp_by_chemical, 
  usage_measure = yearly_items, 
  title_label = "GP Prescribing (Yearly Items)",
  n_breaks = 10
)

fig_yearly_gp_by_chemical

# FOI prescribing
df_yearly_foi_by_chemical <- calc_yearly_total(
  total_foi_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_foi_by_chemical <- plot_med_usage_shapes(
  data = df_yearly_foi_by_chemical,  
  usage_measure = yearly_items, 
  title_label = "FOI Prescribing (Yearly Items)",
  n_breaks = 10
)

fig_yearly_foi_by_chemical

# SCMD issuing - we want to aggregate from Apr 1st to Mar 31st to enable comparisons with ICD-10 code usage
df_yearly_scmd_by_chemical <- total_scmd_over_time |> 
  mutate(
    # calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
    # Add end date of reporting year
    end_date = as.Date(paste0(reporting_year, "-03-31"))
  ) |> 
  group_by(bnf_chemical_name, end_date) |> 
  summarise(
    yearly_ddd_quantity = sum(monthly_ddd_quantity, na.rm = TRUE), .groups = "drop"
  )

fig_yearly_scmd_by_chemical <- plot_med_usage_shapes(
  data = df_yearly_scmd_by_chemical, 
  usage_measure = yearly_ddd_quantity, 
  title_label = "SCMD Issuing (Yearly DDD Quantity)",
  n_breaks = 6
)

fig_yearly_scmd_by_chemical
```

Not doing a shared x-axis as the yearly aggregation periods are different. 

### Combined figures

```{r}
combined_yearly_plot <- (fig_yearly_gp_by_chemical / fig_yearly_foi_by_chemical / fig_yearly_scmd_by_chemical) + 
  plot_annotation(tag_levels = "A") +
  # Shared legend
  plot_layout(guides = "collect") &
	theme(
		plot.tag = element_text(size = 20),
    legend.position = "bottom",
		legend.title = element_blank(),
		legend.box = "horizontal"
	)

ggsave(here("output", "figures", "fig_yearly_med_coding.png"), combined_yearly_plot, width = 10, height = 12)
```

# Results: ICB-level analysis

## GP ICB-level analysis

We obtained GP data from 2015-08-01 to 2025-07-31 from BigQuery to match SNOMED CT code usage reporting years. Therefore, we will aggregate from Aug 1st to Jul 31st the next year. Also need to map ICB codes to ICB names for this one.

```{r}
# The `stp` column contains the correct ICB codes as of 2026-03-01
unique(gp_adhd_prescribing$stp)
# 43 unique values including "NA" (unidentified doctors)

# Helper function for calculating ICB yearly totals for GP and FOI data ONLY
icb_yearly_total <- function(data, icb_col, monthly_col, dataset){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jul 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by({{icb_col}}, end_date) |> 
    summarise(
      yearly_items = sum({{monthly_col}}, na.rm = TRUE), .groups = "drop"
    ) |> 
    mutate(
      end_date = as.factor(end_date),
      dataset = dataset # This is useful for summary stats later
    )
}

# Map ICB codes to ICB names
gp_adhd_icb_yearly <- icb_yearly_total(gp_adhd_prescribing, stp, items_by_chemical, "GP") |>  
  rename(icb_code = stp) |>
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_code)

# Helper function for ranking ICBs according to yearly medication usage
# `yearly_col` is the column name of yearly quantity measures (items or ddd quantity)

rank_icb_yearly <- function(data, yearly_col){
  data |> 
    group_by(end_date) |>
    mutate(
      icb_name = coalesce(icb_name, "Unidentified doctors"),
      # Ranking ICBs by yearly items each year
      year_rank = min_rank(desc({{yearly_col}})),
      max_rank = max(year_rank, na.rm = T),
      # Create columns to highlight top/bottom 5 ICBs in the fig
      year_top5 = year_rank <= 5,
      year_bottom5 = year_rank > max_rank - 5
    ) |> 
    # Turn ICB names into title case and remove "NHS" prefix
    mutate(icb_name = icb_name |> 
        str_remove("^NHS\\s*") |>
        str_remove("\\s*Integrated Care Board$") |> 
        str_to_title()
    ) |> 
    ungroup()
}

gp_adhd_icb_ranked <- rank_icb_yearly(gp_adhd_icb_yearly, yearly_items)

# Helper function for creating a legend for top 5 ICBs
# Need to be very specific here because I want the legend as
# "ICB name, (years the ICB was in the top 5) 
create_top_legend <- function(data){
  data |> 
    filter(year_top5 == T) |> 
    mutate(year = format(as.Date(end_date), "%Y")) |> 
    distinct(icb_name, year) |> 
    group_by(icb_name) |> 
    summarise(
      years = paste(sort(unique(year)), collapse = ", "),
      .groups = "drop"
    ) |> 
    mutate(label = paste0(icb_name, "<br>(", years, ")"))
}

# Helper function for creating a legend for bottom 5 ICBs - same logic as above

create_bottom_legend <- function(data){
  data |> 
    filter(year_bottom5 == T) |> 
    mutate(year = format(as.Date(end_date), "%Y")) |> 
    distinct(icb_name, year) |> 
    group_by(icb_name) |> 
    summarise(
      years = paste(sort(unique(year)), collapse = ", "),
      .groups = "drop"
    ) |> 
    mutate(label = paste0(icb_name, "<br>(", years, ")"))
}

top_gp_legend <- create_top_legend(gp_adhd_icb_ranked)
bottom_gp_legend <- create_bottom_legend(gp_adhd_icb_ranked)

# Create violin plots
# `plot_yearly_violins` is in `medication_plots.R`
gp_shape_values <- c(21, 22, 23, 24, 25, 8)
fig_gp_icb_violins <- plot_yearly_violins(
  data = gp_adhd_icb_ranked,
  usage_measure = yearly_items,
  top_legend = top_gp_legend,
  bottom_legend = bottom_gp_legend,
  shape_values = gp_shape_values,
  title_label = "Yearly GP Prescribing by ICB"
)

fig_gp_icb_violins

ggsave(here("output", "figures", "fig_gp_icb_violins.png"), fig_gp_icb_violins, width = 15, height = 10)
```

## FOI ICB-level analysis

FOI data prior to 2020-04-01 is given with Area Team codes instead of ICB codes. Without practice-level information, it's impossible to map Area Team onto ICB accurately. Therefore, I will only analyse FOI data from 2020-08-01 to 2025-07-31 (to match with SNOMED CT code usage reporting years).

```{r}
foi_adhd_20to25 <- foi_adhd_prescribing |> 
  filter(month >= "2020-08-01" & month < "2025-08-01") |> 
  # Using the mapping file to ensure consistent formatting of ICB names
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_name.x) |> 
  rename(icb_name = icb_name.y)

setequal(na.omit(foi_adhd_20to25$icb_name), icb_code_name_map$icb_name)
# ICB names match the mapping file but FOI data includes "Unidentified doctors")

# Calculate yearly prescribing for each ICB
foi_adhd_icb_yearly <- icb_yearly_total(foi_adhd_20to25, icb_name, items, "FOI")
# Rank ICBs according to yearly prescribing
foi_adhd_icb_ranked <- rank_icb_yearly(foi_adhd_icb_yearly, yearly_items)

# Create top and bottom legends
top_foi_legend <- create_top_legend(foi_adhd_icb_ranked)
bottom_foi_legend <- create_bottom_legend(foi_adhd_icb_ranked)

# Create violin plot
foi_shape_values <- c(21, 22, 23, 24, 25, 8, 11)

fig_foi_icb_violins <- plot_yearly_violins(
  data = foi_adhd_icb_ranked,
  usage_measure = yearly_items,
  top_legend = top_foi_legend,
  bottom_legend = bottom_foi_legend,
  shape_values = foi_shape_values,
  title_label = "Yearly FOI Prescribing by ICB"
)

fig_foi_icb_violins

ggsave(here("output", "figures", "fig_foi_icb_violins.png"), fig_foi_icb_violins, width = 15, height = 10)
```

## SCMD ICB-level analysis

SCMD data is only available from 2019-01-01 to 2025-12-31. To match analysis time periods with ICD-10 code usage reporting years, I will only analyse SCMD hospital issuing from 2019-04-01 to 2025-03-31. 

```{r}
# ICB names in SCMD essentially match the mapping file but are formatted differently
scmd_adhd_issuing_caps <- scmd_adhd_issuing |> 
  mutate(
    icb_name = paste0(icb_name, " Integrated Care Board") |> 
    str_to_upper() 
  )
setequal(scmd_adhd_issuing_caps$icb_name, icb_code_name_map$icb_name)
# [1] TRUE

# Calculate yearly DDD quantity for each ICB
# Can't use `icb_yearly_total` as I am aggregating from April 1st to March 31st but the code follows the same logic
scmd_adhd_icb_yearly <- scmd_adhd_issuing |> 
  filter(month >= "2019-04-01" & month < "2025-04-01") |> 
  mutate(
    # calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-03-31"))
    ) |> 
  group_by(icb_name, end_date) |> 
  summarise(yearly_ddd_quantity = sum(value, na.rm = TRUE), .groups = "drop") |> 
  # `end_date` needs to be a factor for the violin plot
  mutate(
    end_date = as.factor(end_date),
    dataset = "SCMD" # This is useful for summary stats later
  )

# Rank ICBs by yearly DDD quantity
scmd_adhd_icb_ranked <- rank_icb_yearly(scmd_adhd_icb_yearly, yearly_ddd_quantity)

# Create top and bottom legends
top_scmd_legend <- create_top_legend(scmd_adhd_icb_ranked)
bottom_scmd_legend <- create_bottom_legend(scmd_adhd_icb_ranked)

# Create violin plot
scmd_shape_values <- c(21, 22, 23, 24, 25, 8, 11)

fig_scmd_icb_violins <- plot_yearly_violins(
  data = scmd_adhd_icb_ranked,
  usage_measure = yearly_ddd_quantity,
  top_legend = top_scmd_legend,
  bottom_legend = bottom_scmd_legend,
  shape_values = scmd_shape_values,
  title_label = "Yearly SCMD Issuing by ICB"
)

fig_scmd_icb_violins

ggsave(here("output", "figures", "fig_scmd_icb_violins.png"), fig_scmd_icb_yearly, width = 15, height = 10)
```

## Summary statistics

Can have the following:

- Total prescribing/issuing quantity for each data source for the entire analysis period (for every single month or to match yearly totals? currently the stats below include all SCMD data which doesn't match yearly aggregated data because there are some extra months) and percentage change

```{r}
# Prepare GP, FOI, and SCMD dataframes to be combined
gp_monthly_summary <- total_gp_over_time |> 
  mutate(dataset = "GP", unit = "items") |> 
  rename(monthly_value = monthly_items)

foi_monthly_summary <- total_foi_over_time |> 
  mutate(dataset = "FOI", unit = "items") |> 
  rename(monthly_value = monthly_items)

scmd_monthly_summary <- total_scmd_over_time |> 
  mutate(dataset = "SCMD", unit = "ddd_quantity") |> 
  rename(monthly_value = monthly_ddd_quantity)

# Create summary table for total usage
df_total_summary <- bind_rows(gp_monthly_summary, foi_monthly_summary, scmd_monthly_summary, .id = NULL) |> 
  filter(bnf_chemical_name == "All chemicals") |> 
  group_by(dataset) |> 
  mutate(
    # Start and end dates for each dataset - can change if appropriate
    start_date = min(month),
    start_value = monthly_value[which.min(month)],
    end_date = max(month),
    end_value = monthly_value[which.max(month)],
    # Total prescribing/issuing across time period 
    total = sum(monthly_value)
  ) |> 
  select(-c(month, monthly_value, bnf_chemical_name)) |> 
  distinct() |> 
  # Percentage change of last month with data compared to first month with data
  # Note that this is not necessarily 2015-08-01 as some chemicals were not recorded at first
  # Have to pick the first month with data as dividing by 0 doesn't work!
  mutate(pct_change = round((end_value - start_value) / start_value * 100, digits = 2))

print(df_total_summary$total)
```

### Percentage of FOI/GP prescribing for each chemical

Need to think about how I do this, as FOI data for atomoxetine and guanfacine is missing for the first few months. Currently doing first year vs last year comparison (but I might want to do second year vs last year, given that the first year won't be complete?).

```{r}
# Same analysis time period: 2015-08-01 to 2025-07-01
df_foi_gp_pct <- bind_rows(df_yearly_gp_by_chemical, df_yearly_foi_by_chemical, .id = "dataset") |> 
  mutate(
    dataset = recode(dataset, "1" = "GP", "2" = "FOI")
  ) |> 
  group_by(dataset, bnf_chemical_name) |> 
  mutate(
    start_value = yearly_items[which.min(end_date)],
    end_value = yearly_items[which.max(end_date)],
    period_dataset_total = sum(yearly_items)
  ) |> 
  ungroup() |>
  select(-c(end_date, yearly_items)) |> 
  distinct() |> 
  group_by(bnf_chemical_name) |> 
  mutate(
    start_total = sum(start_value),
    start_pct = round(start_value / start_total * 100, digits = 2),
    end_total = sum(end_value),
    end_pct = round(end_value / end_total * 100, digits = 2),
    period_total = sum(period_dataset_total),
    period_pct = round(period_dataset_total / period_total * 100, digits = 2)
  ) |>
  select(-c(start_value, end_value, start_total, end_total, period_total, period_dataset_total)) |> 
  arrange(bnf_chemical_name)
```

### Top total prescribing/issuing ICBs across all years

```{r}
# Prepare dfs for combination
gp_icb_yearly_summary <- gp_adhd_icb_yearly |> 
  mutate(unit = "items") |> 
  rename(yearly_value = yearly_items)

foi_icb_yearly_summary <- foi_adhd_icb_yearly |> 
  mutate(unit = "items") |> 
  rename(yearly_value = yearly_items)

scmd_icb_yearly_summary <- scmd_adhd_icb_yearly |> 
  mutate(unit = "ddd_quantity") |> 
  rename(yearly_value = yearly_ddd_quantity)

# Summary df with all datasets
icb_yearly_summary <- bind_rows(gp_icb_yearly_summary, foi_icb_yearly_summary, scmd_icb_yearly_summary) |> 
  # Need to turn factors back into dates
  mutate(end_date = as.Date(end_date))

df_icbs_ranked <- icb_yearly_summary |> 
  group_by(dataset, icb_name) |> 
  # Total prescribing/issuing across entire analysis period for each dataset
  mutate(total = sum(yearly_value)) |> 
  ungroup() |> 
  select(-c(end_date, yearly_value)) |> 
  distinct() |> 
  group_by(dataset) |> 
  mutate(
    icb_name = coalesce(icb_name, "Unidentified doctors"),
    # Ranking ICBs by total prescribing/issuing
    rank = min_rank(desc(total))
  ) |> 
  ungroup()

df_icbs_summary <- df_icbs_ranked |> 
  group_by(dataset) |> 
  mutate(
    all_icbs_total = sum(total),
    top_icb = icb_name[which.min(rank)],
    bottom_icb = icb_name[which.max(rank)],
    unidentified_rank = first(rank[icb_name == "Unidentified doctors"], default = NA_integer_),
    # Calculate quartiles and IQR
    q1 = quantile(total, probs = 0.25, na.rm = TRUE),
    median = quantile(total, probs = 0.5, na.rm = TRUE),
    q3 = quantile(total, probs = 0.75, na.rm = TRUE),
    iqr = q3 - q1
  ) |> 
  select(-c(icb_name, total, rank)) |> 
  distinct()
  
df_icbs_summary
```

### Percentage of GP and FOI prescribing by unidentified doctors (N/A for SCMD)

```{r}
df_unidentified <- icb_yearly_summary |> 
  # No unidentified issuing for SCMD so filtering out the dataset
  filter_out(dataset == "SCMD") |> 
  mutate(icb_name = if_else(is.na(icb_name), "unidentified", "identified")) |> 
  group_by(dataset, icb_name) |> 
  summarise(total_value = sum(yearly_value), .groups = "drop") |> 
  group_by(dataset) |> 
  summarise(pct_unidentified = (total_value[icb_name == "unidentified"] / sum(total_value)) * 100, .groups = "drop")

df_unidentified
```

Unidentified doctors contributed to 0.468% and 0.972% of total GP and FOI prescribing respectively. 
Note: no GP prescribing from unidentified doctors after 2020-07-31.

### ICB quartiles for the first and last year (GP, FOI, SCMD)

```{r}
df_first_last_icb_quartiles <- icb_yearly_summary |>  
  group_by(dataset) |> 
  # Only keep rows for the first and last year of each dataset
  filter(end_date == min(end_date) | end_date == max(end_date)) |> 
  group_by(dataset, end_date) |> 
  mutate(
    # Calculate total yearly value
    total_yearly_value = sum(yearly_value),
    # Calculate quartiles and IQR
    q1 = quantile(yearly_value, probs = 0.25, na.rm = TRUE),
    median = quantile(yearly_value, probs = 0.5, na.rm = TRUE),
    q3 = quantile(yearly_value, probs = 0.75, na.rm = TRUE),
    iqr = q3 - q1
  ) |> 
  ungroup() |> 
  select(-c(yearly_value, icb_name)) |> 
  distinct()
  
```

These quartiles include unidentified doctors for GP and FOI prescribing. 