---
title: "ADHD Medications Descriptive Analysis and Results"
author:
  - Kaylee Poon
format: html
---
# Overview

This analysis explores prescribing/issuing data for ADHD medications using the following data sources:

1. GP prescribing data (OpenPrescribing) from 2015-08-01 to 2025-07-01 (monthly items). 
2. FOI prescribing data from 2015-08-01 to 2025-07-01 (monthly items).
3. SCMD issuing data (OpenPrescribing-Hospitals website) from 2019-01-01 to 2025-12-01 (monthly DDD quantity).

I will produce the following figures for the thesis: 

- Annual total ADHD medication prescriptions from each data source

- Regional trends and variations in ADHD medications in primary care (OpenPrescribing), secondary care (OpenPrescribing-Hospitals), and private care (NHS BSA FOI)

# Methods
```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(nhstt)
library(arrow)
library(patchwork)
library(ggtext)

source(here("functions", "medication_plots.R"))


gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
    )
)

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)

scmd_adhd_issuing <- read_csv(
  here(
    "output",
    "web-openprescribing-hospitals",
    "adhd_oph_web_ddd_quantity.csv"
    )
) |> 
  janitor::clean_names() |> 
  rename(
    month = date,
    bnf_chemical_name = vtm_name
  ) 
```

# Results: medication usage


## Prescribing/issuing by chemical (monthly)
```{r}
# fig_total_gp_over_time <- adhd_gp_prescribing |> 
#   select(month, items_by_chemical) |>
#   group_by(month, bnf_chemical_name) |> 
#   summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop")

# cheating here by borrowing total_gp_over_time from `prescribing_changepoint.qmd` 
fig_monthly_gp_by_chemical <- plot_med_usage(total_gp_over_time, monthly_items, "Monthly items", "GP Prescribing")

fig_monthly_gp_by_chemical

fig_monthly_foi_by_chemical <- plot_med_usage(total_foi_over_time, monthly_items, "Monthly items", "FOI Prescribing")

fig_monthly_foi_by_chemical

fig_monthly_scmd_by_chemical <- plot_med_usage(total_scmd_over_time, monthly_ddd_quantity, "Monthly DDD quantity", "SCMD issuing")

fig_monthly_scmd_by_chemical

combined_monthly_plot <- ((fig_monthly_gp_by_chemical / fig_monthly_foi_by_chemical) | plot_spacer() | fig_monthly_scmd_by_chemical) + 
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_monthly_plot

ggsave(here("output", "figures", "fig_monthly_med_usage.png"), combined_monthly_plot, width = 15, height = 15)
```

I find that if I'm plotting monthly data, line types work a bit better than shapes. Although sometimes it's still a bit difficult to tell indiviudal lines apart. Also I need to do something about the breaks.

Might be good to do some % calculations for each chemical? Also think about if I should aggregate this into yearly data instead. 

## Prescribing/issuing by chemical (yearly)

Will need to interpret the first year or so carefully because of missing data.

```{r}
# Helper function for calculating yearly totals
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
calc_yearly_total <- function(data, yearly_measure, monthly_measure){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jun 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by(bnf_chemical_name, end_date) |> 
    summarise(
      !!yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

# GP prescribing
df_yearly_gp_by_chemical <- calc_yearly_total(
  total_gp_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_gp_by_chemical <- plot_med_usage_shapes(df_yearly_gp_by_chemical, end_date, yearly_items, "GP Prescribing")

fig_yearly_gp_by_chemical

# FOI prescribing
df_yearly_foi_by_chemical <- calc_yearly_total(
  total_foi_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_foi_by_chemical <- plot_med_usage_shapes(df_yearly_foi_by_chemical, end_date, yearly_items, "FOI Prescribing")

fig_yearly_foi_by_chemical

# SCMD issuing
df_yearly_scmd_by_chemical <- calc_yearly_total(
  total_scmd_over_time,
  "yearly_ddd_quantity",
  monthly_ddd_quantity
)

fig_yearly_scmd_by_chemical <- plot_med_usage_shapes(df_yearly_scmd_by_chemical, end_date, yearly_ddd_quantity, "SCMD Prescribing")

fig_yearly_scmd_by_chemical

combined_yearly_plot <- ((fig_yearly_gp_by_chemical / fig_yearly_foi_by_chemical) | plot_spacer() | (fig_yearly_scmd_by_chemical / fig_total_gp_foi)) + 
  plot_annotation(title = "Version with yearly data", tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_yearly_plot

ggsave(here("output", "figures", "fig_yearly_med_usage.png"), combined_yearly_plot, width = 15, height = 15)
```

It's now easier to distinguish between different lines. Will need to fix x-axis tick labelling if we end up reporting yearly data at this point. Can consider having a bar plot for proportion of items prescribed in GP vs FOI data for each chemical. 

### Percentage stacked bar chart for GP and FOI prescribing (total prescribing by chemical)

```{r}
df_total_gp_foi <- bind_rows(df_yearly_gp_by_chemical, df_yearly_foi_by_chemical, .id = "source") |> 
  mutate(
    source = recode(source, "1" = "GP", "2" = "FOI")
  ) |> 
  group_by(bnf_chemical_name, source) |> 
  summarise(total_items = sum(yearly_items), na.rm = T, .groups = "drop")

fig_total_gp_foi <- ggplot(
  df_total_gp_foi,
  aes(fill = source, y = total_items, x = bnf_chemical_name)
) +
  # Stacked percentage bar chart
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
  scale_fill_viridis_d(begin = 0.25, end = 0.5) +
  labs(title = "Percentage of GP/FOI Prescribing by Chemical") +
  theme_bw(base_size = 12) +
  theme(text = element_text(family = "Times New Roman"))

fig_total_gp_foi
```

Kind of ugly but does the job. Can add percentage labels. 

## Prescribing/issuing by chemical (half-yearly)

```{r}
# Helper function for calculating half-yearly totals
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
calc_halfyearly_total <- function(data, half_yearly_measure, monthly_measure){
  data |> 
    # calculate reporting period (Aug 1st - Jan 31st; Feb 1st - Jul 31st) from month
    mutate( 
      period_start = case_when(
      month(month) %in% 8:12 ~ as.Date(paste0(year(month), "-08-01")),
      month(month) == 1 ~ as.Date(paste0(year(month) + 1, "-08-01")),
      month(month) %in% 2:7 ~ as.Date(paste0(year(month), "-02-01"))
      ),
      period_end = case_when(
      month(month) %in% 8:12 ~ as.Date(paste0(year(month) + 1, "-01-31")),
      month(month) == 1 ~ as.Date(paste0(year(month), "-01-31")),
      month(month) %in% 2:7 ~ as.Date(paste0(year(month), "-07-31"))
      )
    ) |> 
    group_by(bnf_chemical_name, period_end) |> 
    summarise(
      !!half_yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

# GP prescribing

df_halfy_gp_by_chemical <- calc_halfyearly_total(total_gp_over_time, "half_yearly_items", monthly_items)

fig_halfy_gp_by_chemical <- plot_med_usage_shapes(df_halfy_gp_by_chemical, period_end, half_yearly_items, "GP Prescribing")

fig_halfy_gp_by_chemical

# FOI prescribing

df_halfy_foi_by_chemical <- calc_halfyearly_total(total_foi_over_time, "half_yearly_items", monthly_items)

fig_halfy_foi_by_chemical <- plot_med_usage_shapes(df_halfy_foi_by_chemical, period_end, half_yearly_items, "FOI Prescribing")

fig_halfy_foi_by_chemical

# SCMD issuing

df_halfy_scmd_by_chemical <- calc_halfyearly_total(total_scmd_over_time, "half_yearly_ddd_quantity", monthly_ddd_quantity)

fig_halfy_scmd_by_chemical <- plot_med_usage_shapes(df_halfy_scmd_by_chemical, period_end, half_yearly_ddd_quantity, "SCMD issuing")

fig_halfy_scmd_by_chemical

combined_half_yearly_plot <- ((fig_halfy_gp_by_chemical / fig_halfy_foi_by_chemical) | plot_spacer() | (fig_halfy_scmd_by_chemical / fig_total_gp_foi)) + 
  plot_annotation(title = "Version with half-yearly data", tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_half_yearly_plot

ggsave(here("output", "figures", "fig_half_yearly_med_usage.png"), combined_half_yearly_plot, width = 15, height = 15)

```

