---
title: "ADHD Medications Descriptive Analysis and Results"
author:
  - Kaylee Poon
format: html
---
# Overview

This analysis explores prescribing/issuing data for ADHD medications using the following data sources:

1. GP prescribing data (OpenPrescribing) from 2015-08-01 to 2025-07-01 (monthly items). 
2. FOI prescribing data from 2015-08-01 to 2025-07-01 (monthly items).
3. SCMD issuing data (OpenPrescribing-Hospitals website) from 2019-01-01 to 2025-12-01 (monthly DDD quantity).

I will produce the following figures for the thesis: 

- Annual total ADHD medication prescriptions from each data source

- Regional trends and variations in ADHD medications in primary care (OpenPrescribing), secondary care (OpenPrescribing-Hospitals), and private care (NHS BSA FOI)

# Methods
```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(nhstt)
library(arrow)
library(patchwork)
library(ggtext)
library(gghighlight)
library(ggnewscale)

source(here("functions", "medication_plots.R"))

# Read in data files
gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
    )
)

total_gp_over_time <- read_csv(here("output", "monthly_op_gp", "adhd_gp_total.csv"))

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)

total_foi_over_time <- read_csv(here("output", "foi_prescribing", "adhd_foi_total.csv"))

scmd_adhd_issuing <- read_csv(
  here(
    "output",
    "web-openprescribing-hospitals",
    "adhd_oph_web_ddd_quantity.csv"
    )
) |> 
  janitor::clean_names() |> 
  rename(
    month = date,
    bnf_chemical_name = vtm_name,
    icb_name = icb
  ) 

total_scmd_over_time <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_total.csv"))

# Read in ICB code -> ICB name mapping
icb_code_name_map <- read_csv(here("output", "icb_mapping", "icb_code_name_map.csv"))
```

# Results: medication usage


## Prescribing/issuing by chemical (monthly)
```{r}
fig_monthly_gp_by_chemical <- plot_med_usage(total_gp_over_time, monthly_items, "Monthly items", "GP Prescribing")

fig_monthly_gp_by_chemical

fig_monthly_foi_by_chemical <- plot_med_usage(total_foi_over_time, monthly_items, "Monthly items", "FOI Prescribing")

fig_monthly_foi_by_chemical

fig_monthly_scmd_by_chemical <- plot_med_usage(total_scmd_over_time, monthly_ddd_quantity, "Monthly DDD quantity", "SCMD issuing")

fig_monthly_scmd_by_chemical
```

I find that if I'm plotting monthly data, line types work a bit better than shapes. Although sometimes it's still a bit difficult to tell indiviudal lines apart. Also I need to do something about the breaks.

Might be good to do some % calculations for each chemical? Also think about if I should aggregate this into yearly data instead. 

## Prescribing/issuing by chemical (yearly)

Will need to interpret the first year or so carefully because of missing data.

```{r}
# Helper function for calculating yearly totals from Aug 1st to Jul 31st
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
calc_yearly_total <- function(data, yearly_measure, monthly_measure){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jul 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by(bnf_chemical_name, end_date) |> 
    summarise(
      !!yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

# GP prescribing
df_yearly_gp_by_chemical <- calc_yearly_total(
  total_gp_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_gp_by_chemical <- plot_med_usage_shapes(df_yearly_gp_by_chemical, end_date, yearly_items, "GP Prescribing")

fig_yearly_gp_by_chemical

# FOI prescribing
df_yearly_foi_by_chemical <- calc_yearly_total(
  total_foi_over_time,
  "yearly_items",
  monthly_items
)

fig_yearly_foi_by_chemical <- plot_med_usage_shapes(df_yearly_foi_by_chemical, end_date, yearly_items, "FOI Prescribing")

fig_yearly_foi_by_chemical

# SCMD issuing - we want to aggregate from Apr 1st to Mar 31st to enable comparisons with ICD-10 code usage
df_yearly_scmd_by_chemical <- total_scmd_over_time |> 
  mutate(
    # calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
    # Add end date of reporting year
    end_date = as.Date(paste0(reporting_year, "-03-31"))
  ) |> 
  group_by(bnf_chemical_name, end_date) |> 
  summarise(
    yearly_ddd_quantity = sum(monthly_ddd_quantity, na.rm = TRUE), .groups = "drop"
  )

fig_yearly_scmd_by_chemical <- plot_med_usage_shapes(df_yearly_scmd_by_chemical, end_date, yearly_ddd_quantity, "SCMD Prescribing")

fig_yearly_scmd_by_chemical
```

It's now easier to distinguish between different lines. Will need to fix x-axis tick labelling if we end up reporting yearly data at this point. 

### Percentage stacked bar chart for GP and FOI prescribing (total prescribing by chemical)

```{r}
df_total_gp_foi <- bind_rows(df_yearly_gp_by_chemical, df_yearly_foi_by_chemical, .id = "source") |> 
  mutate(
    source = recode(source, "1" = "GP", "2" = "FOI")
  ) |> 
  group_by(bnf_chemical_name, source) |> 
  summarise(total_items = sum(yearly_items), na.rm = T, .groups = "drop")

fig_total_gp_foi <- ggplot(
  df_total_gp_foi,
  aes(fill = source, y = total_items, x = bnf_chemical_name)
) +
  # Stacked percentage bar chart
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
  scale_fill_viridis_d(begin = 0.25, end = 0.5) +
  labs(title = "Percentage of GP/FOI Prescribing by Chemical") +
  theme_bw(base_size = 12) +
  theme(text = element_text(family = "Times New Roman"))

fig_total_gp_foi
```

Kind of ugly but does the job. Can add percentage labels. 

### Combined figures

```{r}
combined_monthly_plot <- ((fig_monthly_gp_by_chemical / fig_monthly_foi_by_chemical) | plot_spacer() | fig_monthly_scmd_by_chemical / fig_total_gp_foi) + 
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_monthly_plot

ggsave(here("output", "figures", "fig_monthly_med_usage.png"), combined_monthly_plot, width = 15, height = 15)

combined_yearly_plot <- ((fig_yearly_gp_by_chemical / fig_yearly_foi_by_chemical) | plot_spacer() | (fig_yearly_scmd_by_chemical / fig_total_gp_foi)) + 
  plot_annotation(title = "Version with yearly data", tag_levels = "A") +
  plot_layout(widths = c(1, 0.05, 1))

combined_yearly_plot

ggsave(here("output", "figures", "fig_yearly_med_usage.png"), combined_yearly_plot, width = 15, height = 15)
```


# Results: ICB-level analysis

Create dataframe with the total number of prescribed ADHD items (all chemicals) per ICB
```{r}
unique(scmd_adhd_issuing$icb_name)
# 42 unique values which seem to match current ICBs 

# aggregating from april 1st to march 31st each year - will only have data from 2019/20 to 2024-25
scmd_adhd_icb_yearly <- scmd_adhd_issuing |> 
  mutate(
    # calculate reporting year (Apr 1st - Mar 31st) from month 
    reporting_year = if_else(month(month) >= 4, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-03-31"))
    ) |> 
  group_by(icb_name, end_date) |> 
  summarise(yearly_ddd_quantity = sum(value, na.rm = TRUE), .groups = "drop") |> 
  # Removing data before 2019-03-31 and after 2025-04-01 as the years are incomplete
  filter_out(end_date == c("2019-03-31", "2026-03-31")) |> 
  # `end_date` needs to be a factor for the violin plot
  mutate(end_date = as.factor(end_date)) |> 
  group_by(end_date) |> 
  mutate(
    # Ranking ICBs by yearly DDD quantity each year
    year_rank = min_rank(desc(yearly_ddd_quantity)),
    max_rank = max(year_rank, na.rm = T),
    # Create columns to highlight top/bottom 5 ICBs in the fig
    year_top5 = year_rank <= 5,
    year_bottom5 = year_rank > max_rank - 5
  ) |> 
  mutate(icb_name = icb_name |> 
    str_remove("^NHS\\s*") |>
    str_remove("\\s*Integrated Care Board$") |> 
    str_to_title()
  ) |> 
  ungroup()

# Top 5 legend
# Need to be very specific here because I want the legend as
# "ICB name, (years the ICB was in the top five)"
top_scmd_legend <- scmd_adhd_icb_yearly |> 
  filter(year_top5 == T) |> 
  mutate(year = format(as.Date(end_date), "%Y")) |> 
  distinct(icb_name, year) |> 
  group_by(icb_name) |> 
  summarise(
    years = paste(sort(unique(year)), collapse = ", "),
    .groups = "drop"
  ) |> 
  mutate(label = paste0(icb_name, "<br>(", years, ")"))

bottom_scmd_legend <- scmd_adhd_icb_yearly |> 
  filter(year_bottom5 == T) |> 
  mutate(year = format(as.Date(end_date), "%Y")) |> 
  distinct(icb_name, year) |> 
  group_by(icb_name) |> 
  summarise(
    years = paste(sort(unique(year)), collapse = ", "),
    .groups = "drop"
  ) |> 
  mutate(label = paste0(icb_name, "<br>(", years, ")"))

# violin plot
scmd_shape_values <- c(21, 22, 23, 24, 25, 8, 11)

fig_scmd_icb_yearly <- plot_yearly_violins(
  data = scmd_adhd_icb_yearly,
  usage_measure = yearly_ddd_quantity,
  top_legend = top_scmd_legend,
  bottom_legend = bottom_scmd_legend,
  shape_values = scmd_shape_values,
  title_label = "Yearly SCMD Issuing by ICB"
)

fig_scmd_icb_yearly

ggsave(here("output", "figures", "fig_scmd_icb_yearly.png"), fig_scmd_icb_yearly, width = 15, height = 10)
```

## GP ICB-level analysis

Need to map ICB codes to ICB names!
```{r}
# The `stp` column contains the correct ICB codes as of 2026-03-01
unique(gp_adhd_prescribing$stp)
# 43 unique values including "NA" (unidentified doctors)

# Helper function for calculating ICB yearly totals for GP and FOI data ONLY
# Pass `yearly_measure` as a string and `monthly_measure` as a variable
icb_yearly_total <- function(data, icb_col, yearly_measure, monthly_measure){
  data |> 
    mutate(
      # calculate reporting year (Aug 1st - Jun 31st) from month 
      reporting_year = if_else(month(month) >= 8, year(month) + 1, year(month)),
      # Add end date of reporting year
      end_date = as.Date(paste0(reporting_year, "-07-31"))
    ) |> 
    group_by({{icb_col}}, end_date) |> 
    summarise(
      !!yearly_measure := sum({{monthly_measure}}, na.rm = TRUE), .groups = "drop"
  )
}

gp_adhd_icb <- icb_yearly_total(gp_adhd_prescribing, stp, "yearly_items", items_by_chemical)

# This is the same logic as SCMD above but with the mapping and slightly different column names
# Can be turned into a function with some work
gp_adhd_icb_yearly <- gp_adhd_icb |> 
  # `end_date` needs to be a factor for the violin plot
  mutate(end_date = as.factor(end_date)) |> 
  group_by(end_date) |> 
  rename(icb_code = stp) |>
  left_join(icb_code_name_map, by = "icb_code") |> 
  select(-icb_code) |> 
  mutate(
    icb_name = coalesce(icb_name, "Unidentified doctors"),
    # Ranking ICBs by yearly items each year
    year_rank = min_rank(desc(yearly_items)),
    max_rank = max(year_rank, na.rm = T),
    # Create columns to highlight top/bottom 5 ICBs in the fig
    year_top5 = year_rank <= 5,
    year_bottom5 = year_rank > max_rank - 5
  ) |> 
  # Turn ICB names into title case and remove "NHS" prefix
  mutate(icb_name = icb_name |> 
      str_remove("^NHS\\s*") |>
      str_remove("\\s*Integrated Care Board$") |> 
      str_to_title()
  ) |> 
  ungroup()

top_gp_legend <- gp_adhd_icb_yearly |> 
  filter(year_top5 == T) |> 
  mutate(year = format(as.Date(end_date), "%Y")) |> 
  distinct(icb_name, year) |> 
  group_by(icb_name) |> 
  summarise(
    years = paste(sort(unique(year)), collapse = ", "),
    .groups = "drop"
  ) |> 
  mutate(label = paste0(icb_name, "<br>(", years, ")"))

bottom_gp_legend <- gp_adhd_icb_yearly |> 
  filter(year_bottom5 == T) |> 
  mutate(year = format(as.Date(end_date), "%Y")) |> 
  distinct(icb_name, year) |> 
  group_by(icb_name) |> 
  summarise(
    years = paste(sort(unique(year)), collapse = ", "),
    .groups = "drop"
  ) |> 
  mutate(label = paste0(icb_name, "<br>(", years, ")"))

gp_shape_values <- c(21, 22, 23, 24, 25, 8)
fig_gp_icb_yearly <- plot_yearly_violins(
  data = gp_adhd_icb_yearly,
  usage_measure = yearly_items,
  top_legend = top_gp_legend,
  bottom_legend = bottom_gp_legend,
  shape_values = gp_shape_values,
  title_label = "Yearly GP Prescribing by ICB"
)

fig_gp_icb_yearly

ggsave(here("output", "figures", "fig_gp_icb_yearly.png"), fig_gp_icb_yearly, width = 15, height = 10)
```

## FOI ICB-level analysis

```{r}
unique(foi_adhd_icb$icb_code)
# 70 unique values including "NA" (unidentified doctors), will need a mapping file!
```