---
title: "Descriptive analysis results: SNOMED CT and ICD-10 diagnosis coding"
author:
  - Kaylee Poon
format: html
---
# Overview
This descriptive analysis explores: 

1. SNOMED CT coding activity related to ADHD diagnosis in English primary care (2011-2025)
2. ICD-10 coding activity related to ADHD diagnosis in English Hospital Admitted Patient Care. (2012-2025) 

The aim is to understand which clinical codes are used to record ADHD diagnoses and related administrative activity, identify usage patterns over time, and inform the development of codelists for future electronic health record research on this topic.

# Methods

## Set up

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(opencodecounts)
library(patchwork)
library(stringr)
library(flextable)
library(officer)
library(ggtext)

# Function for creating facet plots for age breakdowns
source(here("functions", "create_facet_plots.R"))

# Functions for creating usage plots
source(here("functions", "usage_plots.R"))

# Load SNOMED CT ADHD diagnosis coding activity codelist from OpenCodeCounts
snomed_diag_codelist <- get_codelist("https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/adhd_cod/20250912/")

snomed_diag_usage <- snomed_usage |>
  filter(snomed_code %in% snomed_diag_codelist$code)

# Load ICD-10 ADHD diagnosis coding activity codelist from OpenCodeCounts
icd10_diag_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/kayleepn/adhd-icd-10-test/66fd3a31/")

icd10_diag_usage <- icd10_usage |>
  filter(icd10_code %in% icd10_diag_codelist$code)

# Load ICD-10 usage with age, sex, and diagnosis position breakdowns
icd10_diag_breakdowns <- icd10_usage_breakdowns |>
  filter(icd10_code %in% icd10_diag_codelist$code)

# Set flextable defaults
set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 12,
  font.colour = "black",
  border.colour = "black",
  layout = "autofit",
  width = 1
)
```

## Data source

Data extraction methods are based on the [opencodecounts](https://bennettoxford.github.io/opencodecounts/) R package.

We analyse the publicly available NHSE [SNOMED CT code usage in primary care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care) and [Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity) datasets.

For SNOMED CT, the annual usage count (from 1st August to  31st July) reflects how many times each listed SNOMED CT code was added to a GP electronic health record in England during the reporting period.

For ICD-10, we extended the methods to also extract diagnosis position and demographic breakdowns of the counts from the Hospital Admitted Patient Care data. Activity is captured in Finished Consultant Episodes (FCEs), where each FCE can have up to 20 recorded ICD-10 diagnoses. 

## Codelists

2 codelists indicating ADHD diagnosis for SNOMED CT[^codelist-adhd-snomed] and ICD-10[^codelist-adhd-icd10] respectively were developed using OpenCodelists. Note that the ICD-10 code only has 1 code. 

[^codelist-adhd-snomed]: https://www.opencodelists.org/codelist/user/kayleepn/adhd-diagnosis-snomed-ct-clone2/45183c88/
[^codelist-adhd-icd10]: https://www.opencodelists.org/codelist/user/kayleepn/adhd-icd-10-test/66fd3a31/

### ICD-10 Breakdown variables

**Diagnostic codes**: Only the four-character ICD-10 code "Disturbance of activity and attention (F900)" was included.

**Diagnostic position**: Usage counts represent FCEs with a specific code in any diagnostic position ("all diagnoses") or as the primary reason for admission ("main diagnosis").

**Demographic breakdowns**: Demographic breakdowns are available by age groups (with ages 15-19 reported individually and aggregated here for comparability with other age bands) and sex (male, female, unknown).

# Results

## Yearly usage for all codes (SNOMED CT vs ICD-10)

```{r}
# Get total yearly usage for all codes

snomed_all_codes_usage <- summarise_yearly(snomed_usage, "All SNOMED CT codes")
  
fig_snomed_all_usage <- plot_code_usage(
  data = snomed_all_codes_usage, 
  title = "Yearly SNOMED CT code usage", 
  show_legend = FALSE, 
  show_x = TRUE,
  n_breaks = 14
)

fig_snomed_all_usage

icd10_all_codes_usage <- summarise_yearly(icd10_usage, "All ICD-10 codes")

fig_icd10_all_usage <- plot_code_usage(
  data = icd10_all_codes_usage, 
  title = "Yearly ICD-10 code usage", 
  show_legend = FALSE, 
  show_x = TRUE,
  n_breaks = 13
)
fig_icd10_all_usage
```

### Combine figures

```{r}
fig_all_codes_usage <- fig_snomed_all_usage / fig_icd10_all_usage + 
	plot_layout(guides = "collect") +
	plot_annotation(
		tag_levels = "A",
		theme = theme(
			plot.title = element_text(size = 16), 
			legend.position = "bottom",
			legend.title = element_blank())
	)

fig_all_codes_usage

```

## Yearly ADHD code usage

### SNOMED CT
```{r}
#| label: fig-snomed-adhd-vs-all
#| fig-cap: "SNOMED CT ADHD diagnosis code usage compared to all SNOMED CT codes. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

# Get total yearly usage for top 5 SNOMED CT ADHD diagnosis codes
top5_snomed_diag_codes <- snomed_diag_usage |>
  group_by(snomed_code, description) |> 
  summarise(total_usage = sum(usage), .groups = "drop") |> 
  slice_max(total_usage, n = 5) |> 
  pull(snomed_code)

top5_snomed_usage <- snomed_diag_usage |>
  group_by(end_date, snomed_code, description) |> 
  summarise(usage = sum(usage), .groups = "drop") |> 
  filter(snomed_code %in% top5_snomed_diag_codes) |> 
  # Remove semantic tag in brackets so legend labels don't get too long
  mutate(description = sub(" \\(.*", "", description)) |> 
  dplyr::select(-snomed_code)

# Combing yearly usage for all ADHD codes and top 5 ADHD codes

snomed_adhd_usage <- bind_rows(
  summarise_yearly(snomed_diag_usage, "All ADHD diagnosis-related SNOMED CT codes"),
  top5_snomed_usage
)  

# Plot as graph

# None of these figures are going to look good until the `ggsave()` step because of scaling issues
fig_snomed_adhd_usage <- plot_code_usage(
  data = snomed_adhd_usage, 
  title = "ADHD diagnosis-related SNOMED CT code usage",
  show_legend = TRUE,
  show_x = TRUE,
  n_breaks = 14
) 

fig_snomed_adhd_usage
```

### ICD-10 
```{r}
# There is only one ICD-10 code in the codelist so don't need to deal with top 5 codes
icd10_adhd_usage <- summarise_yearly(
  icd10_diag_usage, 
  "F90.0 Disturbance of activity and attention"
)

# Plot figure
# None of these figures are going to look good until the `ggsave()` step because of scaling issues
fig_icd10_adhd_usage <- plot_code_usage(
  data = icd10_adhd_usage, 
  title = "ADHD diagnosis-related ICD-10 code usage",
  show_legend = FALSE,
  show_x = TRUE,
  n_breaks = 14
) 

fig_icd10_adhd_usage
```

### Combining both graphs with patchwork

```{r}
# Combined figure for SNOMED CT and ICD-10 ADHD code usage
fig_adhd_codes_usage <- fig_snomed_adhd_usage / fig_icd10_adhd_usage + 
  plot_layout(heights = c(1.5, 1)) +
	plot_annotation(tag_levels = "A") &
	theme(
		plot.title = element_text(size = 20),
		legend.title = element_blank()
  )

ggsave(here("output", "figures", "fig_adhd_codes_usage.png"), width = 15, height = 11)
```


## Usage counts for top 5 codes in each codelist

Aim: produce table with total usage (across all years) for all codes and ADHD codes for SNOMED CT (only the top 5 codes) and ICD-10 + proportion of coding activity contributed by each code.

ADHD coding only takes up 0.0068% of all SNOMED CT coding so it makes more sense to calculate the proportion of coding each of the top 5 codes take up within ADHD codes.

```{r}
# Helper function for calculating total usage and each code's contribution to total ADHD coding
# `ref_desc` is the description for the reference codes (all ADHD diagnosis-related codes in this case)
calc_total_usage <- function(data_all, data_adhd, ref_desc, ref_ignore){
  bind_rows(data_all, data_adhd, .id = NULL) |> 
    group_by(description) |> 
    summarise(total_usage = sum(usage), .groups = "drop") |> 
    arrange(desc(total_usage)) |>
    # Calculate percentage of code usage compared to all ADHD code usage 
    mutate(ref = total_usage[description == ref_desc]) |> 
    # Doesn't make sense to calculate all SNOMED usage as a percentage of all ADHD code usage so setting the value in that row to NA
    mutate(percentage = if_else(
      description == ref_ignore,
      NA_character_,
      paste0(round(total_usage/ref * 100, digits = 2), "%")
    )) |> 
    dplyr::select(-ref)
}


df_snomed_total_usage <- calc_total_usage(
  snomed_all_codes_usage,
  snomed_adhd_usage,
  ref_desc = "All ADHD diagnosis-related SNOMED CT codes",
  ref_ignore = "All SNOMED CT codes"
)

df_icd10_total_usage <- calc_total_usage(
  icd10_all_codes_usage,
  icd10_adhd_usage,
  ref_desc = "F90.0 Disturbance of activity and attention",
  ref_ignore = "All ICD-10 codes"
)

# Calculate usage percentage change compared to start of analysis period
# Important: note that "Attention deficit hyperactivity disorder, combined type" only starts being recorded in 2018-19

# Helper function to calculate pct change
calc_pct_change <- function(data_all, data_adhd){
  bind_rows(data_all, data_adhd, .id = NULL) |> 
  group_by(description) |> 
  summarise(
    min_date = min(end_date, na.rm = TRUE),
    max_date = max(end_date, na.rm = TRUE),
    start_usage = usage[end_date == min_date][1],
    end_usage = usage[end_date == max_date][1],
    change = end_usage - start_usage,
    pct_change = paste0(round(change / start_usage * 100, digits = 2), "%"),
    .groups = "drop"
   ) |> 
  select(description, pct_change)
}

# Calculate pct changes for SNOMED CT and ICD-10
df_snomed_pct_change <- calc_pct_change(snomed_all_codes_usage, snomed_adhd_usage)
df_icd10_pct_change <- calc_pct_change(icd10_all_codes_usage, icd10_adhd_usage)

```

### Constructing the usage table

```{r}
# Join total usage and percentage change dfs
df_snomed_usage_table <- inner_join(
  df_snomed_total_usage,
  df_snomed_pct_change
)

df_icd10_usage_table <- inner_join(
  df_icd10_total_usage,
  df_icd10_pct_change
)

# Bind SNOMED CT and ICD10 dfs
df_usage_table <- bind_rows(
  tibble(description = "SNOMED CT Codes (2011-2025)", total_usage = NA, percentage = NA, pct_change = NA),  # header row
  df_snomed_usage_table, 
  tibble(description = "ICD-10 Codes (2012-2025)", total_usage = NA, percentage = NA, pct_change = NA),  # header row
  df_icd10_usage_table
)

# Header rows will need to be reformated after exporting to Word - merging with flextable isn't preserved in Word
# Table works fine in html
usage_table <- flextable(df_usage_table) |> 
  theme_booktabs() |> 
  set_caption("ICD-10 and SNOMED CT Code Usage Summary Table") |> 
  set_header_labels(
    description = "Code description",
    total_usage = "Total code usage",
    percentage = "Percentage of total ADHD diagnosis-related usage",
    pct_change = "Percent change compared to start of analysis period"
  ) |> 
  colformat_char(na_str = "N/A") |> 
  # Formatting header rows for SNOMED CT and ICD-10 codes
  merge_at(i = 1, j = 1:4) |> 
  merge_at(i = 9, j = 1:4) |> 
  bold(i = c(1, 9)) |>
  align(i = c(1, 9), align = "center", part = "body") |> 
  hline(
    i = c(1, 8, 9),
    border = fp_border(color = "#666666", width = 1.5),
    part = "body"
  ) |> 
  footnote(i = 8, j = 1, 
    value = as_paragraph(
      "Note that 'Attention deficit hyperactivity disorder, combined type' was first recorded in the 2018-19 reporting period."
      ),
    ref_symbols = "*",
    part = "body"
  ) 

usage_table

save_as_docx(
  "ICD-10 and SNOMED CT Code Usage Summary Table" = usage_table,
  path = "output/tables/descriptive_tables.docx"
)
```



## ICD-10 breakdowns

The usage table above shows that F900 accounts for 97.73% of all F90 code usage. 

```{r}
# Same code as secondary care breakdown analysis report

df_icd10_diag_positions <- icd10_diag_breakdowns |> 
  filter(breakdown %in% c("all_diagnoses", "main_diagnosis")) |>
  group_by(icd10_code, description, breakdown) |>
  summarise(n = sum(usage)) |> 
  pivot_wider(names_from = breakdown, values_from = n) |>
  arrange(icd10_code)

ratio_all_vs_main <- sum(df_icd10_diag_positions$main_diagnosis, na.rm = TRUE) / sum(df_icd10_diag_positions$all_diagnoses, na.rm = TRUE)

round(ratio_all_vs_main * 100, 2)
# [1] 0.79
```

Main diagnoses represent only 0.79% of all F90 diagnosis codes, indicating that ADHD is rarely the primary reason for inpatient admission. Therefore, the following analyses focus on F900 only and do not differentiate between diagnosis positions.

### Age breakdowns

This code is very similar to the previous secondary care report but I aggregate 5-year age groups into 10-year age groups and plot each age group as one line instead of using facets. This allows easier interpretation and takes up less space.

```{r}
#| label: fig-icd10-age-breakdowns
#| fig-cap: "ICD-10 F900 usage over time by age group (aggregated 10-year age groups). Data source: NHS England [Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity), April 2012 to March 2025."
#| fig-width: 10
#| fig-height: 11

df_icd10_age_10_fig <- icd10_diag_breakdowns |> 
  filter(icd10_code == "F900") |>
  filter(str_starts(breakdown, "age")) |>
  mutate(
    breakdown = fct_collapse(
      breakdown,
      "age_0_9" = c("age_0", "age_1_4", "age_5_9"),
      "age_10_19" = c("age_10_14", "age_15", "age_16", "age_17", "age_18", "age_19"),
      "age_20_29" = c("age_20_24", "age_25_29"),
      "age_30_39" = c("age_30_34", "age_35_39"),
      "age_40_49" = c("age_40_44", "age_45_49"),
      "age_50_59" = c("age_50_54", "age_55_59"),
      "age_60_69" = c("age_60_64", "age_65_69"),
      "age_70_79" = c("age_70_74", "age_75_79"),
      "age_80_89" = c("age_80_84", "age_85_89")
    )
  ) |>
  # Need to also group by end_date to plot over time
  group_by(breakdown, end_date) |>
  summarise(usage = sum(usage), .groups = "drop") |>
  ungroup() |>
  group_by(breakdown) |>
    # Change age group labels from "age_x_y" to "Age x - y" and add usage count for each age group across entire time period 
  mutate(
    breakdown = str_replace(breakdown, "age_(\\d+)_(\\d+)", "Age \\1 - \\2"),
    breakdown = str_replace(breakdown, "age_(\\d+)plus", "Age \\1+"),
    breakdown = paste0("**", breakdown, "**<br>n = ", scales::comma(sum(usage)))
  ) |>
  ungroup()

fig_icd10_age_10_breakdowns <- plot_code_usage(df_icd10_age_10_fig, "ICD-10 Code Usage: F900 Age Breakdowns", "Age group")

fig_icd10_age_10_breakdowns
```

Note that the 10 colours are only distinguishable if I use the 'turbo' option in `viridis`, which is a more colour-friendly version of a rainbow scale. 

### Sex breakdowns

Male diagnoses exceeded female diagnoses until approximately 2024, when female diagnoses surpassed male.
Gender unknown remains low throughout the period.

```{r}
#| label: fig-icd10-sex-breakdowns
#| fig-cap: "ICD-10 F900 usage over time by sex (all diagnosis positions). Data source: NHS England [Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity), April 2012 to March 2025."
#| fig-width: 10
#| fig-height: 4

df_icd10_sex_breakdowns_fig <- icd10_diag_breakdowns |>
  filter(icd10_code == "F900") |>
  filter(breakdown %in% c("male", "female", "gender_unknown")) |>
  group_by(breakdown) |>
  mutate(
    breakdown = str_replace(breakdown, "gender_unknown", "unknown"),
    breakdown = str_to_title(breakdown),
    breakdown = paste0(breakdown, " (n = ", scales::comma(sum(usage)), ")")
  )

fig_icd10_sex_breakdowns <- plot_code_usage(df_icd10_sex_breakdowns_fig, "ICD-10 Code Usage: Sex Breakdowns", "Sex")

fig_icd10_sex_breakdowns
```


### Combined figure for ICD-10 breakdowns

```{r}
fig_icd10_breakdowns <- fig_icd10_age_10_breakdowns / fig_icd10_sex_breakdowns + plot_annotation(tag_levels = "A")

ggsave(here("output", "figures", "fig_icd10_breakdowns.png"), width = 15, height = 10)
```