---
title: "Plots for change point analysis of ADHD prescribing/issuing activity"
author:
  - Kaylee Poon
format: html
---

Aiming to plot actual prescribing/issuing over time with the predicted values (+ 95% confidence intervals) from the change point analysis. Only results for analysing all chemicals combined are plotted. One plot for each of:

1. GP data
2. FOI data
3. SCMD data.

Also contains a summary table for change point model coefficients. 

# Methods

```{r}
#| label: set-up
#| output: false

library(tidyverse)
library(here)
library(patchwork)
library(viridis)
library(flextable)
```

## Read in data

Also doing some preparation for creating the plotting dataframe here.

```{r}
#| label: read-data
#| output: false

# Actual prescribing/issuing data
gp_all_prescribing <- read_csv(here("output", "monthly_op_gp", "adhd_gp_all_chemicals.csv")) |> 
  mutate(data_source = "GP") |> 
  rename(observed = monthly_items)

foi_all_prescribing <- read_csv(here("output", "foi_prescribing", "adhd_foi_all_chemicals.csv")) |> 
  mutate(data_source = "FOI") |> 
  rename(observed = monthly_items)

scmd_all_issuing <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_all_chemicals.csv")) |> 
  mutate(data_source = "SCMD") |> 
  rename(observed = monthly_ddd_quantity)

# Predicted values across all 3 datasets
predicted_values <- read_csv(here("output", "changepoint_analysis", "predicted_values.csv"))

# Slopes (including breakpoint locations) across datasets
slopes <- read_csv(here("output", "changepoint_analysis", "model_slopes.csv"))

# Model coefficients across datasets
coefs <- read_csv(here("output", "changepoint_analysis", "model_coefficients.csv"))
```

## Organise data for plotting

```{r}
# Join predicted and observed data dfs
df_plots <- bind_rows(foi_all_prescribing, gp_all_prescribing, scmd_all_issuing) |> 
  left_join(predicted_values, by = join_by(month, data_source)) |> 
  # Remove unnecessary columns
  dplyr::select(-c(bnf_chemical, bnf_chemical_name))

# Get breakpoint dates as a vector
breakpoints <- slopes |> 
	rowwise() |> 
	mutate(vec = list(
		na.omit(c(bp1 = bp1, bp2 = bp2, bp3 = bp3))
	)) |> 
	ungroup() |> 
	dplyr::select(dataset, vec) |> 
	deframe()
```

## Helper function for plotting

```{r}
plot_changepoints <- function(
	source,
	y_label,
	title,
	show_x = TRUE,
	breakpoint_dates,
	label_y_pos
) {
	# Define common x axis scale
	common_x_scale <- scale_x_date(
		date_breaks = "1 year",
		date_labels = "%b\n%Y",
		limits = as.Date(c("2015-08-01", "2025-07-01"))
	)

	# Define plot colours
	plot_colours = viridis(2, alpha = 0.7, begin = 0.25, end = 0.6, option = "D")

	# Subset data to plot
	dat <- df_plots |> 
		subset(data_source == source)

	# Create plot
	plot <- ggplot(dat, aes(x = month)) +
		# Plot observed values
		geom_point(
			aes(y = observed, colour = "Observed value"), 
			size = 2, 
			alpha = 0.7
		) +
		# Plot 95% CI
		geom_ribbon(
			aes(ymin = lci, ymax = uci, fill = "Fitted values with 95% CI (shaded area)"), 
			alpha = .2
		) +
		# Plot line for predicted values
		geom_line(
			aes(y = pred, colour = "Fitted values with 95% CI (shaded area)"), 
			linewidth = 1, 
			alpha = 0.7
		) +
		# Plot breakpoint locations with vertical dashed line
		geom_vline(
			xintercept = breakpoint_dates, 
			linetype = "dashed",
			colour = plot_colours[1], 
			alpha = 0.7) +
		scale_colour_manual(values = plot_colours) +
		scale_fill_manual(values = plot_colours) +
		common_x_scale + # x-axis scale
		scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
		labs(x = "Month", y = y_label, title = title) +
		theme_bw() +
		theme(
			text = element_text(family = "Times New Roman"),
			panel.grid.major.x = element_blank(),
			panel.grid.minor.x = element_blank(),
			plot.title = element_text(size = 20, hjust = .5),
			axis.ticks.x = element_line(),
			axis.title.x = element_text(size = 16),
			axis.text.x = element_text(size = 16),
			legend.text = element_text(size = 14),
			legend.title = element_blank(),
			legend.background = element_rect(
				fill = "white",
				linetype = "solid",
				colour = "black"
			),
			axis.text.y = element_text(size = 16),
			axis.title.y = element_text(size = 16)
		) +
		guides(
			fill = "none", # Remove redundant legend
			colour = guide_legend(nrow = 1)
		) + 
		annotate(
			x = breakpoint_dates, 
			y = label_y_pos, 
			label = breakpoint_dates, 
			geom = "label",
			size = 5,
			family = "Times New Roman"
		) 

	# Common x-axis
	if (show_x == FALSE) {
		plot <- plot +
			theme(
				axis.title.x = element_blank(),
				axis.text.x = element_blank()
			)
	}

	plot
}
```

# Results: plots

```{r}
fig_gp_changepoints <- plot_changepoints(
	source = "GP",
	y_label = "Monthly items",
	title = "GP Prescribing",
	show_x = FALSE,
	breakpoint_dates = breakpoints$GP,
	label_y_pos = c(225000, 225000)
)

fig_gp_changepoints

fig_foi_changepoints <- plot_changepoints(
	source = "FOI",
	y_label = "Monthly items",
	title = "FOI Prescribing",
	show_x = FALSE,
	breakpoint_dates = breakpoints$FOI,
	label_y_pos = c(75000, 50000)
)

fig_foi_changepoints

fig_scmd_changepoints <- plot_changepoints(
	source = "SCMD",
	y_label = "Monthly DDD quantity",
	title = "SCMD Issuing",
	show_x = TRUE,
	breakpoint_dates = breakpoints$SCMD,
	label_y_pos = c(210000, 175000, 100000)
)

fig_scmd_changepoints
```

The warning messages are because the SCMD data has less data available so it doesn't span the whole common x axis. This is normal and it's fine to ignore the warnings. 

## Combined figure

```{r}
fig_combined_changepoints <- fig_gp_changepoints / fig_foi_changepoints / fig_scmd_changepoints + 
	plot_layout(guides = "collect") +
	plot_annotation(tag_levels = "A") &
	theme(
		plot.title = element_text(size = 16),
		plot.tag = element_text(size = 20), 
		legend.position = "bottom",
		legend.title = element_blank(),
		legend.box = "horizontal"
	)
	

ggsave(
	here("output", "figures", "fig_combined_changepoints.png"),
	fig_combined_changepoints,
	width = 10,
	height = 10
)
```

# Results: table

```{r}

```