---
title: "Plots for change point analysis of ADHD prescribing/issuing activity"
author:
  - Kaylee Poon
format: html
---

Aiming to plot actual prescribing/issuing over time with the predicted values (+ 95% confidence intervals) from the change point analysis. Only results for analysing all chemicals combined are plotted. One plot for each of:

1. GP data
2. FOI data
3. SCMD data.

Also contains a summary table for change point model coefficients. 

# Methods

```{r}
library(tidyverse)
library(here)
library(patchwork)
library(viridis)
library(ggnewscale)
library(flextable)
library(officer)

# Set flextable defaults
set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 12,
  font.colour = "black",
  border.colour = "black",
  layout = "autofit",
  width = 1
)
```

## Read in data

Also doing some preparation for creating the plotting dataframe here.

```{r}
#| label: read-data
#| output: false

# Actual prescribing/issuing data
gp_all_prescribing <- read_csv(here("output", "monthly_op_gp", "adhd_gp_all_chemicals.csv")) |> 
  mutate(data_source = "GP") |> 
  rename(observed = monthly_items)

foi_all_prescribing <- read_csv(here("output", "foi_prescribing", "adhd_foi_all_chemicals.csv")) |> 
  mutate(data_source = "FOI") |> 
  rename(observed = monthly_items)

scmd_all_issuing <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_all_chemicals.csv")) |> 
  mutate(data_source = "SCMD") |> 
  rename(observed = monthly_ddd_quantity) |> 
  filter_out(month > as.Date("2024-12-01"))

# Predicted values across all 3 datasets
predicted_values <- read_csv(here("output", "changepoint_analysis", "predicted_values.csv")) |> 
	dplyr::select(-bnf_chemical)

# Slopes (including breakpoint locations) across datasets
slopes <- read_csv(here("output", "changepoint_analysis", "model_slopes.csv")) |> 
	dplyr::select(-bnf_chemical)

# Model coefficients across datasets
coefs <- read_csv(here("output", "changepoint_analysis", "model_coefficients.csv")) |> 
	dplyr::select(-bnf_chemical)
```

## Organise data for plotting

```{r}
# Join predicted and observed data dfs
df_plots <- bind_rows(foi_all_prescribing, gp_all_prescribing, scmd_all_issuing) |> 
  left_join(predicted_values, by = join_by(month, data_source)) |> 
  # Remove unnecessary column
  dplyr::select(-bnf_chemical_name)

# Get breakpoint dates as a vector
breakpoints <- slopes |> 
	rowwise() |> 
	mutate(vec = list(
		na.omit(c(bp1 = bp1, bp2 = bp2))
	)) |> 
	ungroup() |> 
	dplyr::select(dataset, vec) |> 
	deframe()
```

## Helper function for plotting

```{r}
plot_changepoints <- function(
	source,
	title,
	y_label,
	show_x = TRUE,
	show_legend = TRUE,
	breakpoint_dates,
	breakpoint_labels,
	breakpoint_y_pos
) {
	# Define common x axis scale
	common_x_scale <- scale_x_date(
		date_breaks = "1 year",
		date_labels = "%b\n%Y",
		limits = as.Date(c("2015-08-01", "2025-07-01"))
	)

	# Define plot colours
	plot_colours = viridis(2, alpha = 0.7, begin = 0.25, end = 0.6, option = "D")

	# Subset data to plot
	dat <- df_plots |> 
		subset(data_source == source)

	# Create plot
	plot <- ggplot(dat, aes(x = month)) +
		# Plot observed values
		geom_point(
			aes(y = observed, colour = "Observed value"), 
			size = 2
		) +
		# Plot 95% CI
		geom_ribbon(
			aes(ymin = lci, ymax = uci, fill = "Fitted values with 95% CI (shaded area)"), 
			alpha = .2
		) +
		# Plot line for predicted values
		geom_line(
			aes(y = pred, colour = "Fitted values with 95% CI (shaded area)"), 
			linewidth = 1
		) +
		# Plot breakpoint locations with vertical dashed line
		geom_vline(
			xintercept = breakpoint_dates, 
			linetype = "dashed",
			colour = plot_colours[1], 
			alpha = 0.7) +
		# Stylistic decisions below
		scale_colour_manual(values = plot_colours) +
		scale_fill_manual(values = plot_colours) +
		common_x_scale + # x-axis scale
		scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
		labs(x = "Month", title = title, y = y_label) +
		theme_bw() +
		theme(
			text = element_text(family = "Times New Roman"),
			panel.grid.major.x = element_blank(),
			panel.grid.minor.x = element_blank(),
			panel.grid.major.y = element_blank(),
			panel.grid.minor.y = element_blank(),
			plot.title = element_text(size = 20, hjust = .5),
			axis.ticks.x = element_line(),
			axis.title.x = element_text(size = 16),
			axis.text.x = element_text(size = 16),
			axis.text.y = element_text(size = 16),
			axis.title.y = element_text(size = 16),
			legend.text = element_text(size = 16),
			legend.title = element_blank(),
			legend.background = element_rect(
				fill = "white",
				linetype = "solid",
				colour = "black"
			)
		) +
		guides(
			fill = "none", # Remove redundant legend
			colour = guide_legend(nrow = 1)
		) + 
		# Add labels for breakpoint dates
		annotate(
			x = breakpoint_dates, 
			y = breakpoint_y_pos, 
			label = breakpoint_labels, 
			geom = "label",
			size = 6,
			family = "Times New Roman"
		) 

	# Control common x-axis appearance
	if (show_x == FALSE) {
		plot <- plot +
			theme(
				axis.title.x = element_blank(),
				axis.text.x = element_blank()
			)
	}

	# Control legend appearance
  if (show_legend == FALSE) {
    plot <- plot +
      theme(legend.position = "none")
  }

	plot
}
```

# Results: plots

These will _not_ look good until the `ggsave()` step because of scaling issues.

```{r}
# Figure for GP model
fig_gp_changepoints <- plot_changepoints(
	source = "GP",
	title = "GP prescribing",
	y_label = "Monthly items",
	show_x = FALSE,
	breakpoint_dates = breakpoints$GP,
	breakpoint_labels = "Aug 2020",
	# Ensure breakpoint date labels don't overlap with data
	breakpoint_y_pos = 225000 
)

fig_gp_changepoints

# Figure for FOI model
fig_foi_changepoints <- plot_changepoints(
	source = "FOI",
	title = "Private controlled drug prescribing",
	y_label = "Monthly items",
	show_x = FALSE,
	breakpoint_dates = breakpoints$FOI,
	breakpoint_labels = c("Aug 2020", "Dec 2020"),
	breakpoint_y_pos = c(75000, 50000)
)

fig_foi_changepoints

# Figure for SCMD model
# Create df for shading COVID-19 disruption periods in the plot - only the SCMD changepoint analysis needs this specified in the model
scmd_covid <- tibble(
	xmin = as.Date(c("2020-04-01", "2020-07-01")),
	xmax = as.Date(c("2020-06-01", "2021-02-01")),
	ymin = -Inf,
	ymax = Inf,
	period = c("1st COVID-19 disruption period", "2nd COVID-19 disruption period")
)

# Plot figure
fig_scmd_changepoints <- plot_changepoints(
	source = "SCMD",
	title = "Hospital issuing",
	y_label = "Monthly DDD quantity",
	show_x = TRUE,
	show_legend = FALSE,
	breakpoint_dates = breakpoints$SCMD,
	breakpoint_labels = "Oct 2021",
	breakpoint_y_pos = 210000
) +
	# Add shading for COVID-19 disruption periods
	new_scale_fill() + # `plot_changepoints` already has one 'fill' scale
	geom_rect(
	data = scmd_covid,
	aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = period),
	inherit.aes = FALSE,
	alpha = 0.3
	) +
	# Just making things look good
	scale_fill_viridis_d(name = waiver(), alpha = .5, begin = .25, end = .75, option = "turbo") 

fig_scmd_changepoints
```

The warning messages are because the SCMD data has less data available so it doesn't span the whole common x axis. This is normal and it's fine to ignore the warnings. 

## Combined figure

I need to first combine the GP and FOI figures and collect the line/point legends, then have the SCMD figure as its own object so that the legends for shading the COVID-19 disruption periods exists in the SCMD figure. Then I can combine all the figures together for the final figure. 

```{r}
# Combining GP and FOI figures with common legend at the bottom
p1 <- fig_gp_changepoints / fig_foi_changepoints +
	plot_layout(guides = "collect") &
	theme(
		plot.title = element_text(size = 20),
		legend.position = "bottom",
		legend.text = element_text(size = 16),
		legend.title = element_blank(),
		legend.box = "horizontal"
	)
	
p1

# Specifying legend position for COVID-19 period shading and keeping it in the final fig
p2 <- fig_scmd_changepoints +
	plot_layout(guides = "keep") +
	guides(colour = "none") +
	theme(
		legend.position = c(0.01, 0.95),
		legend.justification = c("left", "top")
	)
	
p2

# Combined figures with all 3 datasets and annotated panels
fig_combined_changepoints <- p1 / p2 + 
	plot_annotation(tag_levels = "A") &
	theme(plot.tag = element_text(size = 20))

ggsave(
	here("output", "figures", "fig_combined_changepoints.png"),
	fig_combined_changepoints,
	width = 10,
	height = 10
)
```

# Results: table

```{r}
# Create dataframe for slopes
df_slopes <- slopes 

# Put lower and upper 95% CI bounds in brackets after each slope estimate
slope_cols <- names(slopes) |> 
	str_subset("^slope")

for(col in slope_cols) {
	# Telling R which columns contain lower and upper 95% CI bounds
	idx <- str_extract(col, "\\d?$")
	
	lci_col <- paste0("s", idx, "_lci")
	uci_col <- paste0("s", idx, "_uci")

	# Formatting the slopes as "estimate(lci to uci)"
	df_slopes[[col]] <- paste0(
		df_slopes[[col]], " (",
		df_slopes[[lci_col]], " to ",
		df_slopes[[uci_col]], ")"
	)
}
# Repeat for 95% CIs for breakpoints
bp_cols <- names(df_slopes) |> 
	str_subset("^bp")

for(col in bp_cols) {
	idx <- str_extract(col, "\\d?$")
	
	lci_col <- paste0("bp", idx, "_lci")
	uci_col <- paste0("bp", idx, "_uci")

	df_slopes[[col]] <- paste0(
		df_slopes[[col]], " (",
		df_slopes[[lci_col]], " to ",
		df_slopes[[uci_col]], ")"
	)
}

# Remove `lci` and `uci` columns now that they aren't needed
df_slopes <- df_slopes |> 
	select(-c(contains("lci"), contains("uci")))
```

Now repeating the above process for model coefficients:

```{r}
df_coefs <- coefs |> 
	mutate(
		est = paste0(
			est, " (",
			lci, " to ",
			uci, ")"
		)
	) |> 
	dplyr::select(-c(lci, uci)) |> 
	pivot_wider(names_from = coef, values_from = c(est, holm_pvals)) |>
	# Remove monthly factor dummies - will report in text but not table 
	dplyr::select(-contains("factor")) |> 
	# Remove `est_month_number` as it is identical to slope_1 in `df_slopes`
	dplyr::select(-est_month_number)
```

Note that the columns `est_covid1` and `est_covid2` represent the level change during the 1st and 2nd COVID-19 disruption periods as defined in the model. The level changes are -37289.17 (95% CI: -41014.43 to -33563.91) and -17809.68 (95% CI: -23569.28 to -12050.07) respectively compared to the initial slope. 

Likewise, the columns `holm_pvals_covid1` and `holm_pvals_covid2` represent the p-values for the above level changes. Both p-values are <0.001 but they should be interpreted carefully as the COVID-19 disruption periods were added manually to control COVID-19 variation and allow more accurate change point estimation. These columns will be reported in the main text and not in the table below to avoid confusion. 

Combine dfs and construct table

```{r}
df_changepoint_table <- left_join(df_slopes, df_coefs) |> 
	# Remove COVID-19 level change columns
	dplyr::select(-contains("covid"))	|> 
	# Rename initial slope p-value columns - useful because initial slope for SCMD is not significantly different from 0
	rename(p_slope0 = holm_pvals_month_number) |> 
	# Rename other `est` columns (slope change compared to the previous segment)
	rename_with(
		~ paste0("est", str_extract(.x, "\\d+")),
		starts_with("est")
	) |> 
	# Rename p-value columns (for the change in slope)
	rename_with(
		~ paste0("p_slope", str_extract(.x, "\\d+")),
		starts_with("holm")
	) |> 
	# Subtract 1 from slope number as slope1 is actually the initial slope
	rename_with(
		~ paste0("slope", as.numeric(str_extract(.x, "\\d+")) - 1),
		starts_with("slope")
	) |> 
	# Turn everything into characters for `pivot_longer` below
  mutate(across(everything(), as.character)) |> 
	# Pivot longer into an easier-to-read format
	pivot_longer(
		cols = -dataset,
		# Extracting characters in original column names for the new column names
		# And extracting numbers for `idx` (breakpoint number)
		names_to = c(".value", "idx"),
		names_pattern = "(.+?)(\\d+)"
	) |> 
	# Reordering rows
	mutate(dataset = factor(dataset, levels = c("GP", "FOI", "SCMD"))) |> 
	arrange(dataset, as.numeric(idx)) |> 
	# Remove rows for breakpoints/changepoints that don't actually exist
	filter_out(slope == "NA (NA to NA)") |> 
	# More accurate representation of rounded p-values
	mutate(p_slope = ifelse(p_slope < 0.001, "<0.001", p_slope)) |> 
	relocate(bp, .after = idx)
```

## Format table

If you're reading this for code review reasons, just stop here and read the thesis later because this table is not in its final formatted form. I've just turned the dataframe above into a crude table in Word.

In the interest of time I'm not going to bother with too much formatting using flextable in R, because I want rows with change-point number/idx = 0 to be a merged sub-header row showing the data source and the initial monthly change. Then I'll delete the "Data source" column. From making a table for code usage results previously, I know MS word won't play nice with flextable merging anyway so I might as well format in Word/Gdocs. Adding sub-header rows in the middle of a flextable is also a massive pain. I later decided to remove the p-value column and put it in brackets in the level change column instead. 

```{r}
changepoint_table <- flextable(df_changepoint_table) |> 
  theme_booktabs() |> 
  set_caption("Changes in ADHD medication prescribing/issuing") |> 
  set_header_labels(
    dataset = "Data source",
		idx = "Change-point number",
    slope = "Segment slope",
    bp = "Change-point location estimate",
		est = "Change in slope",
    p_slope = "P-value of slope change"
  ) |> 
	# Formatting what will later become sub-header rows
	bold(i = c(1, 4, 7)) |> 
	hline(
    i = c(1, 3, 4, 6, 7),
    border = fp_border(color = "#666666", width = 1.5),
    part = "body"
  )
	
changepoint_table

# Save as Word document
save_as_docx(
  "Changepoint Analysis Summary Table" = changepoint_table,
  path = "output/tables/changepoint_table.docx"
)
```