---
title: "Plots for change point analysis of ADHD prescribing/issuing activity"
author:
  - Kaylee Poon
format: html
---

Aiming to plot actual prescribing/issuing over time with the predicted values (+ 95% confidence intervals) from the change point analysis. Only results for analysing all chemicals combined are plotted. One plot for each of:

1. GP data
2. FOI data
3. SCMD data.

Also contains a summary table for change point model coefficients. 

# Methods

```{r}
#| label: set-up
#| output: false

library(tidyverse)
library(here)
library(patchwork)
library(viridis)
library(flextable)
library(officer)

# Set flextable defaults
set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 12,
  font.colour = "black",
  border.colour = "black",
  layout = "autofit",
  width = 1
)
```

## Read in data

Also doing some preparation for creating the plotting dataframe here.

```{r}
#| label: read-data
#| output: false

# Actual prescribing/issuing data
gp_all_prescribing <- read_csv(here("output", "monthly_op_gp", "adhd_gp_all_chemicals.csv")) |> 
  mutate(data_source = "GP") |> 
  rename(observed = monthly_items)

foi_all_prescribing <- read_csv(here("output", "foi_prescribing", "adhd_foi_all_chemicals.csv")) |> 
  mutate(data_source = "FOI") |> 
  rename(observed = monthly_items)

scmd_all_issuing <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_all_chemicals.csv")) |> 
  mutate(data_source = "SCMD") |> 
  rename(observed = monthly_ddd_quantity)

# Predicted values across all 3 datasets
predicted_values <- read_csv(here("output", "changepoint_analysis", "predicted_values.csv"))

# Slopes (including breakpoint locations) across datasets
slopes <- read_csv(here("output", "changepoint_analysis", "model_slopes.csv"))

# Model coefficients across datasets
coefs <- read_csv(here("output", "changepoint_analysis", "model_coefficients.csv"))
```

## Organise data for plotting

```{r}
# Join predicted and observed data dfs
df_plots <- bind_rows(foi_all_prescribing, gp_all_prescribing, scmd_all_issuing) |> 
  left_join(predicted_values, by = join_by(month, data_source)) |> 
  # Remove unnecessary columns
  dplyr::select(-c(bnf_chemical, bnf_chemical_name))

# Get breakpoint dates as a vector
breakpoints <- slopes |> 
	rowwise() |> 
	mutate(vec = list(
		na.omit(c(bp1 = bp1, bp2 = bp2, bp3 = bp3))
	)) |> 
	ungroup() |> 
	dplyr::select(dataset, vec) |> 
	deframe()
```

## Helper function for plotting

```{r}
plot_changepoints <- function(
	source,
	y_label,
	title,
	show_x = TRUE,
	breakpoint_dates,
	label_y_pos
) {
	# Define common x axis scale
	common_x_scale <- scale_x_date(
		date_breaks = "1 year",
		date_labels = "%b\n%Y",
		limits = as.Date(c("2015-08-01", "2025-07-01"))
	)

	# Define plot colours
	plot_colours = viridis(2, alpha = 0.7, begin = 0.25, end = 0.6, option = "D")

	# Subset data to plot
	dat <- df_plots |> 
		subset(data_source == source)

	# Create plot
	plot <- ggplot(dat, aes(x = month)) +
		# Plot observed values
		geom_point(
			aes(y = observed, colour = "Observed value"), 
			size = 2, 
			alpha = 0.7
		) +
		# Plot 95% CI
		geom_ribbon(
			aes(ymin = lci, ymax = uci, fill = "Fitted values with 95% CI (shaded area)"), 
			alpha = .2
		) +
		# Plot line for predicted values
		geom_line(
			aes(y = pred, colour = "Fitted values with 95% CI (shaded area)"), 
			linewidth = 1, 
			alpha = 0.7
		) +
		# Plot breakpoint locations with vertical dashed line
		geom_vline(
			xintercept = breakpoint_dates, 
			linetype = "dashed",
			colour = plot_colours[1], 
			alpha = 0.7) +
		scale_colour_manual(values = plot_colours) +
		scale_fill_manual(values = plot_colours) +
		common_x_scale + # x-axis scale
		scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
		labs(x = "Month", y = y_label, title = title) +
		theme_bw() +
		theme(
			text = element_text(family = "Times New Roman"),
			panel.grid.major.x = element_blank(),
			panel.grid.minor.x = element_blank(),
			plot.title = element_text(size = 20, hjust = .5),
			axis.ticks.x = element_line(),
			axis.title.x = element_text(size = 16),
			axis.text.x = element_text(size = 16),
			legend.text = element_text(size = 14),
			legend.title = element_blank(),
			legend.background = element_rect(
				fill = "white",
				linetype = "solid",
				colour = "black"
			),
			axis.text.y = element_text(size = 16),
			axis.title.y = element_text(size = 16)
		) +
		guides(
			fill = "none", # Remove redundant legend
			colour = guide_legend(nrow = 1)
		) + 
		annotate(
			x = breakpoint_dates, 
			y = label_y_pos, 
			label = breakpoint_dates, 
			geom = "label",
			size = 5,
			family = "Times New Roman"
		) 

	# Common x-axis
	if (show_x == FALSE) {
		plot <- plot +
			theme(
				axis.title.x = element_blank(),
				axis.text.x = element_blank()
			)
	}

	plot
}
```

# Results: plots

```{r}
fig_gp_changepoints <- plot_changepoints(
	source = "GP",
	y_label = "Monthly items",
	title = "GP Prescribing",
	show_x = FALSE,
	breakpoint_dates = breakpoints$GP,
	label_y_pos = c(225000, 225000)
)

fig_gp_changepoints

fig_foi_changepoints <- plot_changepoints(
	source = "FOI",
	y_label = "Monthly items",
	title = "FOI Prescribing",
	show_x = FALSE,
	breakpoint_dates = breakpoints$FOI,
	label_y_pos = c(75000, 50000)
)

fig_foi_changepoints

fig_scmd_changepoints <- plot_changepoints(
	source = "SCMD",
	y_label = "Monthly DDD quantity",
	title = "SCMD Issuing",
	show_x = TRUE,
	breakpoint_dates = breakpoints$SCMD,
	label_y_pos = c(210000, 175000, 100000)
)

fig_scmd_changepoints
```

The warning messages are because the SCMD data has less data available so it doesn't span the whole common x axis. This is normal and it's fine to ignore the warnings. 

## Combined figure

```{r}
fig_combined_changepoints <- fig_gp_changepoints / fig_foi_changepoints / fig_scmd_changepoints + 
	plot_layout(guides = "collect") +
	plot_annotation(tag_levels = "A") &
	theme(
		plot.title = element_text(size = 16),
		plot.tag = element_text(size = 20), 
		legend.position = "bottom",
		legend.title = element_blank(),
		legend.box = "horizontal"
	)
	

ggsave(
	here("output", "figures", "fig_combined_changepoints.png"),
	fig_combined_changepoints,
	width = 10,
	height = 10
)
```

# Results: table

```{r}
df_slopes <- slopes |> 
	select(-bnf_chemical)  

slope_cols <- names(df_slopes) |> 
	str_subset("^slope")

for(col in slope_cols) {
	idx <- str_extract(col, "\\d?$")
	
	lci_col <- paste0("s", idx, "_lci")
	uci_col <- paste0("s", idx, "_uci")

	df_slopes[[col]] <- paste0(
		df_slopes[[col]], " (",
		df_slopes[[lci_col]], " to ",
		df_slopes[[uci_col]], ")"
	)
}
# Repeat for 95% CIs for breakpoints
bp_cols <- names(df_slopes) |> 
	str_subset("^bp")

for(col in bp_cols) {
	idx <- str_extract(col, "\\d?$")
	
	lci_col <- paste0("bp", idx, "_lci")
	uci_col <- paste0("bp", idx, "_uci")

	df_slopes[[col]] <- paste0(
		df_slopes[[col]], " (",
		df_slopes[[lci_col]], " to ",
		df_slopes[[uci_col]], ")"
	)
}

df_slopes <- df_slopes |> 
	select(-c(contains("lci"), contains("uci")))

```

Model coefficients

```{r}
df_coefs <- coefs |> 
	mutate(
		est = paste0(
			est, " (",
			lci, " to ",
			uci, ")"
		)
	) |> 
	dplyr::select(-c(bnf_chemical, lci, uci)) |> 
	pivot_wider(names_from = coef, values_from = c(est, holm_pvals)) |> 
	dplyr::select(-contains("factor"))

```

Combine dfs and construct table

```{r}
df_changepoint_table <- left_join(df_slopes, df_coefs) |> 
	select(-c(est_month_number, holm_pvals_month_number)) |> 
	# Reordering rows
	mutate(dataset = factor(dataset, levels = c("GP", "FOI", "SCMD"))) |> 
	# Rename `est` columns (slope change compared to the previous segment)
	rename_with(
		~ paste0("est", str_extract(.x, "\\d+")),
		starts_with("est")
	) |> 
	# Rename p-value columns (for the change in slope)
	rename_with(
		~ paste0("level_p", str_extract(.x, "\\d+")),
		starts_with("holm")
	) |> 
	# Subtract 1 from slope number as slope1 is actually the initial slope
	rename_with(
		~ paste0("slope", as.numeric(str_extract(.x, "\\d+")) - 1),
		starts_with("slope")
	) |> 
	pivot_longer(
		cols = -dataset,
		names_to = c(".value", "idx"),
		names_pattern = "([a-z]+)(\\d+)"
	) |> 
	arrange(dataset, as.numeric(idx)) |> 
	# Remove rows for breakpoints/changepoints that don't actually exist
	filter_out(slope == "NA (NA to NA)") |> 
	# More accurate representation of rounded p-values
	mutate(p = ifelse(p < 0.001, "<0.001", p)) |> 
	relocate(bp, .after = idx)
```

## Format table

If you're reading this for code review reasons, just stop here and read the thesis later because this table is not in its final formatted form. I've just turned the dataframe above into a crude table in Word.

In the interest of time I'm not going to bother with too much formatting using flextable in R, because I want rows with change-point number/idx = 0 to be a merged sub-header row showing the data source and the initial monthly change. Then I'll delete the "Data source" column. From making a table for code usage results previously, I know MS word won't play nice with flextable merging anyway so I might as well format in Word/Gdocs. Adding sub-header rows in the middle of a flextable is also a massive pain. I later decided to remove the p-value column and put it in brackets in the level change column instead. 

```{r}
changepoint_table <- flextable(df_changepoint_table) |> 
  theme_booktabs() |> 
  set_caption("Changes in ADHD medication prescribing/issuing") |> 
  set_header_labels(
    dataset = "Data source",
		idx = "Change-point number",
    slope = "Segment slope",
    bp = "Change-point location estimate",
		est = "Change in slope",
    p = "P-value of slope change"
  ) |> 
	# Formatting what will later become sub-header rows
	bold(i = c(1, 4, 7)) |> 
	hline(
    i = c(1, 3, 4, 6, 7),
    border = fp_border(color = "#666666", width = 1.5),
    part = "body"
  )
	
changepoint_table

save_as_docx(
  "Changepoint Analysis Summary Table" = changepoint_table,
  path = "output/tables/changepoint_table.docx"
)
```