---
title: "FOI ADHD prescribing analysis"
author:
  - Kaylee Poon
format: html
---
# Overview
This analysis explores FP34PCD prescribing for ADHD medications using FOI request data from 2015-08-01 to 2025-07-01. Note that FP34PCD data may include both NHS-funded Right to Choose and non-NHS private prescribing activity. 

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(nhstt)

adhd_foi_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)
```

## Data source and processing

Data extraction methods are similar to that used for OpenCodeCounts.

For this analysis we downloaded FOI response data from the NHSBSA FOI disclosure log:
- https://opendata.nhsbsa.net/dataset/foi-02082
- https://opendata.nhsbsa.net/dataset/foi-02360
- https://opendata.nhsbsa.net/dataset/foi-02722
- https://opendata.nhsbsa.net/dataset/foi-02847
- https://opendata.nhsbsa.net/dataset/foi-03492 

Data for the following 5 ADHD medications recommended by NICE were analysed.
1. Dexamfetamine / Dexamfetamine sulfate (0404000L0)
2. Methylphenidate / Methylphenidate hydrochloride (0404000M0)
3. Atomoxetine / Atomoxetine hydrochloride (0404000S0)
4. Lisdexamfetamine / Lisdexamfetamine dimesylate (0404000U0)
5. Guanfacine / Guanfacine (0404000V0) 

As the NICE-approved ADHD medications atomoxetine and guanfacine are not controlled drugs and do not need to be reported using FP34PCD forms, FP34PCD data may underestimate atomoxetine and guanfacine prescription counts.

# Results

## Total monthly prescribing for all chemicals 

@fig-total-foi-over-time shows large increases in total monthly prescribing starting in roughly 2021. 

```{r}
#| label: fig-total-foi-over-time
#| fig-cap: "Total ADHD prescribing from FOI data over time"

fig_total_foi_over_time <- adhd_foi_prescribing |> 
  select(month, items) |>
  group_by(month) |> 
  summarise(monthly_items = sum(items, na.rm = TRUE), .groups = "drop") |> 
  ggplot(aes(x = month, y = monthly_items)) +
  geom_line(alpha = .5) 

fig_total_foi_over_time

# ggsave(
#   here("output", "figures", "fig_total_gp_over_time.png"),
#   plot = fig_total_gp_over_time,
#   height = 6,
#   width = 10
# )
```

## By chemical

@fig-by-chemical-over-time shows that there was a lot more prescribing for atomoxetine and guanfacine between 2022 and 2024 compared to other years. Prescribing ativity for dexamfetamine, lisdexamfetamine, and methylphenidate has increased after 2011. A dip in prescribing for lisdexamfetamine (and for methylphenidate to a smaller extent) can be seen around the last quarter of 2023, which coincides with the beginning of ADHD medication shortages (lisdexamfetamine supplies were the most affected initially). 

```{r}
#| label: fig-by-chemical-over-time
#| fig-cap: "ADHD prescribing from FOI data by chemical over time"
fig_foi_by_chemical <- adhd_foi_prescribing |> 
    group_by(icb_code, month, bnf_chemical_name) |> 
    mutate(items_by_chemical = sum(items)) |> 
    ungroup() |> 
    ggplot(aes(x = month, y = items_by_chemical, group = bnf_chemical_name)) +
    geom_line(alpha = .5) +
    facet_wrap(~bnf_chemical_name, scales = "free_y")

fig_foi_by_chemical
```