---
title: "Get ONS yearly population estimates for each ICB"
format: html
---
Code for getting yearly population estimates for each ICB. Heavily inspired by `analysis/build_ons_nhs_england_region_population.R` in the `bennettoxford/lithium_project25` GitHub repository. I extract ICB names instead of NHS region names and codes.

Set up:

```{r}
library(here)
library(tidyverse)
library(readxl)
```

ICB-level analysis should be adjusted for population. Mid-year (30 June) population estimates for ICBs can be obtained from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates). Select the "Mid-2022 revised (Nov 2025) to mid-2024: Integrated Care Boards, 2024 geography edition of this dataset"  and "Mid-2011 to mid-2022: Integrated Care Boards, 2024 geography edition of this dataset" editions and download the .xlsx files. The ONS named the files "sapeicb20222024.xlsx" and "sapeicb202420112022.xlsx" respectively. I have placed them in output/icb_mapping. 

```{r}
# Rename Excel files
file.rename(
  from = c(
    here("output", "icb_mapping", "sapeicb20222024.xlsx"),
    here("output", "icb_mapping", "sapeicb202420112022.xlsx")
	),
  to = c(
		here("output", "icb_mapping", "ons_pop_estimates_2022-2024.xlsx"),
		here("output", "icb_mapping", "ons_pop_estimates_2011-2022.xlsx")
	)
)

# Set paths for reading Excel files
paths <- tibble(
  workbook_label = c(
    "ons_pop_estimates_2011-2022.xlsx",
    "ons_pop_estimates_2022-2024.xlsx"
  ),
  path = here("output", "icb_mapping", workbook_label)
)

stopifnot(all(file.exists(paths$path)))

# Select sheets to read using name of sheet
sheet_config <- bind_rows(
  tibble(
    workbook_label = "ons_pop_estimates_2011-2022.xlsx",
    estimate_year = 2011:2021
  ),
  tibble(
    workbook_label = "ons_pop_estimates_2022-2024.xlsx",
    estimate_year = 2022:2024
  )
) |> 
  mutate(sheet = paste0("Mid-", estimate_year, " ICB 2024"))

# Select ICB name and total populations columns to read in
column_config <- tibble(
  workbook_label = c(
    "ons_pop_estimates_2011-2022.xlsx",
    "ons_pop_estimates_2022-2024.xlsx"
  ),
  name_col = c("ICB 2024 Name", "ICB 2024 Name"),
  total_col = c("Total", "Total")
)
```

Helper functions for reading Excel files:

```{r}
read_icb_year_sheet <- function(path, sheet, workbook_label, estimate_year, name_col, total_col) {
  df <- read_xlsx(path, sheet = sheet, skip = 3)

  required_cols <- c(name_col, total_col)
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0L) {
    stop(
      "Missing expected columns in sheet ", sheet, " of ", workbook_label, ": ",
      paste(missing_cols, collapse = ", ")
    )
  }

  df |> 
    transmute(
      icb_name = str_squish(as.character(.data[[name_col]])),
      population = as.numeric(.data[[total_col]])
    ) |> 
    filter(!is.na(icb_name), icb_name != "", !is.na(population)) |> 
    mutate(
      estimate_year = estimate_year,
      source_workbook = workbook_label,
      .before = 1
    )
}

summarise_regions <- function(path, workbook_label) {
  sheet_rows <- sheet_config |> filter(workbook_label == .env$workbook_label)
  col_row <- column_config |> filter(workbook_label == .env$workbook_label)

  map_dfr(seq_len(nrow(sheet_rows)), function(i) {
    read_icb_year_sheet(
      path = path,
      sheet = sheet_rows$sheet[[i]],
      workbook_label = workbook_label,
      estimate_year = sheet_rows$estimate_year[[i]],
      name_col = col_row$name_col[[1]],
      total_col = col_row$total_col[[1]]
    ) |> 
      group_by(estimate_year, source_workbook, icb_name) |> 
      summarise(population = sum(population, na.rm = TRUE), .groups = "drop")
  })
}
```

Check data and write csv

```{r}
# `pmap` maps parallel inputs, `_dfr` combines all returned outputs into one df by rows
pop_raw <- pmap_dfr(list(paths$path, paths$workbook_label), summarise_regions)

pop_long <- pop_raw

if (nrow(pop_long) == 0L) {
  stop("No NHS England ICB rows were extracted")
}

expected_regions <- sort(unique(pop_long$icb_name))
if (length(expected_regions) != 42L) {
  warning(
    "Expected 42 NHS England regions; found ", length(expected_regions), ": ",
    paste(expected_regions, collapse = ", ")
  )
}

# Write csv
write_csv(pop_long, here("output", "icb_mapping", "ons_icb_population_estimates.csv"))
```