---
title: "GP Prescribing (OpenPrescribing) analysis"
author:
  - (Redacted for blind review)
format: html
---
# Overview
This analysis explores GP prescribing records for ADHD medications using OpenPrescribing/English Prescribing Dataset data from 2015-08-01 to 2025-07-01. 

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(nhstt)

adhd_gp_prescribing <- read_csv(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.csv"
    )
)
```

## Data source and processing

Data extraction methods are based on the [OpenPrescribing](https://github.com/bennettoxford/openprescribing) web app.

For this analysis we extracted GP prescribing data from the [English Prescribing Dataset](https://opendata.nhsbsa.net/dataset/english-prescribing-dataset-epd-with-snomed-code/resource/9ab2d5e6-76eb-4bbc-acc1-789952ae3454) through BigQuery.

Data for the following 5 ADHD medications recommended by NICE were extracted using the chemical's pseudoBNF code:

1. Dexamfetamine / Dexamfetamine sulfate (0404000L0)
2. Methylphenidate / Methylphenidate hydrochloride (0404000M0)
3. Atomoxetine / Atomoxetine hydrochloride (0404000S0)
4. Lisdexamfetamine / Lisdexamfetamine dimesylate (0404000U0)
5. Guanfacine / Guanfacine (0404000V0)

We then summed the number of items for each chemical at each practice.

# Results

## Total prescribing over time

```{r}
#| label: fig-total-over-time
#| fig-cap: "Total ADHD prescribing by GP practices over time"

fig_total_gp_over_time <- adhd_gp_prescribing |> 
  select(month, items_by_chemical) |>
  group_by(month) |> 
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop") |> 
  ggplot(aes(x = month, y = monthly_items)) +
  geom_line(alpha = .5) 

fig_total_gp_over_time

ggsave(
  here("output", "figures", "fig_total_gp_over_time.png"),
  plot = fig_total_gp_over_time,
  height = 6,
  width = 10
)
```
## Monthly items per chemical for each practice

@fig-op-by-practice shows monthly items per chemical for each practice. We can visually identify that some individual practices are 'outliers' and prescribe many more items than others. We will probably not put this graph in the thesis as it looks messy but it's useful to keep in mind that some individual practices have much higher prescribing activity. 

```{r}
#| label: fig-op-by-practice
#| fig-cap: "Regional total items by chemical" 

fig_op_by_practice <- adhd_gp_prescribing |>
  select(month, practice, bnf_chemical_name, items_by_chemical) |>
  ggplot(aes(x = month, y = items_by_chemical, group = practice)) +
  geom_line(alpha = .1) +
  facet_wrap(~bnf_chemical_name, scales = "free_y")

fig_op_by_practice

ggsave(
  here("output", "figures", "fig_op_by_practice.png"),
  plot = fig_op_by_practice,
  height = 6,
  width = 10
)
```


## Prescribing by chemical and regional team

Regional team codes are taken from the [Office for National Statistics website](https://geoportal.statistics.gov.uk/documents/46b634b42ceb45cbbfbe9c960fb77ec9/about).

ccg_id and pct columns are identical.

@fig-op-by-region shows total items for each chemical by regional team.

```{r}
#| label: fig-op-by-region
#| fig-cap: "Items by region" 
region_names <- c(
  "Y56" = "London",
  "Y58" = "South West","Y59" = "South East",
  "Y60" = "Midlands",
  "Y61" = "East of England",
  "Y62" = "North West",
  "Y63" = "North East and Yorkshire"
)

fig_op_by_region <- adhd_gp_prescribing |>
  select(month, regional_team, bnf_chemical_name, items_by_chemical) |>
  filter(!regional_team %in% c("Y54", "Y55", "Y57")) |> 
  group_by(regional_team, month, bnf_chemical_name) |> 
  summarise(items_by_region = sum(items_by_chemical, na.rm = TRUE), .groups = "drop") |> 
  ggplot(aes(
    x = month, 
    y = items_by_region, 
    colour = bnf_chemical_name, 
    group = bnf_chemical_name
  )) +
  geom_point(alpha = .1, size = .5) +
  geom_line(alpha = .5) +
  facet_wrap(~regional_team, labeller = as_labeller(region_names, label_wrap_gen(width = 20)))
fig_op_by_region

ggsave(
  here("output", "figures", "fig_op_by_region.png"),
  plot = fig_op_by_region,
  height = 6,
  width = 10
)
```

## Functions needed for decile charts
```{r}
#| label: write-functions

get_percentiles <- function(data, period_col, value_col, group_by) {
  data |>
    nhstt:::summarise_percentiles(
      period_col = {{ period_col }},
      value_col = {{ value_col }},
      group_by = {{ group_by }}
    ) |>
    pivot_wider(
      names_from = percentile,
      values_from = value,
      names_prefix = "p"
    )
}


plot_percentiles <- function(
  data,
  x_col,
  title = NULL,
  y_label = "Value",
  ribbon_colour = "#6baed6",
  median_colour = "#ff000d"
) {
  decile_colours <- c(
    "10th-90th" = alpha(ribbon_colour, 0.2),
    "20th-80th" = alpha(ribbon_colour, 0.3),
    "30th-70th" = alpha(ribbon_colour, 0.4),
    "40th-60th" = alpha(ribbon_colour, 0.5),
    "Median" = median_colour
  )

  p <- ggplot(data, aes(x = {{ x_col }})) +
    # Ribbons first (background)
    geom_ribbon(aes(ymin = p10, ymax = p90, fill = "10th-90th")) +
    geom_ribbon(aes(ymin = p20, ymax = p80, fill = "20th-80th")) +
    geom_ribbon(aes(ymin = p30, ymax = p70, fill = "30th-70th")) +
    geom_ribbon(aes(ymin = p40, ymax = p60, fill = "40th-60th")) +
    # Median line and points on top
    geom_line(aes(y = p50, colour = "Median"), linewidth = 0.5) +
    geom_point(aes(y = p50, colour = "Median"), size = 1) +
    # Scales
    scale_fill_manual(
      values = decile_colours,
      breaks = c("10th-90th", "20th-80th", "30th-70th", "40th-60th"),
      name = NULL
    ) +
    scale_colour_manual(
      values = decile_colours,
      breaks = "Median",
      name = NULL
    ) +
    scale_y_continuous(
      labels = scales::number_format(),
      name = y_label
    ) +
    scale_x_date(
      breaks = unique(pull(data, {{ x_col }}))[seq(
        1,
        length(unique(pull(data, {{ x_col }}))),
        by = 6
      )],
      date_labels = "%b\n%Y",
      name = NULL
    ) +
    guides(
      fill = guide_legend(order = 1),
      colour = guide_legend(order = 2)
    ) +
    theme_minimal() +
    theme(legend.position = "right")

  if (!is.null(title)) {
    p <- p + labs(title = title)
  }

  p
}

```

## Decile chart of total items per chemical (practice level)

@fig-deciles-by-practice is not very useful as median items are quite low for atomoxetine, dexamfetamine, and guanfacine. There is also little practice variation so it is more useful to look at regional team or ICB-level data instead. 

```{r}
#| label: fig-deciles-by-practice
#| fig-cap: "Decile chart of monthly total items per chemical." 

fig_op_deciles_by_med_practice <- adhd_gp_prescribing |>
  select(month, practice, bnf_chemical_name, items_by_chemical) |>
  get_percentiles(
    period_col = month,
    value_col = items_by_chemical,
    group_by = bnf_chemical_name
  ) |>
  plot_percentiles(
    x_col = month,
    y_label = "Total items"
  ) +
  facet_wrap(~bnf_chemical_name, ncol = 1, scales = "free_y")

fig_op_deciles_by_med_practice

ggsave(
  here("output", "figures", "fig_op_deciles_by_med_practice.png"),
  plot = fig_op_deciles_by_med_practice,
  height = 12,
  width = 10
)
```

## Decile chart of total items per practice 

@fig-practice-deciles shows a small increase (from 3 to 5) prescribing items per practice over time. This change is largely driven by the top 20% of prescribers. Not a very good graph for the thesis but shows that median monthly total items is quite low and top prescribers show larger increases.

```{r}
#| label: fig-practice-deciles
#| fig-cap: "Decile chart of monthly total items." 

fig_op_deciles_by_practice <- adhd_gp_prescribing |>
  select(month, practice, bnf_chemical_name, items_by_chemical) |>
  get_percentiles(
    period_col = month,
    value_col = items_by_chemical
  ) |>
  plot_percentiles(
    x_col = month,
    y_label = "Total items"
  ) 

fig_op_deciles_by_practice

ggsave(
  here("output", "figures", "fig_op_deciles_by_practice.png"),
  plot = fig_op_deciles_by_practice,
  height = 4,
  width = 10
)
```

