---
title: "GP Prescribing (OpenPrescribing) analysis"
author:
  - Kaylee Poon
format: html
---
# Overview
This analysis explores GP prescribing records for ADHD medications using OpenPrescribing/English Prescribing Dataset data from 2015-08-01 to 2025-07-01. 

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)

adhd_gp_prescribing <- read_csv(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing.csv"
    )
)
```

## Data source and processing

Data extraction methods are based on the [OpenPrescribing](https://github.com/bennettoxford/openprescribing) web app.

For this analysis we extracted GP prescribing data from the [English Prescribing Dataset](https://opendata.nhsbsa.net/dataset/english-prescribing-dataset-epd-with-snomed-code/resource/9ab2d5e6-76eb-4bbc-acc1-789952ae3454) through BigQuery.

Data for the following 5 ADHD medications recommended by NICE were extracted using the chemical's pseudoBNF code:

1. Dexamfetamine / Dexamfetamine sulfate (0404000L0)
2. Methylphenidate / Methylphenidate hydrochloride (0404000M0)
3. Atomoxetine / Atomoxetine hydrochloride (0404000S0)
4. Lisdexamfetamine / Lisdexamfetamine dimesylate (0404000U0)
5. Guanfacine / Guanfacine (0404000V0)

We then summed the number of items for each chemical at each practice.

# Initial analysis

```{r}
# get unique number of ccg_id
unique(adhd_gp_prescribing$ccg_id)
# 213
unique(adhd_gp_prescribing$pct)
# 213

# are ccg_id and pct columns the same?
identical(adhd_gp_prescribing$ccg_id,adhd_gp_prescribing$pct)
# [1] TRUE

```