# Exploratory analysis for ADHD diagnosis ICD-10 code usage in hospital admitted patient care

This data comes from ICD-10 xlsx files. The following breakdowns are available:

* Age
* Sex
* Diagnosis position 

## Setup and load data

```{r}
#| output: false
library(tidyverse)
library(here)
icd10_f90_breakdowns <- read_csv(here("output", "opencodecounts", "icd10_f90_breakdowns.csv"))
```


# Age breakdowns (all diagnosis positions)

```{r}
icd10_f90_breakdowns |> 
#O nly looking at F900 now since it makes up most of the coding activity
  filter(icd10_code == "F900") |> 
  # filter(breakdown %in% c("male", "female")) |> 
  filter(str_starts(breakdown, "age")) |>
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  facet_wrap(~breakdown)
```

Note that the "age-5_9" group is positioned between the "age_45_49" and "age_50_54" groups but this can be fixed later.
Also interesting to see what happens if we aggregate ages 15-19.

# Sex breakdowns (all diagnosis positions)

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  filter(breakdown %in% c("male", "female", "gender_unknown")) |> 
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3)
```

# Diagnosis position breakdowns

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  # filter(breakdown %in% c("male", "female")) |> 
  filter(breakdown %in% c("main_diagnosis", "all_diagnoses")) |> 
  # filter(str_starts(breakdown, "age")) |>
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  facet_wrap(~breakdown, scales = "free_y")
```

As expected, the majority of code usage is recorded in other diagnosis positions. This is because ADHD being the primary reason for inpatient care is rare.