---
title: "Exploratory analysis for ICD-10 age, sex, and diagnosis position breakdowns"
format:
    html:
        toc: true
        toc-location: body
        code-fold: true
execute:
  message: false
---


This data comes from ICD-10 xlsx files. The following breakdowns are available:

* Age
* Sex
* Diagnosis position 

## Setup and load data

```{r}
#| output: false
library(tidyverse)
library(here)
icd10_f90_breakdowns <- read_csv(here("output", "opencodecounts", "icd10_f90_breakdowns.csv"))
```


# Age breakdowns (all age groups and diagnosis positions)

```{r}
icd10_f90_breakdowns |> 
# Only looking at F900 now since it makes up most of the coding activity
  filter(icd10_code == "F900") |> 
  filter(str_starts(breakdown, "age")) |>
  # Turn age bands into factor to ensure correct display order
  mutate(breakdown = factor(
      breakdown,
      levels = unique(breakdown)[order(
        parse_number(unique(breakdown))
      )]
    )) |> 
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  theme(legend.position = "none") +
  facet_wrap(~breakdown, ncol = 4)
```

## Age breakdowns (ages 15-19 aggregated)

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  filter(str_starts(breakdown, "age")) |>
  mutate(breakdown = factor(
      breakdown,
      levels = unique(breakdown)[order(
        parse_number(unique(breakdown))
      )]
    )) |> 
  # Aggregate factor levels age_15 to age_19
  mutate(breakdown = fct_collapse(
    breakdown,
    "age_15_19" = c("age_15", "age_16", "age_17", "age_18", "age_19")
  )) |> 
  # Need to also group by end_date to plot over time
  group_by(breakdown, end_date) |> 
  summarise(usage = sum(usage), .groups = "drop") |> 
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  theme(legend.position = "none") +
  facet_wrap(~breakdown, ncol = 4)
```

Once ages 15-19 have been aggregated, we can see similar increases in ICD-10 coding over time compared to ages 10-14 and 20-24. 

## Age breakdowns (<18 and >=18)
```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  filter(str_starts(breakdown, "age")) |>
  mutate(breakdown = factor(
      breakdown,
      levels = unique(breakdown)[order(
        parse_number(unique(breakdown))
      )]
    )) |> 
  # Aggregate factor levels <18 and >=18 
  mutate(breakdown = fct_collapse(
    breakdown,
    "age <18" = c(
      "age_0", "age_1_4", "age_5_9", "age_10_14", 
      "age_15", "age_16", "age_17"),
    "age >=18" = c(
      "age_18", "age_19", "age_20_24", "age_25_29",
      "age_30_34", "age_35_39", "age_40_44", "age_45_49",
      "age_50_54", "age_55_59", "age_60_64", "age_65_69",
      "age_70_74", "age_75_79", "age_80_84", "age_85_89",
      "age_90plus"
    )
  )) |> 
  group_by(breakdown, end_date) |> 
  summarise(usage = sum(usage), .groups = "drop") |> 
  ggplot(aes(x = end_date, y = usage, shape = breakdown, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3)
```

# Sex breakdowns (all diagnosis positions)

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  filter(breakdown %in% c("male", "female", "gender_unknown")) |> 
  ggplot(aes(x = end_date, y = usage, shape = breakdown, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3)
```

# Diagnosis position breakdowns

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  # filter(breakdown %in% c("male", "female")) |> 
  filter(breakdown %in% c("main_diagnosis", "all_diagnoses")) |> 
  # filter(str_starts(breakdown, "age")) |>
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  facet_wrap(~breakdown, scales = "free_y") + 
  theme(legend.position = "bottom")
```

As expected, the majority of code usage is recorded in other diagnosis positions (note different y axis scales). This is because ADHD being the primary reason for inpatient care is rare.