---
title: "Exploratory analysis for ICD-10 age, sex, and diagnosis position breakdowns"
format:
    html:
        toc: true
        toc-location: body
execute:
  message: false
---


This data comes from ICD-10 xlsx files. The following breakdowns are available:

* Age
* Sex
* Diagnosis position 

## Setup and load data

```{r}
#| output: false
library(tidyverse)
library(here)
icd10_f90_breakdowns <- read_csv(here("output", "opencodecounts", "icd10_f90_breakdowns.csv"))
```


# Age breakdowns (all age groups and diagnosis positions)

```{r}
icd10_f90_breakdowns |> 
# Only looking at F900 now since it makes up most of the coding activity
  filter(icd10_code == "F900") |> 
  filter(str_starts(breakdown, "age")) |>
  # Turn age bands into factor to ensure correct display order
  mutate(breakdown = factor(
      breakdown,
      levels = unique(breakdown)[order(
        parse_number(unique(breakdown))
      )]
    )) |> 
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  theme(legend.position = "none") +
  facet_wrap(~breakdown, ncol = 4)
```

Interesting to see what happens if we aggregate ages 15-19. May also want to do separate graphs for <18 and >=18 age groups.

# Sex breakdowns (all diagnosis positions)

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  filter(breakdown %in% c("male", "female", "gender_unknown")) |> 
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3)
```

# Diagnosis position breakdowns

```{r}
icd10_f90_breakdowns |> 
  filter(icd10_code == "F900") |> 
  # filter(breakdown %in% c("male", "female")) |> 
  filter(breakdown %in% c("main_diagnosis", "all_diagnoses")) |> 
  # filter(str_starts(breakdown, "age")) |>
  ggplot(aes(x = end_date, y = usage, colour = breakdown)) +
  geom_point() +
  geom_line(alpha = .3) +
  facet_wrap(~breakdown, scales = "free_y")
```

As expected, the majority of code usage is recorded in other diagnosis positions. This is because ADHD being the primary reason for inpatient care is rare.