---
title: "Change point analysis of ADHD prescribing activity"
author:
  - Kaylee Poon
format: html
---
# Overview
This analysis uses the `segmented` R package to conduct change point analysis on the rate of change of ADHD prescribing or medicine issuing activity. The aim is to identify when significant changes in ADHD medication prescribing or issuing occur. 

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(ggplot2)
library(segmented)
library(performance)
library(arrow)
library(see)

source(here("functions", "changepoint_functions.R"))

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
    )
)

gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
    )
)
```

## Data sources 

We will start with FP34PCD prescribing FOI data from 2015-08-01 to 2025-07-01 as it is the most simple. Note that only prescription item counts are available and DDD quantity cannot be calculated without quantity information. Here we calculate monthly prescribing for indiviudal chemicals and all chemicals. 

```{r}
#| label: calculate-monthly-prescribing
#| output: false

# Individual chemicals
foi_by_chemical <- foi_adhd_prescribing |> 
  dplyr::select(month, items, bnf_chemical_name) |>
  group_by(month, bnf_chemical_name) |> 
  summarise(monthly_items = sum(items, na.rm = TRUE), .groups = "drop")

foi_all_chemicals <- foi_adhd_prescribing |>
  dplyr::select(month, items) |>
  group_by(month) |>
  summarise(monthly_items = sum(items, na.rm = TRUE), .groups = "drop") |>
  mutate(bnf_chemical_name = "All chemicals")

total_foi_over_time <- bind_rows(foi_by_chemical, foi_all_chemicals) |> 
  arrange(month, bnf_chemical_name) |>
  group_by(bnf_chemical_name) |>
  # Add column for month number starting from 0, this makes interpreting the model easier
  mutate(month_number = row_number() - 1) |> 
  ungroup()
```

Note that data for atomoxetine and guanfacine (not controlled drugs) only starts being reported on 2015-12-01 and 2016-06-01 respectively. 

```{r}
total_foi_over_time[total_foi_over_time$bnf_chemical_name == "Atomoxetine", ][1, ]
# Date: 2015-12-01

total_foi_over_time[total_foi_over_time$bnf_chemical_name == "Guanfacine", ][1, ]
# Date: 2016-06-01
```

Later we will add GP prescribing data and Secondary Care Medicines Data.

# Results

## FP34PCD 

### All chemicals 
First, we can use Andrea's `mod_aic` function to calculate AIC for a model with 0, 1, and 2 breakpoints. Note that `glm()` is used to create an initial Poisson model.

Both `mod_aic` and `selgmented` work. I've left in the `mod_aic` function for now so we can check that both methods do produce essentially the same results, but later I use models built with `selgmented` because it lets me use fewer functions and avoids mistakes with manually inputing the number of changepoints.
```{r}
#| label: create-segmented-models
#| output: false

mod_aic(total_foi_over_time, "All chemicals")
# 2 segments has the lowest AIC 

# Create initial model
foi_all_model0 <- glm(monthly_items ~ month_number, family = poisson, data = subset(total_foi_over_time, bnf_chemical_name == "All chemicals"))

# Fit segmented model with 2 changepoints
foi_all_segmented <- segmented(foi_all_model0, npsi = 2)
summary(foi_all_segmented)

# To achieve the same effect without needing the function, `selgmented` can also determine whether 0, 1, or 2 breakpoints produces a model with the lowest AIC and fit the optimal model minimising AIC using the code below. This directly outputs a segmented object. The calculated AIC values can be slightly different e.g. 56905.97 vs 56906.29 but it should not be significant.


# Use Kmax to adjust maximum number of breakpoints
foi_all_selgmented <- mod_select(total_foi_over_time, "All chemicals")

summary(foi_all_selgmented)
```

Now we can use a modified version of Andrea's `predicted` function to calculate predicted values and `slope` functions to calculate slope. Note that because I used `selgmented` above to generate the model directly, I only need one predicted function regardless of how many changepoints there are. For `slope`, I still need to use `slope0`, `slope1`, or slope`2` depending on the number of changepoints.
```{r}
# Generate df using segmented model with predicted monthly items and 95% CI bounds over time
pred_foi_all <- predicted(foi_all_selgmented, "All chemicals", total_foi_over_time, "FOI")

# Generate df with slopes and 95% CI bounds for each segment
slope_foi_all <- slope2(foi_all_selgmented, "All chemicals", "FOI")
```

Checking model fit and assumptions
```{r}
acf(residuals(foi_all_selgmented))
# Looks fine
plot(foi_all_chemicals$month, residuals(foi_all_selgmented, type = "deviance"))
plot(fitted(foi_all_selgmented), residuals(foi_all_selgmented))
# Both of these don't look very good
check_overdispersion(foi_all_selgmented)
# Dispersion ratio = 480, overdispersion detected

# Fit negative binomial distribution instead
foi_nb_model0 <- glm.nb(monthly_items ~ month_number,  data = subset(total_foi_over_time, bnf_chemical_name == "All chemicals"))

foi_nb_segmented <- segmented.default(foi_nb_model0, npsi = 1)

summary(foi_nb_segmented)
# No breakpoint could be detected 

pred_foi_nb_model0 <- predicted(foi_nb_model0, "All chemicals", total_foi_over_time, "FOI")

check_overdispersion(foi_nb_model0)
# Dispersion ratio = 0.386, underdispersion detected

```

Fitting a log-transformed linear segmented model for FOI data

```{r}
foi_log_model <- log_mod_select(total_foi_over_time, "All chemicals")

summary(foi_log_model)
# Estimated break points: 59.999, 64.117

# Check model fit
check_model(foi_log_model)
# Doesn't look too bad according to Andrea

acf(residuals(foi_log_model))
# No autocorrelation present

pred_foi_log_model <- predicted_exp(
  foi_log_model, "All chemicals", total_foi_over_time, "FOI"
)

plot(foi_log_model)
```

## Plotting

Plotting all 3 models with observed data.

```{r}
# Rounding to nearest month
foi_break_month_number <- round(c(slope_foi_all$change1, slope_foi_all$change2, 59.999, 64.117))

# Convert month number into date
foi_break_dates <- foi_all_chemicals$month[foi_break_month_number + 1]
# "2020-04-01", "2023-01-01", "2020-08-01", "2020-12-01"

ggplot() +
  geom_point(
    data = foi_all_chemicals,
    aes(x = month, y = monthly_items, colour = "Observed"),
    alpha = .7
  ) +
  geom_line(
    data = pred_foi_all,
    aes(x= month, y = pred, colour = "Segmented Poisson model"),
    size = 1
  ) +
  geom_vline(xintercept = foi_break_dates, colour = c("red", "red", "green", "green"), alpha = 0.5) +
  geom_line(
    data = pred_foi_nb_model0,
    aes(x = month, y = pred, colour = "Negative binomial model"), 
    size = 1
  ) +
  geom_line(
    data = pred_foi_log_model,
    aes(x = month, y = pred, colour = "Segmented log-linear model"),
    size = 1
  ) +
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented Poisson model" = "red",
      "Negative binomial model" = "blue",
      "Segmented log-linear model" = "green"
    )
  ) +
  annotate(x = ymd("2020-04-01"), y = 35000, label = "2020-04-01", geom = "label", colour = "red") +
  annotate(x = ymd("2023-01-01"), y = 65000, label = "2023-01-01", geom = "label", colour = "red") +
  annotate(x = ymd("2020-08-01"), y = 45000, label = "2020-08-01", geom = "label", colour = "green") +
  annotate(x = ymd("2020-12-01"), y = 55000, label = "2020-12-01", geom = "label", colour = "green") +
  
  theme(legend.position = "right")
```

## GP Prescribing Data (items)

Fitting Poisson, negative binomial, and log-transformed models for GP prescribing data from 2015-08-01 to 2025-07-01 then calculating change points. Using items as calculating DDDs is non-trivial and not the best use of limited time. Note that prescribing for Guanfacine was only recorded starting from 2016-02-01.

```{r}
#| label: calculate-monthly-gp-prescribing
#| output: false

# Same code as FOI monthly prescribing but need to fix slightly different column names
# Individual chemicals
gp_by_chemical <- gp_adhd_prescribing |> 
  dplyr::select(month, items_by_chemical, bnf_chemical_name) |>
  group_by(month, bnf_chemical_name) |> 
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop")

gp_all_chemicals <- gp_adhd_prescribing |>
  dplyr::select(month, items_by_chemical) |>
  group_by(month) |>
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop") |>
  mutate(bnf_chemical_name = "All chemicals")

total_gp_over_time <- bind_rows(gp_by_chemical, gp_all_chemicals) |> 
  arrange(month, bnf_chemical_name) |>
  group_by(bnf_chemical_name) |>
  # Add column for month number starting from 0, this makes interpreting the model easier
  mutate(month_number = row_number() - 1) |> 
  ungroup()
```

Creating Poisson linear segmented model for GP data
```{r}
#| label: create-gp-segmented-models
#| output: false

mod_aic(total_gp_over_time, "All chemicals")
# 2 segments has the lowest AIC 

# Create initial model
gp_all_model0 <- glm(monthly_items ~ month_number, family = poisson, data = subset(total_gp_over_time, bnf_chemical_name == "All chemicals"))

# Fit segmented model with 2 changepoints
gp_all_segmented <- segmented(gp_all_model0, npsi = 2)
summary(gp_all_segmented)
# Estimated break points: 35.948 and 66.001

# Using `selgmented`
# Use Kmax to adjust maximum number of breakpoints
gp_all_selgmented <- mod_select(total_gp_over_time, "All chemicals")

summary(gp_all_selgmented)
# Estimated break points: 36.055 and 66.012

# Generate df using segmented model with predicted monthly items and 95% CI bounds over time
pred_gp_all <- predicted(gp_all_selgmented, "All chemicals", total_gp_over_time, "GP")

# Generate df with slopes and 95% CI bounds for each segment
slope_gp_all <- slope2(gp_all_selgmented, "All chemicals", "GP")
```

Checking model fit and assumptions for the Poisson segmented model
```{r}
acf(residuals(gp_all_selgmented))
# A bit of autocorrelation
plot(gp_all_chemicals$month, residuals(gp_all_selgmented, type = "deviance"))
# Not too bad
plot(fitted(gp_all_selgmented), residuals(gp_all_selgmented))
# Clustering towards the left
check_overdispersion(gp_all_selgmented)
# Dispersion ratio = 256, overdispersion detected

check_model(gp_all_selgmented)
```

Creating negative binomial model for GP data

```{r}
gp_nb_model0 <- glm.nb(monthly_items ~ month_number,  data = subset(total_gp_over_time, bnf_chemical_name == "All chemicals"))

gp_nb_segmented <- segmented.default(gp_nb_model0, npsi = 1)

summary(gp_nb_segmented)
# No breakpoint could be detected 

pred_gp_nb_model0 <- predicted(gp_nb_model0, "All chemicals", total_gp_over_time, "FOI")

check_overdispersion(gp_nb_model0)
# Dispersion ratio = 0.997, no underdispersion detected

check_model(gp_nb_model0)
acf(residuals(gp_nb_model0))
# Significant autocorrelation
```

Creating "continuous with a log transformation" linear model for GP data

```{r}
gp_log_model <- log_mod_select(total_gp_over_time, "All chemicals")

summary(gp_log_model)
# Estimated break point: 60

# Check model fit
check_model(gp_log_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions and no significant outliers

acf(residuals(gp_log_model))
# Autocorrelation present

pred_gp_log_model <- predicted_exp(
  gp_log_model, "All chemicals", total_gp_over_time, "GP"
)
```

This log-transformed linear segmented model seems to be a good model choice but we will need to adjust for autocorrelation.

Plotting all models with observed data
```{r}
# Rounding breakpoints to nearest month
gp_break_month_number <- round(c(slope_gp_all$change1, slope_gp_all$change2, 60))

# Convert month number into date
gp_break_dates <- gp_all_chemicals$month[gp_break_month_number + 1]
# "2018-08-01" "2021-02-01" "2020-08-01"

ggplot() +
  geom_point(
    data = gp_all_chemicals,
    aes(x = month, y = monthly_items, colour = "Observed"),
    alpha = .7
  ) +
  geom_line(
    data = pred_gp_all,
    aes(x= month, y = pred, colour = "Segmented Poisson model")
  ) +
  geom_vline(xintercept = gp_break_dates, colour = c("red", "red", "green"), alpha = 0.5) +
  geom_line(
    data = pred_gp_nb_model0,
    aes(x = month, y = pred, colour = "Negative binomial model")
  ) +
  geom_line(
    data = pred_gp_log_model,
    aes(x = month, y = pred, colour = "Segmented log-linear model")
  ) +
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented Poisson model" = "red",
      "Negative binomial model" = "blue",
      "Segmented log-linear model" = "green"
    )
  ) +
  annotate(x = ymd("2018-08-01"), y = 225000, label = "2018-08-01", geom = "label", colour = "red") +
  annotate(x = ymd("2021-02-01"), y = 225000, label = "2021-02-01", geom = "label", colour = "red") +
  annotate(x = ymd("2020-08-01"), y = 175000, label = "2020-08-01", geom = "label", colour = "green") +
  theme(legend.position = "right")
```

Looks like a decent fit.