---
title: "Change point analysis of ADHD prescribing activity"
author:
  - Kaylee Poon
format: html
---
# Overview
This analysis uses the `segmented` R package to conduct change point analysis on the rate of change of ADHD prescribing or medicine issuing activity. The aim is to identify when significant changes in ADHD medication prescribing or issuing occur. 

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(ggplot2)
library(segmented)
library(performance)
library(arrow)
library(see)
library(forecast)
library(sandwich)
library(lmtest)

source(here("functions", "changepoint_functions.R"))
```

## Data sources 

Data cleaning was performed in `get_foi_data.R`, `get-adhd-meds-op`, and `get-web-oph-adhd-meds.qmd`.

### FOI data
We will start with FP34PCD prescribing FOI data from 2015-08-01 to 2025-07-01 as it is the most simple. Note that only prescription item counts are available and DDD quantity cannot be calculated without quantity information. Here we calculate monthly prescribing for indiviudal chemicals and all chemicals. 

```{r}
#| label: calculate-monthly-prescribing
#| output: false

# Deliberately not using functions to simplify this as keeping the intermediate dfs is useful for plotting later. This will not be the final plot but is good for an intuitive visual check of model fit. 

foi_by_chemical <- read_csv(here("output", "foi_prescribing", "adhd_foi_by_chemical.csv"))

foi_all_chemicals <- read_csv(here("output", "foi_prescribing", "adhd_foi_all_chemicals.csv"))

total_foi_over_time <- read_csv(here("output", "foi_prescribing", "adhd_foi_total.csv")) |> 
  # Add column for month number starting from 0, this makes interpreting the model easier
  group_by(bnf_chemical_name) |> 
  mutate(month_number = row_number() - 1) |> 
  ungroup()
```

Check for missing data: if every chemical (6 in total: the 5 medications + "all") has been prescribed at least once per month, the resulting total_foi_over_time df would have 120 * 6 = 720 rows. 
```{r}
n_distinct(total_foi_over_time$month)
# [1] 120
# As 120 * 6 = 720 > 659, not every chemical was prescribed each month.

total_foi_over_time[total_foi_over_time$bnf_chemical_name == "Atomoxetine", ][1, ]
# Date: 2015-12-01

total_foi_over_time[total_foi_over_time$bnf_chemical_name == "Guanfacine", ][1, ]
# Date: 2016-06-01
```


Note that data for atomoxetine and guanfacine (not controlled drugs) only starts being reported on 2015-12-01 and 2016-06-01 respectively. This does not entirely account for the rows of missing data so more thorough checks are needed if analysing individual chemicals.

### GP prescribing data

Analysing GP prescribing data from 2015-08-01 to 2025-07-01. Using items here as calculating DDDs is non-trivial and not the best use of limited time. 

```{r}
gp_by_chemical <- read_csv(here("output", "monthly_op_gp", "adhd_gp_by_chemical.csv"))

gp_all_chemicals <- read_csv(here("output", "monthly_op_gp", "adhd_gp_all_chemicals.csv"))

total_gp_over_time <- read_csv(here("output", "monthly_op_gp", "adhd_gp_total.csv")) |> 
  # Add column for month number starting from 0, this makes interpreting the model easier
  group_by(bnf_chemical_name) |> 
  mutate(month_number = row_number() - 1) |> 
  ungroup()

# Checking for missing data
n_distinct(total_gp_over_time$month)
# [1] 120
# As 120 * 6 = 720 > 714, not every chemical was prescribed each month.

total_gp_over_time[total_gp_over_time$bnf_chemical_name == "Guanfacine", ][1, ]
# Date: 2016-02-01
```

Guanfacine prescribing was not recorded until 2016-02-01. The 6 months from 2015-07-01 to 2016-01-01 account for all missing rows. 

### OP Hospitals SCMD

Adding SCMD data (from the OpenPrescribing-Hospitals website) from 2019-01-01 to 2025-12-01.

```{r}
#| label: calculate-scmd-monthly-issuing
#| output: false

scmd_by_chemical <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_by_chemical.csv"))

scmd_all_chemicals <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_all_chemicals.csv"))

total_scmd_over_time <- read_csv(here("output", "monthly_oph_scmd", "adhd_scmd_total.csv")) |> 
  group_by(bnf_chemical_name) |>   
# Add column for month number starting from 0, this makes interpreting the model easier
  mutate(month_number = row_number() - 1) |> 
  ungroup()

# Same logic for missing data check
n_distinct(total_scmd_over_time$month)
# [1] 84
# As 84 * 6 = 504 (= number of rows), each chemical has been issued at least once each month.
```

## Change point analysis

Final model choice for each dataset:

- FOI data: fit linear segmented model (2 breakpoints) with log-transformed y-axis (monthly items)
- GP data: fit linear segmented model (1 breakpoint) with log-transformed y-axis (monthly items) and monthly dummy factors
- OPH SCMD data: fit linear segmented model (3 breakpoints) with UNTRANSFORMED y-axis (monthly DDD quantity)

I am currently only analysing all chemicals combined,  but the code is written in a way that should allow for change point analysis of individual chemicals for each dataset.

# Results

## FOI FP34PCD data

### All chemicals 

A linear segmented model (2 breakpoints) with a log-transformed y-axis (monthly items) is the best model choice. 

First, we can use `log_mod_select` to fit log-transformed models with 0, 1, 2, and 3 breakpoints, then select the model with the lowest AIC. It is based on `selgmented` in the `segmented` package. It produces very similar results to Andrea's `mod_aic` function but helps me avoid mistakes with manually inputing the number of changepoints.

#### Fitting a log-transformed linear segmented model for FOI data

```{r}
foi_all_log_model <- log_mod_select(total_foi_over_time, "All chemicals", "monthly_items")
# 2 breakpoints has lowest AIC (0.4490)
summary(foi_all_log_model)
# Estimated break points: 59.999 (SE = 1.133), 64.117 (SE = 1.383)

# Check model fit
check_model(foi_all_log_model)
# Residuals vs fitted values mostly linear
# A bit of heteroskedasticity but Andrea says it doesn't look too bad
# 2 outliers detected but only just outside significance bounds
# High collinearity for breakpoint estimates but normal for breakpoint models
# Overall, doesn't look too bad according to Andrea

acf(residuals(foi_all_log_model))
# No significant autocorrelation present

plot(foi_all_log_model)

# Convert FOI data into `ts` object for seasonality plot
ts_foi_all <- ts(data = foi_all_chemicals$monthly_items, frequency = 12, start = c(2015, 8), end = c(2025, 7))
# Season plot
ggseasonplot(ts_foi_all)
# No clear seasonality pattern
```

#### Calculating predicted values and slope

`predicted_exp` is a function based on Andrea's `predicted` functions but modified to deal with log-transformed data. It generates a df using the model with predicted monthly items and 95% CI bounds over time.

`extract_exp_slopes` extracts the slopes and 95% CIs for each segment, exponentiates them, then outputs the interpretation of slopes as a percentage change with 95% CIs. 

```{r}
pred_foi_all_log <- predicted_exp(
  foi_all_log_model, "All chemicals", total_foi_over_time, "FOI"
)

slope_foi_all_log <- extract_exp_slopes(foi_all_log_model, "All chemicals", "FOI") |> 
  # Round breakpoint locations
  mutate(bp1 = round(bp1), bp2 = round(bp2)) |> 
  # Convert to date
  mutate(
    bp1 = foi_all_chemicals$month[bp1 + 1],
    bp2 = foi_all_chemicals$month[bp2 + 1]
  )
```

#### Plotting

Plotting the log-transformed model with observed data.

```{r}
# Simple plot to check model fit

ggplot() +
  geom_point(
    data = foi_all_chemicals,
    aes(x = month, y = monthly_items, colour = "Observed"),
    alpha = .7
  ) +
  geom_vline(xintercept = c(slope_foi_all_log$bp1, slope_foi_all_log$bp2), colour = c("red", "red"), alpha = 0.5) +
  geom_line(
    data = pred_foi_all_log,
    aes(x = month, y = pred, colour = "Segmented log-linear model"),
    size = 1
  ) +
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented log-linear model" = "red"
    )
  ) +
  annotate(x = ymd("2020-08-01"), y = 55000, label = "2020-08-01", geom = "label", colour = "red") +
  annotate(x = ymd("2020-12-01"), y = 45000, label = "2020-12-01", geom = "label", colour = "red") +
  theme(legend.position = "right")
```

Overall it's a reasonable fit.  

## GP Prescribing Data (items)

### All chemicals

Fitting a linear segmented model (2 breakpoints) with a log-transformed y-axis (monthly items) and dummy monthly factors is the best choice for GP data. A log-linear model with dummy monthly factors for each month provides a better fit than a linear model and a log-linear model with dummy monthly factors for only months that are significantly different from January. 

#### Fitting a log-transformed linear segmented model with dummy monthly factors for GP data

I will start with a seasonality plot to observe seasonal patterns.

```{r}
# Convert GP data into `ts` object for seasonality plot
ts_gp_all <- ts(data = gp_all_chemicals$monthly_items, frequency = 12, start = c(2015, 8), end = c(2025, 7))
# Season plot
ggseasonplot(ts_gp_all)
# Generally, there seems to be a spike in March and a dip in August
```

Now to fit the model (same logic as FOI data, but with dummy variables): 

```{r}
# Create dummy month factors
total_gp_over_time$month_factor <- factor(
  format(total_gp_over_time$month, "%b"),
  levels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
)

# Can't use `log_mod_select` since I want to include dummy monthly factors in the model
gp_all_months_model0 <- lm(
  log(monthly_items) ~ month_number + month_factor,
  data = total_gp_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

# Use `selgmented` to fit optimal log-linear model like in `log_mod_select`
gp_all_months_model <- selgmented(
  gp_all_months_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 3, # max number of changepoints = 3
  check.dslope = T # removes breakpoints with small slope differences
)
# 2 breakpoints (after removing 1 due to small slope difference) has lowest AIC (-486.6976) which is lower than the model without the dummy month variable. 

summary(gp_all_months_model) 
# Estimated breakpoints (SE): 32.132 (5.786), 66.000 (3.226)
# SE for breakpoints might be optimistic due to autocorrelation but Newey-West can't adjust for breakpoint SEs
# SE and p values for other coefficients not accurate at this point due to autocorrelation 
# Will use Newey-West adjustment later

# Check model fit
check_model(gp_all_months_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions and no significant outliers. 
# Appears to be a better fit than without the monthly dummy variable.

acf(residuals(gp_all_months_model))
# Autocorrelation still present

plot(gp_all_months_model)
```

This log-transformed linear segmented model is a good model choice, but there is still some autocorrelation. So, I will use Newey-West heteroscedasticity and autocorrelation consistent (HAC) covaraince matrix estimators to adjust 95% confidence intervals (95% CIs) and p values. I will then use the Holm corection for multiple comparisons.

```{r}
# Automatically select lag (bandwidth) according to the model
gp_nw <- NeweyWest(gp_all_months_model, lag = NULL, prewhite = FALSE, verbose = T)
# Lag = 5 using Bartlett kernels for automatic bandwidth selection

# Extract Newey-West covariance matrix
gp_nw_variance <- NeweyWest(gp_all_months_model, lag = 5, prewhite = F)

# Newey-West adjusted p values (before multiple testing correction)
gp_nw_coef_pvals <- coeftest(gp_all_months_model, vcov. = gp_nw_variance)[,4]

# Add Holm correction for multiple testing to identify months with a significant effect
gp_holm_coef_pvals <- p.adjust(gp_nw_coef_pvals, method = "holm")

# Sanity test for Newey-West corrected 95% CIs
round(coefci(gp_all_months_model), digits = 5)
round(coefci(gp_all_months_model, vcov. = gp_nw_variance), digits = 5)
# CIs are sometimes wider and sometimes narrower after Newey-West correction, which can sometimes happen

# Construct df with Newey-West corrected 95% CI/p values and Holm corrected p values for model 'intercepts' (step changes)
df_gp_nw_coef <- data.frame(
  est = exp(gp_all_months_model$coef),
  exp(coefci(gp_all_months_model, vcov. = gp_nw_variance))
) |> 
  rownames_to_column("coef") |> 
  rename(lci = "X2.5..", uci = "X97.5..") |> 
  mutate(
    est = round(est, 5),
    lci = round(lci, 5),
    uci = round(uci, 5),
    pct_est = round((est - 1) * 100, digits = 2),
    pct_lci = round((lci - 1) * 100, digits = 2),
    pct_uci = round((uci - 1) * 100, digits = 2),
    # Add adjusted p-values
    holm_pvals = round(gp_holm_coef_pvals, digits = 5),
    # Identifier columns 
    data_source = "GP",
    bnf_chemical = "All chemicals"
  ) |>
  # Remove irrelevant rows
  filter_out(coef %in% c("psi1.month_number", "psi2.month_number")) |> 
  # Only show dummy month factors with a sig. effect
  filter(holm_pvals < 0.0500) |> 
  relocate(holm_pvals, .after = coef)
```

Only Feb, Apr, and Aug are significantly different to Jan according to Newey-West p values adjusted for multiple corrections.

#### Calculating predicted values and slope

```{r}
# Need to calculate Newey-West SE to calculate 95%CIs for predicted values
# Extract model matrix
gp_mod_matrix <- model.matrix(gp_all_months_model)
View(as.data.frame(gp_mod_matrix))
# Mostly 0s and 1s and looks as expected

# Check that ncol of model matrix matches nrow NW variance matrix, important for matrix multiplication step
ncol(gp_mod_matrix) == nrow(gp_nw_variance)
# [1] TRUE

# Matrix multiplication to get SE for each predicted value
gp_nw_se <- sqrt(rowSums((gp_mod_matrix %*% gp_nw_variance) * gp_mod_matrix))

# Use `predicted_exp` to get prediction df, same as FOI data except I specify NW SEs
pred_gp_all_months <- predicted_exp(
  gp_all_months_model, "All chemicals", total_gp_over_time, "GP", gp_nw_se
)

# Use `extract_exp_slopes` to get slopes df, same as FOI data except I specify NW variance matrix
slope_gp_all_months <- extract_exp_slopes(gp_all_months_model, "All chemicals", "GP", gp_nw_variance) |> 
  # Round breakpoint locations
  mutate(bp1 = round(bp1), bp2 = round(bp2)) |> 
  # Convert to date
  mutate(
    bp1 = gp_all_chemicals$month[bp1 + 1],
    bp2 = gp_all_chemicals$month[bp2 + 1]
  )
```

#### Plotting log-transformed models with observed data
```{r}
# Simple plot to visually check model fit
ggplot() +
  geom_point(
    data = gp_all_chemicals,
    aes(x = month, y = monthly_items, colour = "Observed"),
    alpha = .7
  ) +
  geom_vline(xintercept = c(slope_gp_all_months$bp1, slope_gp_all_months$bp2), colour = "red", alpha = 0.5) +
  geom_line(
    data = pred_gp_all_months,
    aes(x = month, y = pred, colour = "Segmented log-linear model w/ month dummies")
  ) +  
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented log-linear model w/ month dummies" = "red"
    )
  ) +
  annotate(x = ymd("2018-04-01"), y = 175000, label = "2018-04-01", geom = "label", colour = "red") +
  annotate(x = ymd("2021-02-01"), y = 175000, label = "2021-02-01", geom = "label", colour = "red") +
  theme(legend.position = "right")
```

Looks like a decent fit. Adding a dummy variable improves the fit visually (although change points are slightly harder to see)compared to the log-linear model with no dummy months (not plotted) but doesn't improve autocorrelation. Therefore, the Newey-West heteroscedasticity and autocorrelation consistent (HAC) covariance matrix estimators were used to re-estimate 95% CIs and p-values.     

## SCMD Issuing Data (DDD quantity)

### All chemicals

Fitting a linear segmented model (3 breakpoints) with an UNTRANSFORMED y-axis (monthly DDD quantity) is the best choice. Although models with monthly dummies or just dummies for significant months have slightly lower AIC, those models only find 2 breakpoints which is a poorer visual fit for the data. 

For now, I'm keeping the code for the monthly dummy model because it shows that there's a significant dip in prescribing in August (compared to January). Could probably just do a seasonal test or what later. 

```{r}
# Create initial model
# `log_mod_select` or `mod_select` won't work for this because I wamt a linear model 
scmd_all_model0 <- lm(
  monthly_ddd_quantity ~ month_number, 
  data = total_scmd_over_time, 
  subset = bnf_chemical_name == "All chemicals"
)

# `log_mod_select` or `mod_select` won't work for this because I don't want a Poisson model 
# Using `selgmented` to fit the model with 3 breakpoints - this will generate very similar results to Andrea's `mod_aic`

scmd_all_lin3_model <- selgmented(
  scmd_all_model0,
  type = "aic",
  refit = T,
  Kmax = 3,
  check.dslope = T
  )
# 3 breakpoints has lowest AIC (1828.880)
# If I use Kmax = 5, 3 breakpoints still has the lowest AIC

summary(scmd_all_lin3_model)
# Estimated break points: 12.948 (SE = 0.678), 15.694 (SE = 0.773), 70.992 (SE = 0.942)
```

Checking model fit and assumptions 

```{r}
acf(residuals(scmd_all_lin3_model))
# A tiny bit of significant autocorrelation at lag = 6, 7, 12 and 13 but not too bad. Possibly a seasonal effect.

check_model(scmd_all_lin3_model)
# Reasonable linearity, homogeneity of variance, and normality of residuals. 2 outliers detected. Some high collinearity but not a big concern. 
```

Seasonality plot to observe seasonal patterns:

```{r}
# Convert SCMD data into `ts` object for seasonality plot
ts_scmd_all <- ts(data = scmd_all_chemicals$monthly_ddd_quantity, frequency = 12, start = c(2019, 1), end = c(2025, 12))
# Season plot
ggseasonplot(ts_scmd_all)
# Seasonal effect less obvious but there seems to be a dip in August
```

As for the GP data, due to the presence of autocorrelation, Newey-West adjustments for 95% confidence intervals (95% CIs) and p values. I will then use the Holm corection for multiple comparisons.

```{r}
# Automatically select lag (bandwidth) according to the model
scmd_lin3_nw <- NeweyWest(scmd_all_lin3_model, lag = NULL, prewhite = FALSE, verbose = T)
# Lag = 8 using Bartlett kernels for automatic bandwidth selection

# Extract Newey-West covariance matrix
scmd_lin3_nw_variance <- NeweyWest(scmd_all_lin3_model, lag = 8, prewhite = F)

# Newey-West adjusted p values (before multiple testing correction)
scmd_lin3_nw_coef_pvals <- coeftest(scmd_all_lin3_model, vcov. = scmd_lin3_nw_variance)[,4]

# Add Holm correction for multiple testing to identify months with a significant effect
scmd_lin3_holm_coef_pvals <- p.adjust(scmd_lin3_nw_coef_pvals, method = "holm")

# Sanity test for Newey-West corrected 95% CIs
round(coefci(scmd_all_lin3_model), digits = 5)
round(coefci(scmd_all_lin3_model, vcov. = scmd_lin3_nw_variance), digits = 5)
# CIs are all narrower after Newey-West correction, which is unusual but can happen

# Construct df with Newey-West corrected 95% CI/p values and Holm corrected p values for model 'intercepts' (step changes)
# Same logic as for GP data but no need to exponentiate
# And no need to interpret coefficients as percentage changes (they are absolute changes for a linear model)
df_scmd_lin3_nw_coef <- data.frame(
  est = scmd_all_lin3_model$coef,
  coefci(scmd_all_lin3_model, vcov. = scmd_lin3_nw_variance)
) |> 
  rownames_to_column("coef") |> 
  rename(lci = "X2.5..", uci = "X97.5..") |> 
  mutate(
    est = round(est, 5),
    lci = round(lci, 5),
    uci = round(uci, 5),
    # Add adjusted p-values
    holm_pvals = round(scmd_lin3_holm_coef_pvals, digits = 5),
    # Identifier columns 
    data_source = "SCMD",
    bnf_chemical = "All chemicals"
  ) |>
  # Remove irrelevant rows
  filter_out(coef %in% c("psi1.month_number", "psi2.month_number", "psi3.month_number")) |> 
  # Only show dummy month factors with a sig. effect
  filter(holm_pvals < 0.0500) |> 
  relocate(holm_pvals, .after = coef)
```


Now to fit the model with dummy month factors: 

```{r}
total_scmd_over_time$month_factor <- factor(
  format(total_scmd_over_time$month, "%b"),
  levels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
)

scmd_all_months_model0 <- lm(
  monthly_ddd_quantity ~ month_number + month_factor,
  data = total_scmd_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

scmd_all_months_model <- selgmented(
  scmd_all_months_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 2, # max number of changepoints = 2 as 3 breakpoints makes calculation fail
  check.dslope = T 
)
# 2 breakpoints has lowest AIC (1822.564), which is lower than the model without the dummy month factor

summary(scmd_all_months_model) 
# Estimated breakpoints: 23.933 (SE = 1.611), 70.805 (SE = 0.989)
# Significant effect for Aug only (uncorrected p-values)

# Check model fit
check_model(scmd_all_months_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions; similar to model w/o monthly dummy variable
# No significant outliers; slightly better fit than model w/o monthly dummy variable in this sense 

acf(residuals(scmd_all_months_model))
# Autocorrelation still present at lag = 1, 7, and 17

plot(scmd_all_months_model)
```

As for the GP data, due to the presence of autocorrelation, Newey-West adjustments for 95% confidence intervals (95% CIs) and p values. I will then use the Holm corection for multiple comparisons.

```{r}
# Automatically select lag (bandwidth) according to the model
scmd_nw <- NeweyWest(scmd_all_months_model, lag = NULL, prewhite = FALSE, verbose = T)
# Lag = 4 using Bartlett kernels for automatic bandwidth selection

# Extract Newey-West covariance matrix
scmd_nw_variance <- NeweyWest(scmd_all_months_model, lag = 4, prewhite = F)

# Newey-West adjusted p values (before multiple testing correction)
scmd_nw_coef_pvals <- coeftest(scmd_all_months_model, vcov. = scmd_nw_variance)[,4]

# Add Holm correction for multiple testing to identify months with a significant effect
scmd_holm_coef_pvals <- p.adjust(scmd_nw_coef_pvals, method = "holm")

# Sanity test for Newey-West corrected 95% CIs
round(coefci(scmd_all_months_model), digits = 5)
round(coefci(scmd_all_months_model, vcov. = scmd_nw_variance), digits = 5)
# CIs are sometimes wider and sometimes narrower after Newey-West correction, which can happen

# Construct df with Newey-West corrected 95% CI/p values and Holm corrected p values for model 'intercepts' (step changes)
# Same logic as for GP data but no need to exponentiate
# And no need to interpret coefficients as percentage changes (they are absolute changes for a linear model)
df_scmd_nw_coef <- data.frame(
  est = scmd_all_months_model$coef,
  coefci(scmd_all_months_model, vcov. = scmd_nw_variance)
) |> 
  rownames_to_column("coef") |> 
  rename(lci = "X2.5..", uci = "X97.5..") |> 
  mutate(
    est = round(est, 5),
    lci = round(lci, 5),
    uci = round(uci, 5),
    # Add adjusted p-values
    holm_pvals = round(scmd_holm_coef_pvals, digits = 5),
    # Identifier columns 
    data_source = "SCMD",
    bnf_chemical = "All chemicals"
  ) |>
  # Remove irrelevant rows
  filter_out(coef %in% c("psi1.month_number", "psi2.month_number")) |> 
  # Only show dummy month factors with a sig. effect
  filter(holm_pvals < 0.0500) |> 
  relocate(holm_pvals, .after = coef)

# Only Aug is significantly different to Jan according to Newey-West p values adjusted for multiple corrections.
```

#### Calculating predicted values and slope

Use `predicted` for calculating predicted values and `extract_lin_slopes` for calculating 95% CI bounds for each segment. Both functions are the linear equivalent of the functions used above. Predicted values and slopes are not calculated or plotted for models other than the 3-breakpoint linear model. 

```{r}
# Need to calculate Newey-West SE to calculate 95%CIs
# Extract model matrix
scmd_mod_matrix <- model.matrix(scmd_all_lin3_model)
View(as.data.frame(scmd_mod_matrix))
# Mostly 0s and 1s and looks as expected

# Check that ncol of model matrix matches nrow NW variance matrix, important for matrix multiplication step
ncol(scmd_mod_matrix) == nrow(scmd_lin3_nw_variance)
# [1] TRUE

# Matrix multiplication to get SE for each predicted value
scmd_lin3_nw_se <- sqrt(rowSums((scmd_mod_matrix %*% scmd_lin3_nw_variance) * scmd_mod_matrix))

# Use functions to get prediction and slopes dfs, specifying NW SE/variance
pred_scmd_all_lin3 <- predicted(scmd_all_lin3_model, "All chemicals", total_scmd_over_time, "SCMD", scmd_lin3_nw_se)

slope_scmd_all_lin3 <- extract_lin_slopes(scmd_all_lin3_model, "All chemicals", "SCMD", scmd_lin3_nw_variance) |> 
  # Round breakpoint locations
  mutate(bp1 = round(bp1), bp2 = round(bp2), bp3 = round(bp3)) |> 
  # Convert to date
  mutate(
    bp1 = scmd_all_chemicals$month[bp1 + 1],
    bp2 = scmd_all_chemicals$month[bp2 + 1],
    bp3 = scmd_all_chemicals$month[bp3 + 1],
  )
```

#### Plotting the linear model with observed data.

```{r}
# Simple plot to visually check model fit

ggplot() +
  geom_point(
    data = scmd_all_chemicals,
    aes(x = month, y = monthly_ddd_quantity, colour = "Observed"),
    alpha = .7
  ) +
  geom_line(
    data = pred_scmd_all_lin3,
    aes(x= month, y = pred, colour = "Segmented linear model (3bps)"),
    size = 1,
    alpha = .7
  ) +
  geom_vline(
    xintercept = c(slope_scmd_all_lin3$bp1, slope_scmd_all_lin3$bp2, slope_scmd_all_lin3$bp3), 
    colour = "blue", 
    alpha = 0.5
  ) +
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented linear model (3bps)" = "blue"
    )
  ) +
  annotate(x = ymd("2020-02-01"), y = 200000, label = "2020-02-01", geom = "label", colour = "blue") +
  annotate(x = ymd("2020-05-01"), y = 175000, label = "2020-05-01", geom = "label", colour = "blue") +
  annotate(x = ymd("2024-12-01"), y = 100000, label = "2024-12-01", geom = "label", colour = "blue") +
  theme(legend.position = "right")
```

The two close breakpoints (2020-02-01 and 2020-05-01) can be interpreted as an 'unusual' period due to the COVID-19 pandemic. 

## Export useful things for plotting

Combining predicted values and slope dataframes and exporting as csv

```{r}
pred_all <- mget(ls(pattern = "^pred_")) |> 
  bind_rows()

write.csv(pred_all, here("output", "changepoint_analysis", "predicted_values.csv"), row.names = FALSE)

slope_all <- mget(ls(pattern = "^slope_")) |> 
  bind_rows() 

write.csv(slope_all, here("output", "changepoint_analysis", "model_slopes.csv"), row.names = FALSE)
```