---
title: "Change point analysis of ADHD prescribing activity"
author:
  - Kaylee Poon
format: html
---
# Overview
This analysis uses the `segmented` R package to conduct change point analysis on the rate of change of ADHD prescribing or medicine issuing activity. The aim is to identify when significant changes in ADHD medication prescribing or issuing occur. 

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(ggplot2)
library(segmented)
library(performance)
library(arrow)
library(see)
library(forecast)

source(here("functions", "changepoint_functions.R"))

foi_adhd_prescribing <- read_csv(
  here(
    "output",
    "foi_prescribing",
    "adhd_foi_prescribing.csv"
  )
)

gp_adhd_prescribing <- read_parquet(
  here(
    "output",
    "openprescribing",
    "adhd_gp_prescribing_chemical.parquet"
  )
)

scmd_adhd_issuing <- read_csv(
  here(
    "output",
    "web-openprescribing-hospitals",
    "adhd_oph_web_ddd_quantity.csv"
  )
) |> 
  janitor::clean_names() |> 
  rename(
    month = date,
    bnf_chemical_name = vtm_name
  ) 
```

## Data sources 

### FOI data
We will start with FP34PCD prescribing FOI data from 2015-08-01 to 2025-07-01 as it is the most simple. Note that only prescription item counts are available and DDD quantity cannot be calculated without quantity information. Here we calculate monthly prescribing for indiviudal chemicals and all chemicals. 

```{r}
#| label: calculate-monthly-prescribing
#| output: false

# Deliberately not using functions to simplify this as keeping the intermediate dfs is useful for plotting later. This will not be the final plot but is good for an intuitive visual check of model fit. 
# Individual chemicals
foi_by_chemical <- foi_adhd_prescribing |> 
  dplyr::select(month, items, bnf_chemical_name) |>
  group_by(month, bnf_chemical_name) |> 
  summarise(monthly_items = sum(items, na.rm = TRUE), .groups = "drop")

# All chemicals
foi_all_chemicals <- foi_adhd_prescribing |>
  dplyr::select(month, items) |>
  group_by(month) |>
  summarise(monthly_items = sum(items, na.rm = TRUE), .groups = "drop") |>
  mutate(bnf_chemical_name = "All chemicals")

# Combine individual and all chemicals
total_foi_over_time <- bind_rows(foi_by_chemical, foi_all_chemicals) |> 
  arrange(month, bnf_chemical_name) |>
  group_by(bnf_chemical_name) |>
  # Add column for month number starting from 0, this makes interpreting the model easier
  mutate(month_number = row_number() - 1) |> 
  ungroup()
```

Check for missing data: if every chemical (6 in total: the 5 medications + "all") has been prescribed at least once per month, the resulting total_foi_over_time df would have 120 * 6 = 720 rows. 
```{r}
n_distinct(total_foi_over_time$month)
# [1] 120
# As 120 * 6 = 720 > 659, not every chemical was prescribed each month.

total_foi_over_time[total_foi_over_time$bnf_chemical_name == "Atomoxetine", ][1, ]
# Date: 2015-12-01

total_foi_over_time[total_foi_over_time$bnf_chemical_name == "Guanfacine", ][1, ]
# Date: 2016-06-01
```


Note that data for atomoxetine and guanfacine (not controlled drugs) only starts being reported on 2015-12-01 and 2016-06-01 respectively. This does not entirely account for the rows of missing data so more thorough checks are needed if analysing individual chemicals.

### GP prescribing data

Analysing GP prescribing data from 2015-08-01 to 2025-07-01. Using items here as calculating DDDs is non-trivial and not the best use of limited time. 

```{r}
#| label: calculate-monthly-gp-prescribing
#| output: false

# Same logic as calculating FOI monthly prescribing but with slightly different column names
# Individual chemicals
gp_by_chemical <- gp_adhd_prescribing |> 
  dplyr::select(month, items_by_chemical, bnf_chemical_name) |>
  group_by(month, bnf_chemical_name) |> 
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop")

gp_all_chemicals <- gp_adhd_prescribing |>
  dplyr::select(month, items_by_chemical) |>
  group_by(month) |>
  summarise(monthly_items = sum(items_by_chemical, na.rm = TRUE), .groups = "drop") |>
  mutate(bnf_chemical_name = "All chemicals")

total_gp_over_time <- bind_rows(gp_by_chemical, gp_all_chemicals) |> 
  arrange(month, bnf_chemical_name) |>
  group_by(bnf_chemical_name) |>
  # Add column for month number starting from 0, this makes interpreting the model easier
  mutate(month_number = row_number() - 1) |> 
  ungroup()

# Checking for missing data
n_distinct(total_gp_over_time$month)
# [1] 120
# As 120 * 6 = 720 > 714, not every chemical was prescribed each month.

total_gp_over_time[total_gp_over_time$bnf_chemical_name == "Guanfacine", ][1, ]
# Date: 2016-02-01
```

Guanfacine prescribing was not recorded until 2016-02-01. The 6 months from 2015-07-01 to 2016-01-01 account for all missing rows. 

### OP Hospitals SCMD

Adding SCMD data (from the OpenPrescribing-Hospitals website) from 2019-01-01 to 2025-12-01.

```{r}
#| label: calculate-scmd-monthly-issuing
#| output: false

# Same logic as above but different quantity measure
scmd_by_chemical <- scmd_adhd_issuing |> 
  group_by(month, bnf_chemical_name) |> 
  summarise(monthly_ddd_quantity = sum(value, na.rm = TRUE), .groups = "drop")

scmd_all_chemicals <- scmd_adhd_issuing |>
  group_by(month) |>
  summarise(monthly_ddd_quantity = sum(value, na.rm = TRUE), .groups = "drop") |>
  mutate(bnf_chemical_name = "All chemicals")

total_scmd_over_time <- bind_rows(scmd_by_chemical, scmd_all_chemicals) |> 
  arrange(month, bnf_chemical_name) |>
  group_by(bnf_chemical_name) |>
  # Add column for month number starting from 0, this makes interpreting the model easier
  mutate(month_number = row_number() - 1,) |> 
  ungroup()

# Same logic for missing data check
n_distinct(total_scmd_over_time$month)
# [1] 84
# As 84 * 6 = 504 (= number of rows), each chemical has been issued at least once each month.
```

## Change point analysis

Final model choice for each dataset:

- FOI data: fit linear segmented model (2 breakpoints) with log-transformed y-axis (monthly items)
- GP data: fit linear segmented model (1 breakpoint) with log-transformed y-axis (monthly items)
- OPH SCMD data: fit linear segmented model (3 breakpoints) with UNTRANSFORMED y-axis (monthly DDD quantity)

I am currently only analysing all chemicals combined,  but the code is written in a way that should allow for change point analysis of individual chemicals for each dataset.

# Results

## FOI FP34PCD data

### All chemicals 

A linear segmented model (2 breakpoints) with a log-transformed y-axis (monthly items) is the best model choice. 

First, we can use `log_mod_select` to fit log-transformed models with 0, 1, 2, and 3 breakpoints, then select the model with the lowest AIC. It is based on `selgmented` in the `segmented` package. It produces very similar results to Andrea's `mod_aic` function but helps me avoid mistakes with manually inputing the number of changepoints.

#### Fitting a log-transformed linear segmented model for FOI data

```{r}
foi_all_log_model <- log_mod_select(total_foi_over_time, "All chemicals", "monthly_items")
# 2 breakpoints has lowest AIC (0.4490)
summary(foi_all_log_model)
# Estimated break points: 59.999, 64.117

# Check model fit
check_model(foi_all_log_model)
# Residuals vs fitted values mostly linear
# A bit of heteroskedasticity but Andrea says it doesn't look too bad
# 2 outliers detected but only just outside significance bounds
# High collinearity for breakpoint estimates but normal for breakpoint models
# Overall, doesn't look too bad according to Andrea

acf(residuals(foi_all_log_model))
# No significant autocorrelation present

plot(foi_all_log_model)
```

#### Calculating predicted values and slope

`predicted_exp` is a function based on Andrea's `predicted` functions but modified to deal with log-transformed data. It generates a df using the model with predicted monthly items and 95% CI bounds over time.

`extract_exp_slopes` extracts the slopes and 95% CIs for each segment, exponentiates them, then outputs the interpretation of slopes as a percentage change with 95% CIs. 

```{r}
pred_foi_all_log <- predicted_exp(
  foi_all_log_model, "All chemicals", total_foi_over_time, "FOI"
)

slope_foi_all_log <- extract_exp_slopes(foi_all_log_model, "All chemicals", "FOI")
```

#### Plotting

Plotting the log-transformed model with observed data.

```{r}
# Rounding break points to nearest month
foi_all_break_months <- round(c(slope_foi_all_log$bp1, slope_foi_all_log$bp2))

# Convert month number into date
foi_all_break_dates <- foi_all_chemicals$month[foi_all_break_months + 1]
# "2020-08-01", "2020-12-01"

ggplot() +
  geom_point(
    data = foi_all_chemicals,
    aes(x = month, y = monthly_items, colour = "Observed"),
    alpha = .7
  ) +
  geom_vline(xintercept = foi_all_break_dates, colour = c("red", "red"), alpha = 0.5) +
  geom_line(
    data = pred_foi_all_log,
    aes(x = month, y = pred, colour = "Segmented log-linear model"),
    size = 1
  ) +
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented log-linear model" = "red"
    )
  ) +
  annotate(x = ymd("2020-08-01"), y = 55000, label = "2020-08-01", geom = "label", colour = "red") +
  annotate(x = ymd("2020-12-01"), y = 45000, label = "2020-12-01", geom = "label", colour = "red") +
  theme(legend.position = "right")
```

Overall, the log model provides the best fit. 

## GP Prescribing Data (items)

### All chemicals
Fitting a linear segmented model (2 breakpoints) with a log-transformed y-axis (monthly items) is the best choice for GP data

#### Fitting a log-transformed linear segmented model for FOI data

This uses the same logic and functions as the FOI data. 

```{r}
gp_all_log_model <- log_mod_select(total_gp_over_time, "All chemicals", "monthly_items")

summary(gp_all_log_model)
# 1 breakpoint has lowest AIC (-399.4453)
# Estimated break point: 60

# Check model fit
check_model(gp_all_log_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions and no significant outliers

acf(residuals(gp_all_log_model))
# Autocorrelation present

plot(gp_all_log_model)
```

This log-transformed linear segmented model is a good model choice but we will need to adjust for autocorrelation.

#### Fitting a log-transformed model with month as a dummy variable

I will start with a seasonality plot to observe seasonal patterns.

```{r}
# Convert GP data into `ts` object for seasonality plot
ts_gp_all <- ts(data = gp_all_chemicals$monthly_items, frequency = 12, start = c(2015, 8), end = c(2025, 7))
# Season plot
ggseasonplot(ts_gp_all)
# Generally, there seems to be a spike in March and a dip in August
```

Now to fit the model: 

```{r}
total_gp_over_time$month_factor <- factor(
  format(total_gp_over_time$month, "%b"),
  levels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
)

gp_all_seasonal_model0 <- lm(
  log(monthly_items) ~ month_number + month_factor,
  data = total_gp_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

gp_all_seasonal_model <- selgmented(
  gp_all_seasonal_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 3, # max number of changepoints = 3
  check.dslope = F # ensures only AIC is used to select model
)
# 3 breakpoints has lowest AIC (-486.8576) which is lower than the model without the dummy month variable.

summary(gp_all_seasonal_model) 
# Estimated breakpoints: 32.202, 69.014, 91.001
# Significant effect for Feb, Mar, Apr, and Aug

# Check model fit
check_model(gp_all_seasonal_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions and no significant outliers. 
# Appears to be a better fit than without the monthly dummy variable.

acf(residuals(gp_all_seasonal_model))
# Autocorrelation still present

plot(gp_all_seasonal_model)
```

Version with only months with a significant effect as a factor

```{r}
total_gp_over_time$sigmonth_factor <- if_else(
  format(total_gp_over_time$month, "%b") %in% c("Feb", "Mar", "Apr", "Aug"), 
  "sig", 
  "not_sig"
) |> 
  factor(levels = c("sig", "not_sig"))

gp_all_sigmonth_model0 <- lm(
  log(monthly_items) ~ month_number + sigmonth_factor,
  data = total_gp_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

gp_all_sigmonth_model <- selgmented(
  gp_all_sigmonth_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 3, # max number of changepoints = 3
  check.dslope = F # ensures only AIC is used to select model
)
# 2 breakpoints has lowest AIC (-416.9572) which is higher than the model without the dummy month variable.

summary(gp_all_sigmonth_model) 
# Estimated breakpoints: 36.001, 65.918
# Significant effect for sigmonths

# Check model fit
check_model(gp_all_sigmonth_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions and no significant outliers. 

acf(residuals(gp_all_sigmonth_model))
# Autocorrelation still present, big spike at lag = 12

plot(gp_all_sigmonth_model)
```

Version with each significant month as a separate dummy variable

```{r}
# Feb, Mar, Apr, and Aug
total_gp_over_time$sigmonth_factors <- format(total_gp_over_time$month, "%b") |> 
  factor() |> 
  fct_other(keep = c("Feb", "Mar", "Apr", "Aug"), other_level = "not_sig")

gp_all_sigmonths_model0 <- lm(
  log(monthly_items) ~ month_number + sigmonth_factors,
  data = total_gp_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

gp_all_sigmonths_model <- selgmented(
  gp_all_sigmonths_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 3, # max number of changepoints = 3
  check.dslope = F # ensures only AIC is used to select model
)
# 2 breakpoints has lowest AIC (-496.9207) which is lower than the model without the dummy month variable.

summary(gp_all_sigmonths_model) 
# Estimated breakpoints: 34.965, 66.000
# Significant effect for sigmonth_factors

# Check model fit
check_model(gp_all_sigmonths_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions and no significant outliers. 

acf(residuals(gp_all_sigmonths_model))
# Autocorrelation still present, big spike at lag = 12

plot(gp_all_sigmonths_model)
```

#### Calculating predicted values and slope

```{r}
pred_gp_all_log <- predicted_exp(
  gp_all_log_model, "All chemicals", total_gp_over_time, "GP"
)

slope_gp_all_log <- extract_exp_slopes(gp_all_log_model, "All chemicals", "GP")

pred_gp_all_seasonal <- predicted_exp(
  gp_all_seasonal_model, "All chemicals", total_gp_over_time, "GP"
)

slope_gp_all_seasonal <- extract_exp_slopes(gp_all_seasonal_model, "All chemicals", "GP")
```

#### Plotting log-transformed models with observed data
```{r}
# Rounding breakpoints to nearest month
gp_log_break_months <- round(c(slope_gp_all_log$bp1, slope_gp_all_seasonal$bp1, slope_gp_all_seasonal$bp2, slope_gp_all_seasonal$bp3))

# Convert month number into date
gp_all_break_dates <- gp_all_chemicals$month[gp_log_break_months + 1]
# "2020-08-01", "2018-04-01", "2021-05-01", "2023-03-01"

ggplot() +
  geom_point(
    data = gp_all_chemicals,
    aes(x = month, y = monthly_items, colour = "Observed"),
    alpha = .7
  ) +
  geom_vline(xintercept = gp_all_break_dates, colour = "red", alpha = 0.5) +
  geom_line(
    data = pred_gp_all_log,
    aes(x = month, y = pred, colour = "Segmented log-linear model")
  ) +
  geom_line(
    data = pred_gp_all_seasonal,
    aes(x = month, y = pred, colour = "Segmented seasonal log-linear model")
  ) +  
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented log-linear model" = "red",
      "Segmented seasonal log-linear model" = "blue"
    )
  ) +
  annotate(x = ymd("2020-08-01"), y = 175000, label = "2020-08-01", geom = "label", colour = "red") +
  theme(legend.position = "right")
```

Looks like a decent fit. Adding a dummy variable improves the fit a bit visually (but change points are harder to see) but doesn't improve autocorrelation. I thihnk I need to explicitly adjust for autocorrelation.  

## SCMD Issuing Data (DDD quantity)

Fitting a linear segmented model (3 breakpoints) with an UNTRANSFORMED y-axis (monthly DDD quantity) is the best choice. 

```{r}
#| label: create-scmd-linear-model
#| output: false

# Create initial model
scmd_all_model0 <- lm(
  monthly_ddd_quantity ~ month_number, 
  data = total_scmd_over_time, 
  subset = bnf_chemical_name == "All chemicals"
)

# `log_mod_select` or `mod_select` won't work for this because I don't want a Poisson model 
# Using `selgmented` to fit the model with 3 breakpoints - this will generate very similar results to Andrea's `mod_aic`

scmd_all_lin3_model <- selgmented(
  scmd_all_model0,
  type = "aic",
  refit = T,
  Kmax = 3,
  check.dslope = F
  )
# 3 breakpoints has lowest AIC (1828.880)
# If I use Kmax = 5, 3 breakpoints still has the lowest AIC

summary(scmd_all_lin3_model)
# Estimated break points: 12.948, 15.694, 70.992
```

Checking model fit and assumptions 

```{r}
acf(residuals(scmd_all_lin3_model))
# A tiny bit of significant autocorrelation at lag = 6, 7, 12 and 13 but not too bad. Possibly a seasonal effect.

check_model(scmd_all_lin3_model)
# Reasonable linearity, homogeneity of variance, and normality of residuals. 2 outliers detected. Some high collinearity but not a big concern. 
```

Again, starting start with a seasonality plot to observe seasonal patterns.

```{r}
# Convert SCMD data into `ts` object for seasonality plot
ts_scmd_all <- ts(data = scmd_all_chemicals$monthly_ddd_quantity, frequency = 12, start = c(2019, 1), end = c(2025, 12))
# Season plot
ggseasonplot(ts_scmd_all)
# Seasonal effect less obvious but there seems to be a spike in March
```

Now to fit the model: 

```{r}
total_scmd_over_time$month_factor <- factor(
  format(total_scmd_over_time$month, "%b"),
  levels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
)

scmd_all_seasonal_model0 <- lm(
  monthly_ddd_quantity ~ month_number + month_factor,
  data = total_scmd_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

scmd_all_seasonal_model <- selgmented(
  scmd_all_seasonal_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 2, # max number of changepoints = 2 as 3 breakpoints makes calculation fail
  check.dslope = F # ensures only AIC is used to select model
)
# 2 breakpoints has lowest AIC (1822.564), which is lower than the model without the dummy month factor

summary(scmd_all_seasonal_model) 
# Estimated breakpoints: 32.202, 69.014, 91.001
# Significant effect for Aug only

# Check model fit
check_model(scmd_all_seasonal_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions; similar to model w/o monthly dummy variable
# No significant outliers; slightly better fit than model w/o monthly dummy variable in this sense 

acf(residuals(scmd_all_seasonal_model))
# Autocorrelation still present at lag = 1, 7, and 17

plot(scmd_all_seasonal_model)
```

Model with only august as a dummy factor

```{r}
total_scmd_over_time$aug_factor <- factor(
  format(total_scmd_over_time$month, "%b"),
  levels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
) |> 
  fct_other(keep = "Aug", other_level = "not_aug")

scmd_all_aug_model0 <- lm(
  monthly_ddd_quantity ~ month_number + aug_factor,
  data = total_scmd_over_time,
  subset = bnf_chemical_name == "All chemicals"
)

scmd_all_aug_model <- selgmented(
  scmd_all_aug_model0,
  seg.Z = ~ month_number,
  type = "aic",
  refit = T, # improves model selection accuracy
  Kmax = 3, # max number of changepoints = 3 
  check.dslope = F # ensures only AIC is used to select model
)
# 2 breakpoints has lowest AIC (1815.751), which is lower than the other two models

summary(scmd_all_aug_model) 
# Estimated breakpoints: 24.547, 70.388
# Significant effect for Aug factor

# Check model fit
check_model(scmd_all_aug_model)
# Meets linearity, homoscedasticity, and normality of residual assumptions; similar to model w/o monthly dummy variable
# No significant outliers; slightly better fit than model w/o monthly dummy variable in this sense 

acf(residuals(scmd_all_aug_model))
# Autocorrelation still present at lag = 7 and 14

plot(scmd_all_aug_model)
```

#### Calculating predicted values and slope

Uses `predicted` for calculating predicted values and `slope3` for calculating 95% CI bounds for each segment. `Predicted` is similar to `predited_exp` but for linear models. `slope3` is very similar to Andrea's various `slope` functions. 

`extract_lin_slopes` extracts slopes and 95% CIs. 

```{r}
pred_scmd_all_lin3 <- predicted(scmd_all_lin3_model, "All chemicals", total_scmd_over_time, "SCMD")

# Generate df with slopes and 95% CI bounds for each segment
slope_scmd_all_lin3 <- extract_lin_slopes(scmd_all_lin3_model, "All chemicals", "SCMD")

pred_scmd_all_seasonal <- predicted(scmd_all_seasonal_model, "All chemicals", total_scmd_over_time, "SCMD")

pred_scmd_all_aug <- predicted(scmd_all_aug_model, "All chemicals", total_scmd_over_time, "SCMD")
```

#### Plotting the linear model with observed data.

```{r}
# Rounding breakpoints to nearest month
scmd_all_break_months <- round(c( 
  slope_scmd_all_lin3$bp1, 
  slope_scmd_all_lin3$bp2,
  slope_scmd_all_lin3$bp3
  )
)

# Convert month number into date
scmd_all_break_dates <- scmd_all_chemicals$month[scmd_all_break_months + 1]
# "2020-02-01", "2020-05-01", "2024-12-01"

ggplot() +
  geom_point(
    data = scmd_all_chemicals,
    aes(x = month, y = monthly_ddd_quantity, colour = "Observed"),
    alpha = .7
  ) +
  geom_line(
    data = pred_scmd_all_lin3,
    aes(x= month, y = pred, colour = "Segmented linear model (3bps)"),
    size = 1,
    alpha = .7
  ) +
  geom_line(
    data = pred_scmd_all_seasonal,
    aes(x= month, y = pred, colour = "Segmented seasonal model"),
    size = 1,
    alpha = .7
  ) +
  geom_line(
    data = pred_scmd_all_aug,
    aes(x= month, y = pred, colour = "Segmented august model"),
    size = 1,
    alpha = .7
  ) +
  geom_vline(xintercept = scmd_all_break_dates, colour = c("blue", "blue", "blue"), alpha = 0.5) +
  scale_colour_manual(
    values = c(
      "Observed" = "black",
      "Segmented linear model (3bps)" = "blue",
      "Segmented seasonal model" = "red",
      "Segmented august model" = "green"
    )
  ) +
  annotate(x = ymd("2020-02-01"), y = 200000, label = "2020-02-01", geom = "label", colour = "blue") +
  annotate(x = ymd("2020-05-01"), y = 175000, label = "2020-05-01", geom = "label", colour = "blue") +
  annotate(x = ymd("2024-12-01"), y = 100000, label = "2024-12-01", geom = "label", colour = "blue") +
  theme(legend.position = "right")
```

The two close breakpoints (2020-02-01 and 2020-05-01) can be interpreted as an 'unusual' period due to the COVID-19 pandemic. 