---
title: "Trends in SNOMED CT code usage indicating ADHD diagnosis and procedures in publicly available NHS primary care data (2012-2025)"
author:
  - (Redacted for blind review)
format:
    html:
        toc: false
        toc-location: body
    pdf:
        toc: false
        fig-pos: "H"
        tbl-pos: "H"
        margin-top: 1in
        margin-bottom: 1in
    docx:
        toc: false
execute:
  echo: false
  message: false
  warning: false
---

# Overview

This analysis examines SNOMED CT coding activity related to ADHD in English primary care. The aim is to understand which clinical codes are used to record ADHD diagnoses and related administrative activity, identify usage patterns over time, and inform the development of codelists for future electronic health record research on this topic.

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(kableExtra)
library(opencodecounts)
```

## Data source

We used publicly available clinical code usage data from NHS England accessed via the [opencodecounts](https://bennettoxford.github.io/opencodecounts/) R package. The annual usage count (from 1st August to  31st July) reflects how many times each listed SNOMED CT code was added to a GP electronic health record in England during the reporting period.

## Codelists

Two draft SNOMED CT codelists were developed using OpenCodelists. These codelists are preliminary and may require refinement based on clinical review.

1. **ADHD diagnosis**[^codelist-adhd-dx]: Codes that likely indicate an ADHD diagnosis. The diagnosis codelist includes some codes under review (highlighted in @tbl-diagnosis-codes) where clinical relevance requires further discussion.
2. **ADHD administrative activity**[^codelist-adhd-px]: Codes that capture referral, assessment, and other administrative activity related to ADHD (excluding diagnosis and treatment codes).

[^codelist-adhd-dx]: https://www.opencodelists.org/codelist/user/(Redacted for blind review)/adhd-diagnosis-snomed-ct-clone1/62ec200e/
[^codelist-adhd-px]: https://www.opencodelists.org/codelist/user/(Redacted for blind review)/adhd-referral-assessment-and-admin-related-codes/38f84b80/

# Results

## ADHD diagnosis

```{r}
#| label: load-diagnosis-codelist
#| output: false

snomed_diag_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/(Redacted for blind review)/adhd-diagnosis-snomed-ct-clone1/62ec200e/")

df_snomed_diag_usage <- snomed_usage |>
  filter(snomed_code %in% snomed_diag_codelist$code)
```

@tbl-diagnosis-codes shows total usage for each diagnosis code. Codes under review are highlighted; these include broader parent codes (e.g., 'Deficits in attention motor control and perception (disorder)').

```{r}
#| label: calc-diagnosis-totals

df_snomed_diag_codes_tbl <- df_snomed_diag_usage |>
  group_by(snomed_code, description) |>
  summarise(total_usage = sum(usage), .groups = "drop") |>
  arrange(desc(total_usage))

codes_under_review <- c(
  192132008,
  229715008,
  192131001,
  28271000087107,
  46745001
)
```

```{r}
#| label: tbl-diagnosis-codes
#| tbl-cap: "SNOMED CT diagnosis codes with total usage. Codes under review highlighted. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."

rows_under_review <- which(df_snomed_diag_codes_tbl$snomed_code %in% codes_under_review)

knitr::kable(
  df_snomed_diag_codes_tbl,
  col.names = c("Code", "Description", "Total usage"),
  align = c("l", "l", "r"),
  format.args = list(big.mark = ",")
) |>
  kable_styling(latex_options = c("scale_down", "HOLD_position")) |>
  column_spec(2, width = "10cm") |>
  row_spec(rows_under_review, background = "#ADD8E6")
```

@fig-diagnosis-trends shows usage trends for the five most frequently used diagnosis codes. All codes show a sharp increase from approximately 2021 onwards except 'child attention deficit disorder'.

```{r}
#| label: fig-diagnosis-trends
#| fig-cap: "SNOMED CT diagnosis code usage over time (top 5 codes). Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

top5_snomed_diag_codes <- df_snomed_diag_codes_tbl |>
  slice_max(total_usage, n = 5) |>
  pull(snomed_code)

df_snomed_diag_fig <- df_snomed_diag_usage |>
  filter(snomed_code %in% top5_snomed_diag_codes) |>
  mutate(code_label = description)

all_dates <- sort(unique(df_snomed_diag_fig$end_date))
scale_x_date_breaks <- all_dates[round(seq(1, length(all_dates), length.out = 4))]

fig_snomed_diag_trends <- ggplot(
  df_snomed_diag_fig,
  aes(
    x = end_date,
    y = usage,
    colour = code_label,
    shape = code_label
  )
) +
  geom_line(alpha = .3) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
  scale_x_date(
    breaks = scale_x_date_breaks,
    labels = scales::label_date("%b\n%Y")
  ) +
  scale_colour_viridis_d(end = .75) +
  labs(x = "End date of yearly reporting period", y = "Usage count", colour = NULL, shape = NULL) +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.text = element_text(size = 12)
  )

fig_snomed_diag_trends
```

@fig-snomed-adhd-vs-all shows code usage for all SNOMED CT ADHD diagnosis codes next to that of all SNOMED CT codes. Code usage for all codes is much higher than that for ADHD diagnosis codes so facets are used to plot this. 

```{r}
#| label: fig-snomed-adhd-vs-all
#| fig-cap: "SNOMED CT ADHD diagnosis code usage compared to all SNOMED CT codes. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

# The helper function summarise_yearly is in usage_plots.R
df_snomed_yearly_usage <- bind_rows(
  summarise_yearly(snomed_usage, "all_codes"),
  summarise_yearly(df_snomed_diag_usage, "adhd_codes")
)  

# Plot as graph
fig_snomed_adhd_vs_all <- ggplot(
  df_snomed_yearly_usage,
  aes(
    x = end_date,
    y = yearly_usage,
    colour = group
  )
 ) +
  geom_line(alpha = .5) +
  scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
  scale_x_date(
    breaks = scale_x_date_breaks,
    labels = scales::label_date("%b\n%Y")
  ) +
  scale_colour_viridis_d(end = .75) +
  labs(x = "End date of yearly reporting period", y = "Usage count", colour = NULL, shape = NULL) +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.text = element_text(size = 12)
  ) +
  facet_wrap(vars(group), scales = "free_y")

fig_snomed_adhd_vs_all
```

@fig-snomed-adhd-all-dual-y compares code usage for all SNOMED CT ADHD diagnosis codes next to that of all SNOMED CT codes but with a dual y axis. A dual y axis can be useful but some people think it's misleading. For this project the facet plot above might work better. 

```{r}
#| label: fig-snomed-adhd-all-dual
#| fig-cap: "SNOMED CT ADHD diagnosis code usage compared to all SNOMED CT codes. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

# Value used to transform the data (2nd y axis)
coeff <- 10000

fig_snomed_adhd_all_dual <- df_snomed_yearly_usage |> 
  pivot_wider(
    names_from = group,
    values_from = yearly_usage
  ) |> 
  ggplot(
  aes(x = end_date)
) +
  geom_line(aes(y = adhd_codes), alpha = .5) +
  geom_point(aes(y = adhd_codes), alpha = .3) +
  geom_col(aes(y = all_codes/coeff), alpha = .3) +
  scale_y_continuous(
    name = "ADHD diagnosis codes", 
    labels = scales::comma,
    sec.axis = sec_axis(~.*coeff, 
    name = "All codes",
    labels = scales::comma)
  )

fig_snomed_adhd_all_dual
```

@fig-pct-change-comparison compares year-on-year percentage changes in code usage for ADHD diagnosis codes and all SNOMED CT codes. There was a decrease in coding activity in 2020 for all codes combined but not for ADHD codes. The rate of increase in code usage is also higher for ADHD codes after 2022.

```{r}
#| label: fig-pct-change-comparison
#| fig-cap: "Year-on-year percentage changes in code usage. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."

df_snomed_pct_change <- df_snomed_yearly_usage |> 
  arrange(end_date, yearly_usage) |> 
  group_by(group) |> 
  mutate(pct_change = (
    (yearly_usage - lag(yearly_usage, n = 1)) / lag(yearly_usage, n=1)
  )) |> 
  mutate(pct_change = round((pct_change*100), digits = 2))

fig_pct_change_comparison <- ggplot(
  df_snomed_pct_change,
  aes(
    x = end_date,
    y = pct_change,
    colour = group
  )
) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  scale_y_continuous(limits = NULL, labels = scales::comma) +
  scale_x_date(
    breaks = scale_x_date_breaks,
    labels = scales::label_date("%b\n%Y")
  ) +
  scale_colour_viridis_d(end = .75) +
  labs(x = "End date of yearly reporting period", y = "Yearly % change in usage", colour = NULL, shape = NULL) +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.text = element_text(size = 12)
  )

fig_pct_change_comparison

ggsave(
  here("output", "figures", "fig_snomed_pct_change_comparison.png"),
  plot = fig_pct_change_comparison,
  height = 6,
  width = 10
)
```

## ADHD administrative activity

```{r}
#| label: load-admin-codelist
#| output: false

snomed_admin_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/(Redacted for blind review)/adhd-referral-assessment-and-admin-related-codes/38f84b80/")

df_snomed_admin_usage <- snomed_usage |>
  filter(snomed_code %in% snomed_admin_codelist$code)
```

@tbl-admin-codes shows total usage for each administrative activity code.

```{r}
#| label: calc-admin-totals

df_snomed_admin_codes_tbl <- df_snomed_admin_usage |>
  group_by(snomed_code, description) |>
  summarise(total_usage = sum(usage), .groups = "drop") |>
  arrange(desc(total_usage))
```

```{r}
#| label: tbl-admin-codes
#| tbl-cap: "SNOMED CT administrative activity codes with total usage. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."

options(knitr.kable.NA = 0)

knitr::kable(
  df_snomed_admin_codes_tbl,
  col.names = c("Code", "Description", "Total usage"),
  align = c("l", "l", "r"),
  format.args = list(big.mark = ",")
) |>
  kable_styling(latex_options = c("scale_down", "HOLD_position")) |>
  column_spec(2, width = "10cm")
```

@fig-admin-trends shows usage trends for the five most frequently used administrative codes. These codes were rarely used before 2022. Some codes, such as 'Diagnostic assessment for ADHD' and 'Referral for diagnostic assessment for ADHD', only appear in the data from 2024 onwards, suggesting recent adoption of more specific coding practices.

```{r}
#| label: fig-admin-trends
#| fig-cap: "SNOMED CT administrative activity code usage over time (top 5 codes). Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

top5_snomed_admin_codes <- df_snomed_admin_codes_tbl |>
  slice_max(total_usage, n = 5) |>
  pull(snomed_code)

df_snomed_admin_fig <- df_snomed_admin_usage |>
  filter(snomed_code %in% top5_snomed_admin_codes) |>
  mutate(code_label = description)

all_dates <- sort(unique(df_snomed_admin_fig$end_date))
scale_x_date_breaks <- all_dates[round(seq(1, length(all_dates), length.out = 4))]

fig_snomed_admin_trends <- ggplot(
  df_snomed_admin_fig,
  aes(
    x = end_date,
    y = usage,
    colour = code_label,
    shape = code_label
  )
) +
  geom_line(alpha = .3) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
  scale_x_date(
    breaks = scale_x_date_breaks,
    labels = scales::label_date("%b\n%Y")
  ) +
  scale_colour_viridis_d(end = .75) +
  labs(x = "End date of yearly reporting period", y = "Usage count", colour = NULL, shape = NULL) +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.text = element_text(size = 12)
  )

fig_snomed_admin_trends
```
