---
title: "Age and sex breakdowns for ADHD diagnosis in NHS England open Hospital Admitted Patient Care Activity data (2012-2025)"
author:
  - Kaylee Poon
format:
    html:
        toc: false
        toc-location: body
    pdf:
        toc: false
        fig-pos: "H"
        tbl-pos: "H"
        margin-top: 1in
        margin-bottom: 1in
    docx:
        toc: false
execute:
  echo: false
  message: false
  warning: false
---

# Overview

This analysis explores age, sex, and diagnosis position breakdowns for ICD-10 F90 codes (Hyperkinetic disorders, including ADHD) in publicly available NHS England Hospital Admitted Patient Care Activity data from April 2012 to March 2025.

# Methods

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(here)
library(ggtext)
library(kableExtra)

source(here("functions", "create_plots.R"))

icd10_f90_breakdowns <- read_csv(
  here(
    "output",
    "opencodecounts",
    "icd10_f90_breakdowns.csv"
    )
)
```

## Data source

Data extraction methods are based on the [opencodecounts](https://bennettoxford.github.io/opencodecounts/) R package.
For this analysis we extended the methods to also extract procedural and demographic breakdowns of the counts from publicly available [NHS England Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity) data.
Activity is captured in Finished Consultant Episodes (FCEs), where each FCE can have up to 20 recorded ICD-10 diagnoses. 

## Codelists

A draft ICD-10 codelist[^icd10-codelist-adhd-dx] including codes that likely indicate an ADHD diagnosis was developed using OpenCodelists. 

This codelist is preliminary and may require refinement based on clinical review. The diagnosis codelist includes some codes under review (highlighted in @tbl-icd10-codes) where clinical relevance requires further discussion.

[^icd10-codelist-adhd-dx]: https://www.opencodelists.org/codelist/user/kayleepn/adhd-icd-10-test/27569f95/

### Breakdown variables

**Diagnostic codes**: The following four-character ICD-10 codes were included: Disturbance of activity and attention (F900); _Hyperkinetic conduct disorder_ (F901); _Other hyperkinetic disorders_ (F908); _Hyperkinetic disorder, unspecified_ (F909).

**Diagnostic position**: Usage counts represent FCEs with a specific code in any diagnostic position ("all diagnoses") or as the primary reason for admission ("main diagnosis").

**Demographic breakdowns**: Demographic breakdowns are available by age groups (with ages 15-19 reported individually and aggregated here for comparability with other age bands) and sex (male, female, unknown).


## Results

### Diagnostic codes

```{r}
#| label: calc-f90-diag-percentages

df_icd10_codes_tbl <- icd10_f90_breakdowns |>
  filter(breakdown %in% c("all_diagnoses", "main_diagnosis")) |>
  group_by(icd10_code, description, breakdown) |>
  summarise(n = sum(usage)) |>
  pivot_wider(names_from = breakdown, values_from = n) |>
  arrange(icd10_code)

ratio_f900_vs_f90x <- df_icd10_codes_tbl$all_diagnoses[df_icd10_codes_tbl$icd10_code == "F900"] / sum(df_icd10_codes_tbl$all_diagnoses)
ratio_all_vs_main <- sum(df_icd10_codes_tbl$main_diagnosis, na.rm = TRUE) / sum(df_icd10_codes_tbl$all_diagnoses, na.rm = TRUE)
```

@tbl-icd10-codes shows that F900 accounts for `r round(ratio_f900_vs_f90x * 100, 2)`% of all F90 code usage. 
Main diagnoses represent only `r round(ratio_all_vs_main * 100, 2)`% of all F90 diagnosis codes, indicating that ADHD is rarely the primary reason for inpatient admission. Therefore, the following analyses focus on F900 only and do not differentiate between diagnosis positions.

```{r}
#| label: codes-under-review

codes_under_review <- c(
  "F901",
  "F908",
  "F909"
)
```

```{r}
#| label: tbl-icd10-codes
#| tbl-cap: "ICD-10 F90 codes with total usage by diagnosis position. Main diagnoses as percentage of all diagnoses: F900 <1%; F901 = 8.5%; F909 = 6%. Data source: NHS England [Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity), April 2012 to March 2025."

rows_under_review <- which(df_icd10_codes_tbl$icd10_code %in% codes_under_review)

options(knitr.kable.NA = 0)

knitr::kable(
  df_icd10_codes_tbl,
  col.names = c("Code", "Description", "All diagnoses", "Main diagnosis"),
  align = c("l", "l", "r", "r"),
  format.args = list(big.mark = ","),
) |> 
  kable_styling() |>
  row_spec(rows_under_review, background = "#ADD8E6")
```

### Age breakdowns

@fig-f90-age-breakdowns shows F900 usage by age group.
All age groups show upward trends, with steep increases from approximately 2021 onwards.
Age groups 5-9 through 25-29 show a notable decline during 2020-2021.

```{r}
#| label: fig-f90-age-breakdowns
#| fig-cap: "ICD-10 F900 usage over time by age group (ages 15-19 aggregated). Data source: NHS England [Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity), April 2012 to March 2025."
#| fig-width: 10
#| fig-height: 11

df_f90_age_breakdowns_fig <- icd10_f90_breakdowns |>
  filter(icd10_code == "F900") |>
  filter(str_starts(breakdown, "age")) |>
  # Aggregate factor levels age_15 to age_19
  mutate(
    breakdown = fct_collapse(
      breakdown,
      "age_15_19" = c("age_15", "age_16", "age_17", "age_18", "age_19")
    )
  ) |>
  # Need to also group by end_date to plot over time
  group_by(breakdown, end_date) |>
  summarise(usage = sum(usage), .groups = "drop") |>
  ungroup() |>
  group_by(breakdown) |>
  # Change age group labels from "age_x_y" to "Age x - y" and add usage count for each age group across entire time period 
  mutate(
    breakdown = str_replace(breakdown, "age_(\\d+)_(\\d+)", "Age \\1 - \\2"),
    breakdown = str_replace(breakdown, "age_(\\d+)plus", "Age \\1+"),
    breakdown = str_replace(breakdown, "age_(\\d)$", "Age \\1"),
    breakdown = paste0("**", breakdown, "**<br>n = ", scales::comma(sum(usage)))
  ) |>
  ungroup() |>
  mutate(
    breakdown = fct_reorder(breakdown, parse_number(breakdown))
  )

fig_f90_age_breakdowns <- plot_breakdown_facet(df_f90_age_breakdowns_fig)

fig_f90_age_breakdowns
```

## Sex breakdowns

@fig-f90-sex-breakdowns shows F900 usage by sex across all diagnosis positions.
Male diagnoses exceeded female diagnoses until approximately 2024, when female diagnoses surpassed male.
Gender unknown remains low throughout the period.

```{r}
#| label: fig-f90-sex-breakdowns
#| fig-cap: "ICD-10 F900 usage over time by sex (all diagnosis positions). Data source: NHS England [Hospital Admitted Patient Care Activity](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity), April 2012 to March 2025."
#| fig-width: 10
#| fig-height: 4

df_f90_sex_breakdowns_fig <- icd10_f90_breakdowns |>
  filter(icd10_code == "F900") |>
  filter(breakdown %in% c("male", "female", "gender_unknown")) |>
  group_by(breakdown) |>
  mutate(
    breakdown = str_replace(breakdown, "gender_unknown", "unknown"),
    breakdown = str_to_title(breakdown),
    breakdown = paste0(breakdown, " (n = ", scales::comma(sum(usage)), ")")
  )

all_dates <- sort(unique(df_f90_sex_breakdowns_fig$end_date))
scale_x_date_breaks <- all_dates[round(seq(
  1,
  length(all_dates),
  length.out = 4
))]

fig_f90_sex_breakdowns <- ggplot(
  df_f90_sex_breakdowns_fig,
  aes(
    x = end_date,
    y = usage,
    shape = breakdown,
    colour = breakdown
  )
) +
  geom_point(size = 2) +
  geom_line(alpha = .3) +
  scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
  scale_x_date(
    breaks = scale_x_date_breaks,
    labels = scales::label_date("%b\n%Y")
  ) +
  scale_colour_viridis_d(end = .75) +
  labs(x = "End date of yearly reporting period", y = NULL) +
  theme(
    text = element_text(size = 16),
    legend.position = "right",
    legend.title = element_blank()
  )

fig_f90_sex_breakdowns
```

