---
title: "ADHD diagnosis code usage in NHS England open SNOMED Code Usage in Primary Care data (2012-2025)"
author:
  - Kaylee Poon
format:
    html:
        toc: false
        toc-location: body
    pdf:
        toc: false
        fig-pos: "H"
        tbl-pos: "H"
        margin-top: 1in
        margin-bottom: 1in
        margin-left: 1in
        margin-right: 1in
    docx:
        toc: false
execute:
  echo: false
  message: false
  warning: false
---

# Overview

This analysis explores code usage for SNOMED CT codes that may indicate diagnosis or administrative activity in the publicly available NHS England open SNOMED Code Usage in Primary Care data (2012-2025) from 1 Aug 2012 to 31 Jul 2025.

# Methods
```{r}
#| label: setup
#| output: false
library(opencodecounts)
library(here)
library(tibble)
library(dplyr)
library(ggplot2)
library(gt)
library(kableExtra)
```

## Data source

Data is taken from the [opencodecounts](https://bennettoxford.github.io/opencodecounts/) R package.

## Codelists

Two SNOMED CT codelists were created in OpenCodelists: 

1. [ADHD diagnosis codelist](https://www.opencodelists.org/codelist/user/kayleepn/adhd-diagnosis-snomed-ct-clone1/62ec200e/): Codes that likely record a diagnosis of ADHD
2. [ADHD admin codelist](https://www.opencodelists.org/codelist/user/kayleepn/adhd-referral-assessment-and-admin-related-codes/38f84b80/): Codes that record administrative, referral, and assessment activity related to ADHD

# Results

## Usage data for ADHD diagnosis codelist

```{r}
#| label: import-and-filter-codelist
#| output: false
# Import codelist from OpenCodelists.org
snomed_diag_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/kayleepn/adhd-diagnosis-snomed-ct-clone1/62ec200e/")
as_tibble(snomed_diag_codelist)

# Filter code usage data to include codes from SNOMED-CT diagnosis codelist
df_snomed_diag_usage <- snomed_usage |>
    filter(snomed_code %in% snomed_diag_codelist$code)
```

### Total code usage for each code

@tbl-diagnosis-codes shows the total code usage (across all years) for each code. Borderline codes that may not be relevant for the thesis are highlighted in light blue. Note that the parent code 'Disorders of attention and motor control 229712006' is included but has no usage data. We may also want to consider if we are including codes related to drug treatment of ADHD. 

```{r}
#| label: tbl-diagnosis-codes
#| tbl-cap: "SNOMED CT diagnosis codes with total usage. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| tbl-pos: "H"

# Summarise total code usage for each code
all_snomed_diag_codes <- df_snomed_diag_usage |>
    group_by(snomed_code, description) |>
    summarise(total_usage = sum(usage)) |>
    ungroup() |> 
    arrange(desc(total_usage))

# Create list of borderline codes
borderline_snomed_diag_codes <- c(
    192132008,
    229715008,
    192131001,
    28271000087107,
    46745001
)

# Create code usage table with borderline codes highlighted. 
all_snomed_diag_usage <- all_snomed_diag_codes |> 
    gt() |> 
    cols_width(
    snomed_code ~ px(150),
    description ~ px(350),
    total_usage ~ px(120)
    ) |>
    cols_label(
        snomed_code = "Code",
        description = "Description",
        total_usage = "Total usage"
    ) |> 
    tab_header(
        title = "SNOMED ADHD diagnosis code usage") |> 
    tab_style(
        cell_fill(color = "lightblue"),
        locations = cells_body(
            rows = snomed_code %in% borderline_snomed_diag_codes
        ) 
    ) |> 
    tab_style(
        style = cell_text(whitespace = "normal"),
        locations = cells_body(columns = everything())) |> 
    cols_align(
        align = "left",
        columns = snomed_code
    )

all_snomed_diag_usage
```

### Usage trends over time (top 5 codes)

@fig-diagnosis-trends-top-5 shows that code usage increased sharply after 2021.

```{r}
#| label: fig-diagnosis-trends-top-5
#| fig-cap: "SNOMED CT diagnosis code usage over time. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

# Select top 5 codes with most usage 
top5_snomed_diag_codes <- all_snomed_diag_codes |> 
    slice_max(total_usage, n = 5) |> 
    pull(snomed_code)

# Visualise trends over time
plot_snomed_diag_usage <- df_snomed_diag_usage |> 
    filter(snomed_code %in% top5_snomed_diag_codes) |> 
    ggplot(aes(
        x = end_date,
        y = usage,
        colour = paste0(description, "(", snomed_code, ")"),
        shape = paste0(description, "(", snomed_code, ")" )
    )) +
    geom_line(alpha = .3) +
    geom_point() +
    scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
    scale_x_date(date_breaks = "2 years",
    date_labels = "%Y") +
    scale_colour_viridis_d(end = .7) +
    labs(
        x = NULL, y = "Usage count", 
        colour = NULL, shape = NULL
    ) +
    theme(legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.text = element_text(size = 12))

plot_snomed_diag_usage  
```

## Usage data for ADHD admin codelist

This codelist contains codes related to referral, assessment, and administrative activity for ADHD, but not diagnosis or treatment.

```{r}
#| label: import-and-filter-admin-codelist
#| output: false
# Import codelist from OpenCodelists.org
snomed_admin_codelist <- get_codelist("https://www.opencodelists.org/codelist/user/kayleepn/adhd-referral-assessment-and-admin-related-codes/38f84b80/")
as_tibble(snomed_admin_codelist)

# Filter code usage data to include codes from SNOMED-CT admin codelist
df_snomed_admin_usage <- snomed_usage |>
    filter(snomed_code %in% snomed_admin_codelist$code)
```

### Total code usage for each code

@tbl-admin-codes shows the total code usage (across all years) for each code.

```{r}
#| label: tbl-admin-codes
#| tbl-cap: "SNOMED CT admin codes with total usage. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."

all_snomed_admin_codes <- df_snomed_admin_usage |>
    group_by(snomed_code, description) |>
    summarise(total_usage = sum(usage)) |>
    ungroup() |> 
    arrange(desc(total_usage))

options(knitr.kable.NA = 0)

all_snomed_admin_codes |> 
    knitr::kable(
        col.names = c("Code", "Description", "Total usage"),
        align = c("l", "l", "r"),
        format.args = list(big.mark = ",")
    ) |> 
    kable_styling(latex_options = c("scale_down")) |> 
    column_spec(2, width = "9cm")
```

### Usage trends over time (top 5 codes)

@fig-admin-trends-top-5 shows total code usage (across all years) for each admin-related code. Note that admin codes were not frequently used until around 2022, and some codes such as 'Diagnostic assessment for attention deficit hyperactivity disorder (1914891000000100)' and 'Referral for diagnostic assessment for attention deficit hyperactivity disorder (2437121000000109)' first have activity recorded in 2024 and 2025 respectively. 

```{r}
#| label: fig-admin-trends-top-5
#| fig-cap: "SNOMED CT diagnosis code usage over time. Data source: NHS England [SNOMED Code Usage in Primary Care](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care), Aug 2012 to Jul 2025."
#| fig-width: 10
#| fig-height: 6

# Select top 5 codes with most usage
top5_snomed_admin_codes <- all_snomed_admin_codes |> 
    slice_max(total_usage, n = 5) |> 
    pull(snomed_code)

# Visualise trends over time 
plot_snomed_admin_usage <- df_snomed_admin_usage |> 
    filter(snomed_code %in% top5_snomed_admin_codes) |> 
    ggplot(aes(
        x = end_date,
        y = usage,
        colour = paste0(description, "(", snomed_code, ")"),
        shape = paste0(description, "(", snomed_code, ")" )
    )) +
    geom_line(alpha = .3) +
    geom_point() +
    scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
    scale_x_date(date_breaks = "2 years",
    date_labels = "%Y") +
    scale_colour_viridis_d(end = .7) +
    labs(
        x = NULL, y = "Usage count", 
        colour = NULL, shape = NULL
    ) +
    theme(legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.text = element_text(size = 12))
plot_snomed_admin_usage    
```
